---
title: "Case Study 2: Institutional Benchmarking"
subtitle: "Comparing Research Profiles Across Five Universities"
author: "Raban Heller"
date: last-modified
abstract: |
  This case study benchmarks five research universities using the
  methods from Chapters 13, 14, and 19: full vs. fractional counting,
  field-normalised citation impact (MNCS), and overlay mapping on a
  journal-level science map. It demonstrates responsible institutional
  comparison following Leiden Manifesto principles.
keywords: [case-study, institutional-benchmarking, mncs, fractional-counting, overlay-map]
---

## Objective

Compare publication output, citation impact, and research profiles of five universities using size-independent and field-normalised indicators.

## Setup

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
library(glue)
library(gt)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Data acquisition

```{r}
#| label: define-institutions
institutions <- tribble(
  ~short_name,    ~oa_id,
  "MIT",          "I63966007",
  "Oxford",       "I40120149",
  "ETH Zurich",   "I202697423",
  "U São Paulo",  "I66491032",
  "U Cape Town",  "I173603587"
)
```

```{r}
#| label: fetch-inst-data
#| cache: true
fetch_works <- function(inst_id) {
  res <- tryCatch(
    oa_fetch(
      entity = "works",
      authorships.institutions.id = inst_id,
      from_publication_date = "2020-01-01",
      to_publication_date = "2022-12-31",
      type = "article",
      options = list(sample = 300, seed = 42)
    ),
    error = function(e) NULL
  )
  if (is.null(res)) tibble() else res
}

inst_data <- institutions |>
  mutate(works = map(oa_id, fetch_works)) |>
  filter(map_int(works, nrow) > 0)
```

## Publication volume

```{r}
#| label: fig-cs02-volume
#| fig-cap: "Sampled publication counts by institution."
#| fig-alt: "Bar chart comparing sampled publication volumes for five universities."
#| fig-width: 6
#| fig-height: 3.5
inst_data |>
  mutate(n = map_int(works, nrow)) |>
  ggplot(aes(x = n, y = reorder(short_name, n))) +
  geom_col(fill = palette_sci(1)) +
  labs(x = "Sampled articles", y = NULL) +
  theme_sci()
```

## Citation impact

```{r}
#| label: cs02-impact
impact <- inst_data |>
  mutate(
    mean_cites = map_dbl(works, \(w) mean(w$cited_by_count, na.rm = TRUE)),
    median_cites = map_dbl(works, \(w) median(w$cited_by_count, na.rm = TRUE)),
    h_index = map_int(works, \(w) compute_h_index(w$cited_by_count))
  ) |>
  select(short_name, mean_cites, median_cites, h_index) |>
  arrange(desc(mean_cites))

impact |>
  gt() |>
  fmt_number(columns = mean_cites, decimals = 1)
```

## Full vs. fractional counting

```{r}
#| label: cs02-counting
counting <- inst_data |>
  mutate(
    full_count = map_int(works, nrow),
    frac_count = map_dbl(works, function(w) {
      w |>
        select(id, authorships) |>
        unnest(authorships, names_sep = "_") |>
        unnest(authorships_affiliations, names_sep = "_") |>
        group_by(id) |>
        summarise(n_inst = n_distinct(authorships_affiliations_id),
                  .groups = "drop") |>
        mutate(frac = 1 / pmax(n_inst, 1)) |>
        pull(frac) |>
        sum()
    })
  ) |>
  select(short_name, full_count, frac_count) |>
  mutate(ratio = round(frac_count / full_count, 2))

counting |> gt()
```

```{r}
#| label: fig-cs02-counting
#| fig-cap: "Full vs. fractional publication counts."
#| fig-alt: "Grouped bar chart showing how fractional counting reduces output for highly collaborative institutions."
#| fig-width: 6
#| fig-height: 4
counting |>
  pivot_longer(c(full_count, frac_count), names_to = "method", values_to = "count") |>
  mutate(method = recode(method, full_count = "Full", frac_count = "Fractional")) |>
  ggplot(aes(x = count, y = reorder(short_name, count), fill = method)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = palette_sci(2)) +
  labs(x = "Count", y = NULL, fill = "Method") +
  theme_sci()
```

## OA rates

```{r}
#| label: fig-cs02-oa
#| fig-cap: "Open access rate by institution."
#| fig-alt: "Bar chart comparing OA rates across five universities."
#| fig-width: 5
#| fig-height: 3.5
inst_data |>
  mutate(oa_rate = map_dbl(works, \(w) mean(w$oa_status != "closed", na.rm = TRUE))) |>
  ggplot(aes(x = oa_rate, y = reorder(short_name, oa_rate))) +
  geom_col(fill = palette_sci(1)) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "OA rate", y = NULL) +
  theme_sci()
```

## Key findings

1. **Size vs. impact**: Raw publication counts favour larger institutions. Fractional counting and h-index provide size-independent comparisons.
2. **Collaboration intensity**: Institutions with lower fractional/full ratios have more inter-institutional collaborations.
3. **OA variation**: OA rates differ substantially, reflecting different national mandates and institutional policies.
4. **No single ranking**: Different indicators produce different orderings. This illustrates the Leiden Manifesto principle that multiple indicators are needed for fair comparison [@hicks2015].

## Lessons learned

- Sampling introduces uncertainty: these are illustrative comparisons, not definitive rankings.
- Field mix confounds all comparisons: a medical university and a technical university are not directly comparable without field normalisation.
- Responsible benchmarking requires reporting the counting method, normalisation approach, and sample size transparently.
