---
title: "Theoretical Underpinnings"
subtitle: "Lotka, Bradford, Zipf, Price, and the Matthew Effect"
author: "Raban Heller"
date: last-modified
abstract: |
  The empirical regularities of scholarly communication — power-law
  productivity, journal scattering, vocabulary distributions, exponential
  growth, and cumulative advantage — form the theoretical backbone of
  scientometrics. This chapter introduces each law with historical context
  and demonstrates it with reproducible R code on OpenAlex data.
keywords: [scientometrics, lotka, bradford, zipf, price, matthew-effect, power-law]
---

## Learning objectives {#sec-theory-objectives}

After completing this chapter, you will be able to:

- State Lotka's, Bradford's, and Zipf's laws and explain the empirical patterns they describe
- Test whether a corpus of publications approximately follows Lotka's inverse-square law
- Construct a Bradford plot to identify core and peripheral journals in a field
- Explain Price's law of exponential growth and the Matthew Effect in science
- Recognise power-law-like distributions in bibliometric data

## Setup {#sec-theory-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
library(tidytext)
library(glue)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#sec-theory-background}

Scientometrics rests on a small number of empirical regularities discovered across decades of studying scholarly communication:

**Lotka's Law** [@lotka1926] observes that the number of authors producing *n* papers is approximately proportional to 1/*n*². A few prolific authors generate a large share of output, while most researchers publish only one or two papers.

**Bradford's Law** [@bradford1934] describes the scattering of articles across journals. For a given topic, a small core of journals contains roughly one-third of all relevant articles; a larger zone contains the next third; and a much larger periphery the final third. Each successive zone contains roughly the same number of articles but geometrically more journals.

**Zipf's Law** states that the frequency of a word is inversely proportional to its rank. In scientific text, a handful of terms dominate while the vast majority appear rarely.

**Price's Law** [@price1963] generalises these observations: the cumulative output of science grows exponentially, doubling roughly every 10–15 years. Price also noted that the square root of all contributors produce half of all contributions.

The **Matthew Effect** [@merton1968] — named after the Gospel of Matthew ("to those who have, more shall be given") — describes cumulative advantage in science. Well-known researchers attract disproportionate credit and resources, amplifying initial differences in visibility. This mechanism helps explain the skewed distributions observed by Lotka and Price.

## Worked example {#sec-theory-example}

We use a corpus of works from the journal *Scientometrics* to demonstrate each law.

### Fetching the data

```{r}
#| label: fetch-theory-data
#| cache: true
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2015-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 500, seed = 42)
)
```

### Lotka's Law

We unnest authors and count the number of papers per author in this sample.

```{r}
#| label: lotka-data
author_prod <- works |>
  select(id, authorships) |>
  unnest(authorships, names_sep = "_") |>
  count(authorships_display_name, name = "n_papers") |>
  count(n_papers, name = "n_authors") |>
  arrange(n_papers)

author_prod
```

```{r}
#| label: fig-theory-lotka
#| fig-cap: "Author productivity distribution. The dashed line shows the inverse-square law predicted by Lotka."
#| fig-alt: "Log-log scatter plot of the number of authors versus the number of papers they published, with a dashed reference line showing the inverse-square law."
#| fig-width: 6
#| fig-height: 4
lotka_c <- author_prod$n_authors[1]

author_prod |>
  ggplot(aes(x = n_papers, y = n_authors)) +
  geom_point(colour = palette_sci(1), size = 2) +
  geom_line(
    aes(y = lotka_c / n_papers^2),
    linetype = "dashed", colour = "grey40"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Papers per author", y = "Number of authors") +
  theme_sci()
```

The data approximate the inverse-square pattern, though real distributions often deviate from the exact exponent of 2.

### Bradford's Law

We count how many articles each journal source contributes (across a broader query).

```{r}
#| label: bradford-data
#| cache: true
broad_works <- oa_fetch(
  entity = "works",
  search = "bibliometrics",
  from_publication_date = "2020-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 1000, seed = 42)
)

journal_counts <- broad_works |>
  filter(!is.na(source_display_name)) |>
  count(source_display_name, name = "articles", sort = TRUE) |>
  mutate(
    cum_articles = cumsum(articles),
    rank = row_number()
  )
```

```{r}
#| label: fig-theory-bradford
#| fig-cap: "Bradford plot: cumulative articles vs. log journal rank."
#| fig-alt: "Line plot showing cumulative article count on the y-axis versus journal rank on a logarithmic x-axis. The characteristic S-shaped Bradford curve is visible, with a steep rise for core journals."
#| fig-width: 6
#| fig-height: 4
journal_counts |>
  ggplot(aes(x = rank, y = cum_articles)) +
  geom_line(colour = palette_sci(1), linewidth = 1) +
  scale_x_log10() +
  labs(x = "Journal rank (log scale)", y = "Cumulative articles") +
  theme_sci()
```

### Zipf's Law

We extract title words and examine their frequency distribution.

```{r}
#| label: zipf-data
word_freq <- works |>
  select(display_name) |>
  unnest_tokens(word, display_name, token = "words") |>
  anti_join(tidytext::get_stopwords(), by = "word") |>
  count(word, sort = TRUE) |>
  mutate(rank = row_number())
```

```{r}
#| label: fig-theory-zipf
#| fig-cap: "Word frequency vs. rank for article titles, approximating Zipf's law."
#| fig-alt: "Log-log scatter plot of word frequency versus rank, showing an approximately linear relationship consistent with Zipf's law."
#| fig-width: 6
#| fig-height: 4
word_freq |>
  head(200) |>
  ggplot(aes(x = rank, y = n)) +
  geom_point(colour = palette_sci(1), size = 1, alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Rank", y = "Frequency") +
  theme_sci()
```

### Price's growth law

```{r}
#| label: fig-theory-growth
#| fig-cap: "Annual publication counts for sampled works, illustrating the growth of bibliometrics literature."
#| fig-alt: "Bar chart showing increasing annual publication counts from 2015 to 2023."
#| fig-width: 6
#| fig-height: 3.5
works |>
  mutate(year = year(publication_date)) |>
  count(year) |>
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = palette_sci(1)) +
  labs(x = "Year", y = "Works in sample") +
  theme_sci()
```

## Diagnostics and interpretation {#sec-theory-diagnostics}

These laws are *empirical regularities*, not physical laws. They describe approximate patterns:

- Lotka's exponent is often not exactly 2; values between 1.5 and 3.5 are common depending on the field and time window.
- Bradford zones are sensitive to how you define the subject scope.
- Zipf's law holds best for common words; the tail deviates.
- Price's exponential growth eventually saturates — science cannot grow faster than the economy indefinitely.

The value of these laws is not predictive precision but the insight that **skewed distributions are the norm, not the exception**, in scholarly communication.

## Limitations and responsible use {#sec-theory-limitations}

::: {.callout-warning}
## Limitations and responsible use

- These laws describe aggregate patterns, not individual behaviour. A researcher publishing one paper is not "less productive" in any evaluative sense — Lotka's law is a statistical observation, not a performance standard.
- The Matthew Effect [@merton1968] is a mechanism that *amplifies inequality*. Recognising it should prompt caution about using cumulative metrics (like the h-index) that inherit and reinforce this bias.
- Power-law fitting is notoriously difficult. Many distributions that "look" power-law on a log-log plot are better described by lognormal or stretched exponential models. Rigorous fitting requires maximum-likelihood methods and goodness-of-fit tests [@hicks2015].
:::

## Common pitfalls {#sec-theory-pitfalls}

::: {.callout-important}
## Common pitfalls

- **Eyeballing log-log plots.** A straight line on a log-log plot does not prove a power law. Use formal statistical tests.
- **Confusing description with prescription.** "Most authors publish few papers" does not mean they *should* publish more.
- **Ignoring field differences.** The exponent of Lotka's law and the scattering pattern of Bradford's law vary substantially across disciplines.
- **Assuming stationarity.** These patterns can shift over time as publication norms change (e.g., the rise of mega-journals).
:::

## Exercises {#sec-theory-exercises}

1. **Lotka exponent.** Fit a linear model to the log-log author productivity data. What exponent do you estimate? How does it compare to Lotka's original value of 2?

2. **Bradford zones.** Divide the journal list into three zones of roughly equal article counts. How many journals are in each zone? What is the ratio between successive zone sizes?

3. **Your field's Zipf distribution.** Fetch 500 works in a field of your choice. Extract title words and plot the Zipf distribution. Are the most frequent terms what you would expect?

4. **Matthew Effect simulation.** Write a simple simulation: start 100 authors with 1 paper each. At each time step, an author gains a new paper with probability proportional to their current count. After 50 steps, plot the distribution. Does it resemble Lotka's law?

::: {.callout-tip collapse="true"}
## Solutions

Solutions are provided in @sec-solutions.
:::

## Further reading {#sec-theory-reading}

- @lotka1926 — The original observation on author productivity distributions.
- @bradford1934 — Journal scattering and the Bradford law.
- @price1963 — Exponential growth of science and cumulative advantage.
- @merton1968 — The Matthew Effect: how advantage accumulates in science.
- @garfield1955 — Citation indexing, the empirical basis for much of scientometrics.
- @waltman2016 — A modern review contextualising these classical laws within citation impact measurement.

## Session info {#sec-theory-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
