# Parsing Native Exports

## Learning objectives {#parsing-objectives}

After completing this chapter, you will be able to:

- Parse BibTeX files using `bib2df` and `RefManageR`
- Import Web of Science plain-text and Scopus CSV exports via `bibliometrix`
- Standardize field names across formats into a unified schema
- Diagnose common parsing failures and missing fields

## Setup {#parsing-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(bib2df)
library(RefManageR)
library(bibliometrix)
library(glue)

set.seed(20260509)

source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#parsing-background}

Not all bibliometric data comes from an API. Researchers frequently begin with files exported from database search interfaces: BibTeX (`.bib`) from Google Scholar or Zotero, RIS (`.ris`) from Scopus or PubMed, plain text from Web of Science, or CSV from Scopus. Each format encodes the same information — authors, title, year, journal, DOI — in different structures.

The `bibliometrix` package [@aria2017] provides `convert2df()`, a versatile function that reads multiple export formats and returns a standardized data frame. For BibTeX specifically, `bib2df` and `RefManageR` offer complementary parsing capabilities. The key challenge is not parsing per se but *standardization*: ensuring that author names, journal titles, and identifiers are represented consistently regardless of the source format.

## Worked example {#parsing-example}

### Parsing a BibTeX string

We construct a small BibTeX string inline to demonstrate parsing without external files.

```{r}
#| label: parse-bibtex
bib_text <- '
@article{smith2023,
  author = {Smith, John A. and Doe, Jane B.},
  title = {Advances in Bibliometric Methods},
  journal = {Journal of Informetrics},
  year = {2023},
  volume = {17},
  pages = {101--115},
  doi = {10.1234/example.2023}
}
@article{chen2022,
  author = {Chen, Wei and Li, Xiao},
  title = {Citation Networks in Computer Science},
  journal = {Scientometrics},
  year = {2022},
  volume = {127},
  pages = {3201--3220},
  doi = {10.1234/example.2022}
}
'

tmp_bib <- tempfile(fileext = ".bib")
writeLines(bib_text, tmp_bib)

bib_df <- bib2df(tmp_bib)
glimpse(bib_df)
```

### Parsing with RefManageR

```{r}
#| label: parse-refmanager
bib_rm <- ReadBib(tmp_bib)
as.data.frame(bib_rm) |> select(title, author, year, journal, doi)
```

### Parsing WoS plain text with bibliometrix

```{r}
#| label: parse-wos-demo
#| eval: false
# Requires a WoS export file — not bundled
# wos_df <- convert2df("savedrecs.txt", dbsource = "wos", format = "plaintext")
# glimpse(wos_df)
```

### Standardizing field names

```{r}
#| label: standardize-fields
standardize_bib <- function(df) {
  df |>
    transmute(
      title = TITLE,
      authors = AUTHOR,
      year = as.integer(YEAR),
      journal = JOURNAL,
      doi = DOI
    )
}

bib_df |> standardize_bib()
```

### Visualization

```{r}
#| label: fig-parsing-fields
#| fig-cap: "Field completeness in the parsed BibTeX records."
#| fig-alt: "Horizontal bar chart showing which metadata fields are present in the parsed BibTeX records."
#| fig-width: 6
#| fig-height: 3
completeness <- bib_df |>
  summarise(across(everything(), ~ mean(!is.na(.)))) |>
  pivot_longer(everything(), names_to = "field", values_to = "present") |>
  filter(present > 0) |>
  mutate(field = fct_reorder(field, present))

ggplot(completeness, aes(x = present, y = field)) +
  geom_col(fill = palette_sci(1)) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Completeness", y = NULL) +
  theme_sci()
```

## Diagnostics and interpretation {#parsing-diagnostics}

After parsing, always check:

- **Row count**: Does the number of records match the export? Missing entries may indicate parsing errors.
- **Field completeness**: Which fields are populated? Abstracts and keywords are often missing in BibTeX exports.
- **Author format**: Names may appear as "Last, First" or "First Last" depending on the source. Standardize before analysis.
- **Encoding**: Non-ASCII characters (diacritics, CJK) can corrupt during parsing. Ensure UTF-8 encoding.

## Limitations and responsible use {#parsing-limitations}

<div class="callout callout-warn">
## Limitations and responsible use

- Export files are snapshots: they reflect the database state at export time and become stale immediately.
- Field mapping between formats is imperfect. Some formats lack fields (RIS has no standard abstract tag across all databases).
- Parsing errors are silent — always validate record counts and spot-check metadata against the source [@hicks2015].
</div>

## Common pitfalls {#parsing-pitfalls}

<div class="callout callout-stop">
## Common pitfalls

- **Encoding mismatches**: Exporting as Latin-1 and reading as UTF-8 (or vice versa) corrupts names and titles.
- **Truncated exports**: Many databases cap exports at 500 or 1,000 records per file. Large corpora require multiple export batches.
- **Inconsistent author delimiters**: BibTeX uses "and", RIS uses newlines, CSV uses semicolons. Parsing must handle each.
- **Assuming format from extension**: A `.txt` file may be WoS plain text or something else entirely. Check the first few lines.
</div>

## Exercises {#parsing-exercises}

1. **Export and parse.** Export 50 records from Google Scholar as BibTeX. Parse them with `bib2df`. How many fields are populated?

2. **Format comparison.** Export the same set of records from Scopus as both CSV and RIS. Parse both and compare the fields available.

3. **Standardization function.** Write a function that takes a data frame from `convert2df()` and returns a tibble with columns: `doi`, `title`, `authors`, `year`, `journal`, `cited_by`.

<div class="callout callout-ok">
## Solutions

Solutions are provided in \@ref(solutions).
</div>

## Further reading {#parsing-reading}

- @aria2017 — The `bibliometrix` package, including `convert2df()` for multi-format parsing.
- @priem2022 — OpenAlex as an API-first alternative to file-based workflows.

## Session info {#parsing-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
