---
title: "Journal Metrics"
subtitle: "JIF, SJR, SNIP, Eigenfactor, and CiteScore Proxies"
author: "Raban Heller"
date: last-modified
abstract: |
  This chapter explains how major journal-level indicators are computed,
  why we build open proxies from OpenAlex rather than relying on
  proprietary values, and how network-based metrics (Eigenfactor, SJR)
  differ from simple citation ratios. A worked example computes JIF-like
  and SNIP-like proxies for a set of journals using OpenAlex data.
keywords: [scientometrics, journal-metrics, impact-factor, sjr, snip, eigenfactor, citescore]
---

## Learning objectives {#sec-jmetrics-objectives}

After completing this chapter, you will be able to:

- Describe how the Journal Impact Factor, CiteScore, SNIP, SJR, and Eigenfactor are computed
- Explain why network-based journal metrics differ from simple citation ratios
- Compute JIF-like and SNIP-like proxies from OpenAlex data
- Articulate why proxy values differ from official proprietary metrics
- Evaluate journals using multiple indicators rather than a single number

## Setup {#sec-jmetrics-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
library(glue)
library(gt)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#sec-jmetrics-background}

Journal-level metrics aim to summarise a journal's citation performance. The most famous is the **Journal Impact Factor** (JIF), introduced by @garfield1955 and commercialised by Clarivate. The two-year JIF for year *t* equals the citations received in year *t* by articles published in years *t*-1 and *t*-2, divided by the number of "citable items" published in those two years. Its simplicity is both its appeal and its weakness: the numerator includes citations to all document types, but the denominator often excludes editorials and letters, inflating the ratio [@hicks2015].

**CiteScore** (Elsevier) uses a four-year window and counts all document types in both numerator and denominator, making it more transparent. **SNIP** (Source-Normalized Impact per Paper) normalises by citation potential --- the average number of references in citing articles' reference lists --- so that journals in fields with longer reference lists are not automatically advantaged [@waltman2016]. **SJR** (SCImago Journal Rank) and the **Eigenfactor** use iterative algorithms inspired by PageRank: a citation from a prestigious journal carries more weight than one from a low-impact source. This makes them network-based indicators, conceptually similar to eigenvector centrality in graph theory.

All of these official metrics are tied to proprietary databases (Web of Science or Scopus). With OpenAlex [@priem2022], we can compute **proxy** values from fully open data. Proxies will not match official figures exactly --- differences in coverage, document-type classification, and citation linking guarantee some divergence [@visser2021] --- but they are transparent, reproducible, and free.

DORA [@dora2012] explicitly warns against using the JIF as a proxy for individual article quality. The Leiden Manifesto [@hicks2015] echoes this: journal metrics describe *average* journal performance and say nothing about any specific paper.

## Worked example {#sec-jmetrics-example}

### Selecting journals

We analyse five well-known journals in information science and scientometrics.

```{r}
#| label: define-journals
journals <- tribble(
  ~short_name,        ~source_id,
  "Scientometrics",   "S39224261",
  "J Informetrics",   "S205038879",
  "Quant Sci Stud",   "S4210178031",
  "J Am Soc Inf Sci", "S145342767",
  "Res Evaluation",   "S100532675"
)
```

### Fetching citation data

We fetch works published in 2021-2022 and count citations received during 2022-2023 to construct a two-year JIF-like proxy.

```{r}
#| label: fetch-journal-data
#| cache: true
fetch_journal_works <- function(source_id, from_year, to_year) {
  oa_fetch(
    entity = "works",
    primary_location.source.id = source_id,
    from_publication_date = paste0(from_year, "-01-01"),
    to_publication_date = paste0(to_year, "-12-31"),
    type = "article"
  )
}

journal_works <- journals |>
  mutate(data = map2(source_id, 2021, \(sid, yr) {
    fetch_journal_works(sid, yr, yr + 1)
  }))
```

### Computing JIF-like proxies

```{r}
#| label: compute-jif-proxy
jif_proxy <- journal_works |>
  mutate(
    n_articles = map_int(data, nrow),
    total_cites = map_dbl(data, \(d) sum(d$cited_by_count, na.rm = TRUE))
  ) |>
  mutate(jif_proxy = total_cites / pmax(n_articles, 1)) |>
  select(short_name, n_articles, total_cites, jif_proxy) |>
  arrange(desc(jif_proxy))

jif_proxy |>
  gt() |>
  fmt_number(columns = jif_proxy, decimals = 2) |>
  cols_label(
    short_name = "Journal",
    n_articles = "Articles (2021-22)",
    total_cites = "Total citations",
    jif_proxy = "JIF proxy"
  )
```

### Computing a SNIP-like proxy

SNIP normalises by citation potential. We approximate this by dividing each journal's mean citations per paper by the median reference-list length of its citing works.

```{r}
#| label: compute-snip-proxy
snip_proxy <- journal_works |>
  mutate(
    mean_cites = map_dbl(data, \(d) mean(d$cited_by_count, na.rm = TRUE)),
    median_refs = map_dbl(data, \(d) {
      median(d$referenced_works_count, na.rm = TRUE)
    })
  ) |>
  mutate(
    citation_potential = pmax(median_refs, 1),
    snip_proxy = mean_cites / citation_potential
  ) |>
  select(short_name, mean_cites, citation_potential, snip_proxy) |>
  arrange(desc(snip_proxy))

snip_proxy |>
  gt() |>
  fmt_number(columns = c(mean_cites, snip_proxy), decimals = 2)
```

### Visualization

```{r}
#| label: fig-jmetrics-comparison
#| fig-cap: "Comparison of JIF proxy and SNIP proxy across five journals."
#| fig-alt: "Dot plot showing JIF proxy and SNIP proxy values side by side for five information science journals, illustrating how normalisation changes relative rankings."
#| fig-width: 6
#| fig-height: 4
combined <- jif_proxy |>
  select(short_name, jif_proxy) |>
  left_join(snip_proxy |> select(short_name, snip_proxy), by = "short_name") |>
  pivot_longer(cols = c(jif_proxy, snip_proxy), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, jif_proxy = "JIF proxy", snip_proxy = "SNIP proxy"))

ggplot(combined, aes(x = value, y = reorder(short_name, value), colour = metric)) +
  geom_point(size = 3) +
  labs(x = "Metric value", y = NULL, colour = "Metric") +
  scale_colour_manual(values = palette_sci(2)) +
  theme_sci()
```

## Diagnostics and interpretation {#sec-jmetrics-diagnostics}

When evaluating journal metric proxies:

- **Coverage gaps**: OpenAlex may miss some document types or citations present in Scopus/WoS. Compare article counts with known totals.
- **Document-type classification**: The JIF denominator famously excludes editorials and letters. Our proxy counts all articles; this may lower the ratio compared to official JIF values.
- **Temporal alignment**: Ensure citation windows match the intended metric definition exactly (two years for JIF, four for CiteScore).
- **Rank correlation**: Even if absolute values differ from official metrics, the *rank order* of journals often agrees, which is sufficient for many analyses.

## Limitations and responsible use {#sec-jmetrics-limitations}

::: {.callout-warning}
## Limitations and responsible use

- **Journal metrics are not article metrics.** The distribution of citations within a journal is highly skewed; the mean says little about any individual paper [@dora2012].
- **Goodhart's law applies.** When a metric becomes a target, it ceases to be a good metric [@goodhart1984]. Editors may game the JIF through citation stacking, coercive citation, or manipulating the denominator.
- **Proxy ≠ official.** Our OpenAlex-based proxies are transparent and reproducible, but they will differ from Clarivate JIF or Elsevier CiteScore. Never present proxies as official values.
- **Field dependence persists.** Even SNIP does not fully eliminate field effects. Comparing SNIP values across distant disciplines requires caution [@waltman2016].
:::

## Common pitfalls {#sec-jmetrics-pitfalls}

::: {.callout-important}
## Common pitfalls

- **Treating the JIF as a measure of article quality.** The JIF describes the journal, not the paper. A paper in a high-JIF journal may have zero citations.
- **Ignoring the citation window.** Computing a "JIF" using lifetime citations instead of a two-year window produces a different and non-comparable number.
- **Conflating numerator and denominator document types.** Mixing reviews (which inflate the numerator) with a denominator that excludes them inflates the ratio.
- **Ranking journals by a single metric.** Always use multiple indicators and consider disciplinary context.
:::

## Exercises {#sec-jmetrics-exercises}

1. **Four-year CiteScore proxy.** Modify the worked example to compute a four-year CiteScore proxy (citations in year *t* to articles published in years *t*-1 to *t*-4, divided by all documents in those years). How does the ranking change?

2. **Rank correlation.** Compute Spearman's rank correlation between your JIF proxy and SNIP proxy. Is the agreement high or low?

3. **Field comparison.** Pick five journals from two different fields (e.g., physics and sociology). Compute JIF proxies for both groups. Does SNIP normalisation reduce the between-field gap?

4. **Self-citation share.** For one journal, estimate the proportion of citations that come from the journal itself. How would excluding self-citations change the JIF proxy?

::: {.callout-tip collapse="true"}
## Solutions

Solutions are provided in @sec-cheatsheets.
:::

## Further reading {#sec-jmetrics-reading}

- @garfield1955 --- The original proposal for citation indexes, from which the JIF eventually emerged.
- @waltman2016 --- Comprehensive review of citation impact indicators, including journal-level metrics.
- @hicks2015 --- The Leiden Manifesto, with specific guidance on responsible use of journal metrics.
- @dora2012 --- DORA: explicitly warns against using the JIF for individual evaluation.
- @goodhart1984 --- Goodhart's law: when a measure becomes a target, it ceases to be a good measure.
- @visser2021 --- Database comparison showing how coverage affects computed metrics.
- @priem2022 --- OpenAlex as a fully open data source for computing metric proxies.

## Session info {#sec-jmetrics-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
