# Co-word and Keyword Co-occurrence

## Learning objectives {#coword-objectives}

After completing this chapter, you will be able to:

- Explain the principles of co-word analysis and its role in science mapping
- Extract and clean keywords from OpenAlex data (author keywords and concepts)
- Build a keyword co-occurrence matrix and convert it to a network
- Apply community detection to identify topical clusters
- Visualise keyword co-occurrence networks with interpretable layouts

## Setup {#coword-setup}

```{r}
#| label: 17-setup
#| message: false
library(tidyverse)
library(openalexR)
library(igraph)
library(tidygraph)
library(ggraph)
library(glue)
library(gt)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#coword-background}

Co-word analysis maps the conceptual structure of a research field by examining which terms appear together in the same documents. Introduced by @callon1983, the method assumes that when two keywords repeatedly co-occur across publications, they are thematically related. The resulting co-occurrence network reveals the topical landscape of a field: clusters of densely connected keywords represent coherent research themes, while bridges between clusters indicate interdisciplinary connections.

The method works with several types of terms:

- **Author keywords**: Terms selected by the paper's authors. These are intentional descriptors but suffer from inconsistency --- authors may use different terms for the same concept ("machine learning" vs. "ML" vs. "statistical learning").
- **Indexed keywords**: Terms assigned by database indexers (e.g., MeSH terms in PubMed). More consistent but only available in some databases.
- **OpenAlex concepts**: Algorithmically assigned topics at multiple levels of a hierarchical taxonomy. These provide broad coverage but may miss nuanced distinctions [@priem2022].
- **Title/abstract words**: Extracted via text mining. Comprehensive but noisy; requires extensive preprocessing.

Normalisation is important for co-word networks. Raw co-occurrence counts favour high-frequency terms. Common normalisations include the **association strength** (equivalent to pointwise mutual information) and the **equivalence index** (cosine of the co-occurrence vector). @waltman2012 demonstrated that association strength produces the most balanced network structures for bibliometric mapping.

Co-word analysis complements citation-based methods (\@ref(cocit-background)). Citation networks reveal intellectual influence; co-word networks reveal thematic content. Combining both provides a more complete picture of a field's structure.

## Worked example {#coword-example}

### Extracting keywords

We extract author keywords from a sample of scientometrics research.

```{r}
#| label: 17-fetch-coword-data
#| cache: true
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2019-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 500, seed = 42)
)
```

```{r}
#| label: 17-extract-keywords
keywords <- works |>
  select(id, topics) |>
  unnest(topics, names_sep = "_") |>
  filter(topics_display_name != "") |>
  select(work_id = id, keyword = topics_display_name) |>
  mutate(keyword = str_to_lower(str_trim(keyword))) |>
  filter(!is.na(keyword), nchar(keyword) >= 3)

kw_counts <- keywords |>
  count(keyword, sort = TRUE)

cat(glue("Total keyword occurrences: {nrow(keywords)}\n"))
cat(glue("Unique keywords: {n_distinct(keywords$keyword)}\n"))
```

```{r}
#| label: 17-fig-coword-topkeywords
#| fig-cap: "Top 20 most frequent keywords in the Scientometrics sample."
#| fig-alt: "Horizontal bar chart showing the 20 most frequently occurring keywords, with frequency counts on the x-axis."
#| fig-width: 6
#| fig-height: 5
kw_counts |>
  head(20) |>
  mutate(keyword = fct_reorder(keyword, n)) |>
  ggplot(aes(x = n, y = keyword)) +
  geom_col(fill = palette_sci(1)) +
  labs(x = "Frequency", y = NULL) +
  theme_sci()
```

### Building the co-occurrence network

We create edges between keywords that appear in the same paper.

```{r}
#| label: 17-build-coword-network
kw_frequent <- kw_counts |>
  filter(n >= 5) |>
  pull(keyword)

kw_filtered <- keywords |>
  filter(keyword %in% kw_frequent)

coword_pairs <- kw_filtered |>
  inner_join(kw_filtered, by = "work_id", suffix = c("_a", "_b"),
             relationship = "many-to-many") |>
  filter(keyword_a < keyword_b) |>
  count(keyword_a, keyword_b, name = "cooccurrence")

coword_top <- coword_pairs |>
  filter(cooccurrence >= 3)

g_kw <- graph_from_data_frame(
  coword_top |> select(keyword_a, keyword_b, weight = cooccurrence),
  directed = FALSE
) |>
  simplify(edge.attr.comb = list(weight = "sum"))

cat(glue("Keyword network: {vcount(g_kw)} nodes, {ecount(g_kw)} edges\n"))
```

### Community detection for topical clusters

```{r}
#| label: 17-keyword-communities
V(g_kw)$degree <- degree(g_kw)
V(g_kw)$strength <- strength(g_kw)

communities <- cluster_leiden(g_kw, resolution_parameter = 0.8,
                              objective_function = "modularity")
V(g_kw)$community <- as.factor(membership(communities))

cat(glue("Communities: {length(unique(membership(communities)))}\n"))
cat(glue("Modularity: {round(modularity(g_kw, membership(communities)), 3)}\n"))
```

```{r}
#| label: 17-community-keywords
community_summary <- tibble(
  keyword = V(g_kw)$name,
  community = V(g_kw)$community,
  strength = V(g_kw)$strength
) |>
  group_by(community) |>
  slice_max(strength, n = 5) |>
  summarise(top_keywords = paste(keyword, collapse = ", "),
            n_keywords = n(), .groups = "drop")

community_summary |> gt()
```

### Visualisation

```{r}
#| label: 17-fig-coword-network
#| fig-cap: "Keyword co-occurrence network coloured by topical cluster."
#| fig-alt: "Network graph where nodes are keywords and edges connect keywords that frequently co-occur. Node size reflects degree; colours indicate topical communities identified by the Leiden algorithm."
#| fig-width: 7
#| fig-height: 5.5
set.seed(42)
layout <- create_layout(as_tbl_graph(g_kw), layout = "fr")

ggraph(layout) +
  geom_edge_link(aes(width = weight), alpha = 0.15, colour = "grey60") +
  scale_edge_width_continuous(range = c(0.3, 2), guide = "none") +
  geom_node_point(aes(size = degree, colour = community), alpha = 0.8) +
  geom_node_text(aes(label = ifelse(degree > quantile(degree, 0.85),
                                     name, NA_character_)),
                 repel = TRUE, size = 2.5, max.overlaps = 20, na.rm = TRUE) +
  scale_size_continuous(range = c(1, 6), guide = "none") +
  scale_colour_manual(values = palette_sci(
    n_distinct(V(g_kw)$community)
  )) +
  labs(colour = "Cluster") +
  theme_void(base_family = "sans", base_size = 11) +
  theme(legend.position = "bottom")
```

## Diagnostics and interpretation {#coword-diagnostics}

- **Keyword cleaning**: Inconsistent terminology inflates the vocabulary and creates spurious nodes. Standardise spelling, merge synonyms (e.g., "h-index" and "hirsch index"), and remove generic terms ("research", "analysis") that co-occur with everything but convey no topical information.
- **Frequency threshold**: Including rare keywords produces large, sparse, unreadable networks. Start with keywords appearing in at least 5 papers and adjust based on corpus size.
- **Community interpretability**: Each community should correspond to a recognisable research theme. If communities are uninterpretable, the resolution parameter may need adjustment or keywords need further cleaning.
- **Centrality interpretation**: High-degree keywords are thematically central (used across many contexts). High-betweenness keywords bridge distinct topics and may represent interdisciplinary concepts.

## Limitations and responsible use {#coword-limitations}

<div class="callout callout-warn">
## Limitations and responsible use

- **Vocabulary inconsistency.** Author keywords are not controlled vocabulary. The same concept may appear under multiple terms, fragmenting what should be a single node. Merging synonyms requires domain expertise.
- **Algorithmic concepts.** OpenAlex concepts are assigned by machine learning and may contain errors — especially at fine-grained levels. Always validate topic assignments by reading sample papers [@priem2022].
- **Static snapshots.** A co-word network represents a time-averaged view. Emerging topics with few publications may be invisible. Consider building temporal slices to track field evolution.
- **Not evaluative.** Popular keywords indicate research activity, not research quality or societal impact. Do not equate topical centrality with importance [@hicks2015].
</div>

## Common pitfalls {#coword-pitfalls}

<div class="callout callout-stop">
## Common pitfalls

- **Not cleaning keywords.** Punctuation variants ("co-authorship" vs. "coauthorship"), case differences, and trailing whitespace create duplicate nodes. Clean before building the network.
- **Including stop-keywords.** Generic terms like "research", "study", "analysis" co-occur with everything and dominate the network without conveying topical information. Remove them.
- **Comparing co-occurrence counts across corpora of different sizes.** Larger corpora produce higher co-occurrence counts mechanically. Normalise or use relative measures.
- **Over-reading small clusters.** A cluster with two or three keywords may reflect a single paper, not a research theme. Report cluster size alongside content.
</div>

## Exercises {#coword-exercises}

1. **Author keywords vs. concepts.** Build co-occurrence networks from both author keywords (if available) and OpenAlex concepts for the same corpus. Compare the resulting topical clusters.

2. **Temporal evolution.** Split the corpus into yearly slices. For each year, build a co-word network and identify the top 5 keywords by degree. Which keywords are consistently central, and which emerge over time?

3. **Strategic diagram.** Compute density (internal cohesion) and centrality (external connections) for each keyword cluster. Plot them on a strategic diagram (density vs. centrality quadrants). Which clusters are core themes and which are peripheral?

4. **Normalisation comparison.** Build two networks from the same data: one with raw co-occurrence counts and one with association strength normalisation. How do the most central keywords differ?

<div class="callout callout-ok">
## Solutions

Solutions are provided in \@ref(solutions).
</div>

## Further reading {#coword-reading}

- @callon1983 --- The foundational paper on co-word analysis for mapping science.
- @waltman2012 --- Unified framework for bibliometric network construction, including keyword networks.
- @aria2017 --- `bibliometrix` includes co-word analysis via the `biblioNetwork()` function.
- @priem2022 --- OpenAlex concepts and topic classification.

## Session info {#coword-sessioninfo}

```{r}
#| label: 17-session-info
sessionInfo()
```
