# Text Mining Bibliographic Corpora

## Learning objectives {#textmine-objectives}

After completing this chapter, you will be able to:

- Build a text corpus from OpenAlex titles and abstracts
- Apply standard preprocessing steps: tokenisation, stopword removal, stemming
- Construct a document-term matrix (DTM) and a document-feature matrix (DFM)
- Compute TF-IDF weights to identify distinctive terms
- Visualise term frequency patterns across time or groups

## Setup {#textmine-setup}

```{r}
#| label: 21-setup
#| message: false
library(tidyverse)
library(openalexR)
library(quanteda)
library(quanteda.textstats)
library(glue)
library(gt)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#textmine-background}

Bibliometric metadata --- titles, abstracts, and keywords --- constitutes a rich but noisy text corpus. Text mining transforms this unstructured data into structured representations suitable for statistical analysis. The pipeline typically follows three stages: preprocessing, representation, and analysis.

**Preprocessing** converts raw text into a standardised form. Tokenisation splits text into individual words or n-grams. Lowercasing removes case variation. Stopword removal eliminates high-frequency function words ("the", "of", "and") that carry little topical information. Stemming or lemmatisation reduces words to their root forms ("computing", "computed", "computation" → "comput" or "compute"). Each step involves trade-offs: aggressive stemming can merge distinct concepts, while minimal preprocessing retains noise.

The **document-term matrix** (DTM) or **document-feature matrix** (DFM) is the fundamental representation. Rows are documents; columns are terms; cells contain counts or weights. Raw term frequencies overweight common words. **TF-IDF** (term frequency–inverse document frequency) addresses this by upweighting terms that are frequent in a document but rare across the corpus, highlighting words that distinguish one document from others.

`quanteda` [@aria2017] provides a fast, well-designed toolkit for text analysis in R. Its `corpus → tokens → dfm` pipeline integrates naturally with the tidyverse. For very large corpora, `quanteda` uses sparse matrix representations that scale to millions of documents.

Text mining complements network-based methods (\@ref(coword-background)). Co-word networks reveal term associations; text mining provides the frequency distributions, temporal trends, and discriminative features that characterise a field's vocabulary.

## Worked example {#textmine-example}

### Building a text corpus

```{r}
#| label: 21-fetch-textmine-data
#| cache: true
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2019-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 500, seed = 42)
)

text_df <- works |>
  filter(!is.na(abstract), nchar(abstract) > 50) |>
  transmute(
    doc_id = id,
    text = paste(display_name, abstract, sep = ". "),
    year = year(publication_date)
  )

cat(glue("Documents with abstracts: {nrow(text_df)}\n"))
```

```{r}
#| label: 21-build-corpus
corp <- corpus(text_df, docid_field = "doc_id", text_field = "text")
docvars(corp, "year") <- text_df$year

cat(glue("Corpus size: {ndoc(corp)} documents\n"))
cat(glue("Total tokens: {sum(ntoken(corp))}\n"))
```

### Tokenisation and preprocessing

```{r}
#| label: 21-tokenise
toks <- tokens(corp,
               remove_punct = TRUE,
               remove_numbers = TRUE,
               remove_symbols = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en")) |>
  tokens_remove(c("also", "however", "using", "based", "study",
                   "paper", "results", "research", "analysis")) |>
  tokens_wordstem()

cat(glue("Tokens after preprocessing: {sum(ntoken(toks))}\n"))
```

### Document-feature matrix

```{r}
#| label: 21-build-dfm
dfmat <- dfm(toks) |>
  dfm_trim(min_termfreq = 5, min_docfreq = 3)

cat(glue("DFM dimensions: {nrow(dfmat)} docs x {ncol(dfmat)} features\n"))
cat(glue("Sparsity: {scales::percent(sparsity(dfmat))}\n"))
```

```{r}
#| label: 21-fig-textmine-topfeatures
#| fig-cap: "Top 20 most frequent terms in the Scientometrics corpus."
#| fig-alt: "Horizontal bar chart showing the 20 most frequent terms after preprocessing, with term frequency on the x-axis."
#| fig-width: 6
#| fig-height: 5
top_terms <- topfeatures(dfmat, 20) |>
  enframe(name = "term", value = "frequency") |>
  mutate(term = fct_reorder(term, frequency))

ggplot(top_terms, aes(x = frequency, y = term)) +
  geom_col(fill = palette_sci(1)) +
  labs(x = "Frequency", y = NULL) +
  theme_sci()
```

### TF-IDF weighting

```{r}
#| label: 21-compute-tfidf
dfmat_tfidf <- dfm_tfidf(dfmat)

tfidf_by_year <- text_df |>
  group_by(year) |>
  group_keys() |>
  pull(year) |>
  map_dfr(function(yr) {
    docs <- docvars(dfmat_tfidf, "year") == yr
    if (sum(docs) < 5) return(tibble())
    top <- topfeatures(dfmat_tfidf[docs, ], 10)
    tibble(year = yr, term = names(top), tfidf = unname(top))
  })

tfidf_by_year |>
  group_by(year) |>
  slice_max(tfidf, n = 5) |>
  gt() |>
  fmt_number(columns = tfidf, decimals = 1)
```

### Term frequency over time

```{r}
#| label: 21-fig-textmine-temporal
#| fig-cap: "Relative frequency of selected terms over time."
#| fig-alt: "Line chart showing how the relative frequency of key terms changes across publication years, revealing shifts in research vocabulary."
#| fig-width: 6
#| fig-height: 4
target_terms <- c("bibliometr", "open", "network", "impact", "collabor")

freq_by_year <- map_dfr(unique(text_df$year), function(yr) {
  docs <- docvars(dfmat, "year") == yr
  if (sum(docs) < 5) return(tibble())
  freq <- colSums(dfmat[docs, ]) / sum(dfmat[docs, ])
  tibble(
    year = yr,
    term = target_terms,
    rel_freq = freq[target_terms]
  )
})

ggplot(freq_by_year, aes(x = year, y = rel_freq, colour = term)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = palette_sci(length(target_terms))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(x = "Year", y = "Relative frequency", colour = "Term") +
  theme_sci()
```

## Diagnostics and interpretation {#textmine-diagnostics}

- **Vocabulary size vs. documents**: A typical DFM has far more features than documents. Trim aggressively (minimum document frequency of 3–5) to reduce noise and computation.
- **Sparsity**: Bibliometric DFMs are typically 95–99% sparse. This is normal. Sparse matrix representations keep memory usage manageable.
- **Preprocessing sensitivity**: Results change with preprocessing choices. Always report stopword list, stemming method, and trimming thresholds.
- **Abstract availability**: Not all OpenAlex records have abstracts. Report the coverage rate and consider whether missing abstracts introduce bias (e.g., older papers, certain publishers).

## Limitations and responsible use {#textmine-limitations}

<div class="callout callout-warn">
## Limitations and responsible use

- **Abstracts are summaries.** They capture the main claims but miss nuance, methodology details, and negative results. Full-text analysis (\@ref(fulltext-background)) provides richer data but is harder to obtain.
- **Language bias.** Text mining tools work best for English. Non-English abstracts may be poorly tokenised, incorrectly stemmed, or excluded entirely. This biases results toward Anglophone research [@visser2021].
- **Stemming conflates concepts.** "Stem" and "stemming" are merged, but so are "cell" (biology) and "cell" (spreadsheet). Lemmatisation is more precise but slower.
- **Bag-of-words ignores context.** TF-IDF and frequency-based methods treat documents as unordered collections of words. Phrase meaning ("machine learning" vs. "learning machine") is lost unless you use n-grams or embeddings (\@ref(embed-background)).
</div>

## Common pitfalls {#textmine-pitfalls}

<div class="callout callout-stop">
## Common pitfalls

- **Not removing domain-specific stop words.** Generic stopword lists miss terms like "study", "results", "paper" that dominate academic text without conveying topical information.
- **Applying TF-IDF before trimming.** Rare terms get extreme TF-IDF scores. Trim the DFM first, then apply TF-IDF.
- **Comparing raw frequencies across groups of different sizes.** A year with 200 papers will have higher raw counts than a year with 50. Always normalise to relative frequency.
- **Stemming titles but not abstracts (or vice versa).** Apply identical preprocessing to all text fields.
</div>

## Exercises {#textmine-exercises}

1. **Bigram analysis.** Build a DFM using bigrams (two-word sequences) instead of unigrams. What meaningful phrases emerge that single words miss?

2. **Lexical diversity.** Compute the type-token ratio (TTR) for each publication year. Is the vocabulary growing more diverse or more standardised over time?

3. **Keyness analysis.** Use `quanteda.textstats::textstat_keyness()` to identify terms that distinguish one year from all others. What terms characterise 2023 specifically?

4. **Preprocessing sensitivity.** Run the same analysis with and without stemming. How do the top terms differ? Which version is more interpretable?

<div class="callout callout-ok">
## Solutions

Solutions are provided in \@ref(solutions).
</div>

## Further reading {#textmine-reading}

- @silge2017 --- *Text Mining with R*; comprehensive guide to tidy text analysis.
- @aria2017 --- `bibliometrix` text-mining features for bibliometric data.
- @priem2022 --- OpenAlex abstract availability and coverage.

## Session info {#textmine-sessioninfo}

```{r}
#| label: 21-session-info
sessionInfo()
```
