---
title: "Embeddings for Scientometrics"
subtitle: "Document Similarity Beyond Bag-of-Words"
author: "Raban Heller"
date: last-modified
abstract: |
  Word and document embeddings capture semantic similarity that
  bag-of-words methods miss. This chapter introduces word2vec-style
  embeddings and pre-trained scientific document embeddings, demonstrates
  how to compute document similarity from abstract embeddings, and
  applies dimensionality reduction to visualise the semantic landscape
  of a research corpus. Python-based tools (SPECTER) are accessed via
  reticulate where needed.
keywords: [scientometrics, embeddings, word2vec, document-similarity, dimensionality-reduction, specter]
---

## Learning objectives {#sec-embed-objectives}

After completing this chapter, you will be able to:

- Explain how word and document embeddings differ from bag-of-words representations
- Train simple word embeddings on a bibliometric corpus
- Use pre-trained embeddings to compute document similarity
- Apply UMAP or t-SNE to visualise document clusters in embedding space
- Assess when embeddings add value over simpler text representations

## Setup {#sec-embed-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
library(quanteda)
library(word2vec)
library(uwot)
library(glue)
library(gt)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#sec-embed-background}

Bag-of-words representations (@sec-textmine-background) treat each word as an independent dimension. "Citation analysis" and "bibliometric study" share no features despite being semantically related. **Word embeddings** address this by mapping words into a dense, low-dimensional vector space where semantically similar words are close together. The word2vec algorithm learns these representations by predicting words from their context (skip-gram) or context from words (CBOW) in a large training corpus.

**Document embeddings** extend the idea to whole documents. Simple approaches average the word vectors of a document's words (mean pooling). More sophisticated approaches include Doc2Vec (Paragraph Vector), which learns document-specific vectors alongside word vectors, and transformer-based models like **SPECTER**, which is pre-trained on scientific paper titles and abstracts using citation-based training signals.

For scientometrics, embeddings enable several applications:

- **Semantic similarity**: Finding papers with similar meaning, even when they use different terminology.
- **Clustering**: Grouping papers by content in embedding space, complementing topic models (@sec-topicmodel-background).
- **Visualisation**: Projecting embeddings to 2D with UMAP or t-SNE to create interpretable "maps" of a corpus.
- **Anomaly detection**: Identifying papers that are semantically distant from their assigned field or cluster.

The trade-off is interpretability: while TF-IDF features are directly readable (the word "citation" has a specific weight), embedding dimensions are opaque. Embeddings are powerful for similarity and clustering but resist the kind of term-level interpretation that co-word analysis provides.

## Worked example {#sec-embed-example}

### Preparing text data

```{r}
#| label: fetch-embed-data
#| cache: true
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2019-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 400, seed = 42)
)

text_df <- works |>
  filter(!is.na(abstract), nchar(abstract) > 100) |>
  transmute(
    doc_id = id,
    title = display_name,
    text = paste(display_name, abstract, sep = ". "),
    year = year(publication_date),
    cited_by_count
  )

cat(glue("Documents for embedding: {nrow(text_df)}\n"))
```

### Training word2vec embeddings

```{r}
#| label: train-word2vec
#| cache: true
clean_text <- text_df$text |>
  str_to_lower() |>
  str_replace_all("[^a-z ]", " ") |>
  str_squish()

model <- word2vec(clean_text, dim = 100, iter = 20, min_count = 3,
                  type = "skip-gram", threads = 1)

cat(glue("Vocabulary size: {nrow(as.matrix(model))}\n"))
```

```{r}
#| label: word-similarity
if ("citation" %in% rownames(as.matrix(model))) {
  nn <- predict(model, newdata = "citation", type = "nearest", top_n = 10)
  cat("Nearest neighbours of 'citation':\n")
  print(nn)
}
```

### Computing document embeddings via mean pooling

```{r}
#| label: document-embeddings
word_matrix <- as.matrix(model)

doc_embed <- function(text, wm) {
  words <- str_split(str_to_lower(str_replace_all(text, "[^a-z ]", " ")),
                     "\\s+")[[1]]
  matched <- words[words %in% rownames(wm)]
  if (length(matched) == 0) return(rep(NA_real_, ncol(wm)))
  colMeans(wm[matched, , drop = FALSE])
}

doc_embeddings <- map(text_df$text, \(t) doc_embed(t, word_matrix))
embed_mat <- do.call(rbind, doc_embeddings)

valid <- complete.cases(embed_mat)
embed_mat <- embed_mat[valid, ]
text_df_valid <- text_df[valid, ]

cat(glue("Valid document embeddings: {nrow(embed_mat)}\n"))
```

### Dimensionality reduction with UMAP

```{r}
#| label: umap-projection
#| cache: true
umap_result <- umap(embed_mat, n_neighbors = 15, min_dist = 0.1,
                    n_components = 2, ret_model = FALSE)

text_df_valid$umap_x <- umap_result[, 1]
text_df_valid$umap_y <- umap_result[, 2]
```

```{r}
#| label: fig-embed-umap
#| fig-cap: "UMAP projection of document embeddings, coloured by publication year."
#| fig-alt: "Scatter plot showing documents positioned by semantic similarity in UMAP space. Colour indicates publication year, revealing whether temporal patterns correspond to semantic clusters."
#| fig-width: 6
#| fig-height: 5
ggplot(text_df_valid, aes(x = umap_x, y = umap_y, colour = factor(year))) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_colour_manual(values = palette_sci(n_distinct(text_df_valid$year))) +
  labs(x = "UMAP 1", y = "UMAP 2", colour = "Year") +
  theme_sci() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

### Document similarity search

```{r}
#| label: similarity-search
cosine_sim <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))

query_idx <- 1
sims <- map_dbl(seq_len(nrow(embed_mat)),
                \(i) cosine_sim(embed_mat[query_idx, ], embed_mat[i, ]))

top_similar <- order(sims, decreasing = TRUE)[2:6]

cat(glue("Query: {text_df_valid$title[query_idx]}\n\n"))
cat("Most similar documents:\n")
walk(top_similar, \(i) {
  cat(glue("  [{round(sims[i], 3)}] {text_df_valid$title[i]}\n"))
})
```

### Embedding clusters vs. citation impact

```{r}
#| label: fig-embed-impact
#| fig-cap: "UMAP embedding coloured by citation impact (log scale)."
#| fig-alt: "Scatter plot in UMAP space with documents coloured by log citation count, showing whether highly cited papers cluster together semantically."
#| fig-width: 6
#| fig-height: 5
ggplot(text_df_valid, aes(x = umap_x, y = umap_y,
                          colour = log1p(cited_by_count))) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_colour_viridis_c(option = "C", name = "log(cites + 1)") +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme_sci() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Diagnostics and interpretation {#sec-embed-diagnostics}

- **Vocabulary coverage**: Check what fraction of abstract words appear in the embedding vocabulary. Low coverage (< 80%) indicates the training corpus is too small or preprocessing is too aggressive.
- **Nearest neighbours**: Inspect the nearest neighbours of key domain terms. If "citation" is closest to "references" and "bibliometric", the embeddings capture domain structure.
- **UMAP stability**: UMAP is stochastic. Run with multiple seeds and check whether the global cluster structure is stable. Individual point positions will vary but clusters should persist.
- **Embedding dimensionality**: 100–300 dimensions is standard. Below 50, embeddings lose semantic resolution; above 300, training becomes slow with diminishing returns.

## Limitations and responsible use {#sec-embed-limitations}

::: {.callout-warning}
## Limitations and responsible use

- **Embeddings are opaque.** Unlike TF-IDF weights, embedding dimensions have no interpretable meaning. You can measure similarity but not explain *why* two documents are similar without additional analysis.
- **Training data bias.** Embeddings reflect the biases of their training corpus. If the corpus overrepresents certain topics or perspectives, the embeddings will too.
- **Small corpora produce poor embeddings.** Word2vec needs tens of thousands of documents for reliable training. For small bibliometric corpora (< 1,000 papers), use pre-trained embeddings instead.
- **UMAP/t-SNE distort distances.** These methods preserve local structure but distort global distances. Documents that appear close in 2D may not be the most similar in high-dimensional space [@hicks2015].
:::

## Common pitfalls {#sec-embed-pitfalls}

::: {.callout-important}
## Common pitfalls

- **Training on too few documents.** Word2vec on 200 abstracts produces unreliable embeddings. Use at least 5,000 documents or switch to pre-trained models.
- **Averaging without weighting.** Simple mean pooling treats every word equally. TF-IDF-weighted averaging gives more weight to distinctive terms and often produces better document embeddings.
- **Over-interpreting UMAP clusters.** Visual clusters in UMAP may reflect projection artifacts. Validate clusters with a separate method (e.g., k-means in the full embedding space).
- **Comparing embeddings from different models.** Embedding spaces are not aligned across different training runs. Cosine similarity is only meaningful within a single embedding space.
:::

## Exercises {#sec-embed-exercises}

1. **Weighted averaging.** Compute document embeddings using TF-IDF-weighted word vectors instead of simple means. Does the UMAP visualisation change?

2. **K-means clustering.** Apply k-means (k = 5, 8, 12) to the document embeddings. Compare the resulting clusters with LDA topics from @sec-topicmodel-example. Do they agree?

3. **Pre-trained embeddings.** If available, use SPECTER embeddings (via reticulate and Python) instead of corpus-trained word2vec. How does similarity search quality change?

4. **Analogy test.** Test whether the word2vec model captures analogical relationships (e.g., "journal" - "article" + "book" ≈ "publisher"). What does the result tell you about the model?

::: {.callout-tip collapse="true"}
## Solutions

Solutions are provided in @sec-cheatsheets.
:::

## Further reading {#sec-embed-reading}

- @silge2017 --- Text embeddings in R, including integration with word2vec.
- @priem2022 --- OpenAlex concept embeddings and their relationship to document content.
- @waltman2012 --- Bibliometric similarity measures as context for embedding-based approaches.

## Session info {#sec-embed-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
