---
title: "Detecting Emerging Topics"
subtitle: "Burst Detection, Novelty Indicators, and Temporal Topic Tracking"
author: "Raban Heller"
date: last-modified
abstract: |
  Identifying emerging research topics is one of the most valuable
  applications of scientometrics. This chapter covers methods for
  detecting emergence: Kleinberg burst detection on keyword time series,
  temporal topic tracking with rolling-window analysis, novelty and
  transience indicators, and growth-rate metrics. The worked example
  identifies bursting terms in a scientometrics corpus and tracks
  topic evolution over time.
keywords: [scientometrics, emerging-topics, burst-detection, novelty, temporal-analysis, research-fronts]
---

## Learning objectives {#sec-emerging-objectives}

After completing this chapter, you will be able to:

- Define what constitutes an "emerging topic" in scientometric terms
- Apply burst detection to keyword time series
- Track topic prevalence over time using rolling-window analysis
- Compute novelty and transience indicators for documents
- Distinguish genuine emergence from database coverage artefacts

## Setup {#sec-emerging-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
library(quanteda)
library(quanteda.textstats)
library(tidytext)
library(glue)
library(gt)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#sec-emerging-background}

An **emerging topic** is a research area experiencing rapid growth in attention --- new publications, new keywords, increasing citation flows. Detecting emergence early has strategic value for researchers, funders, and policymakers. But distinguishing genuine intellectual emergence from database artefacts (a journal newly indexed, a conference proceedings batch-imported) requires careful methodology.

**Burst detection** formalises the intuition that a keyword or topic has "burst" when its frequency increases sharply relative to baseline. @kleinberg2003 introduced a state-based model: a keyword exists in a "normal" state (baseline frequency) or a "burst" state (elevated frequency). The algorithm identifies time intervals during which each keyword is in its burst state. Cascading bursts at multiple time scales reveal both short-lived spikes and sustained surges.

**Temporal topic tracking** extends topic models (@sec-topicmodel-background) by fitting models to time-sliced corpora or using dynamic topic models where topic distributions evolve smoothly over time. A simpler approach uses rolling windows: fit a model to each 3-year window and track how topic proportions change between windows.

**Novelty and transience indicators** operate at the document level. A paper is *novel* if its content (measured by embedding distance or topic distribution) is distant from prior literature. A paper is *transient* if it is distant from subsequent literature --- it attracted brief attention but did not generate follow-on work. Genuinely emerging topics are characterised by papers that are novel *and* non-transient: they introduce new ideas that the field continues to develop.

**Growth-rate metrics** are the simplest emergence indicators: the year-over-year growth rate of publications containing a keyword, the acceleration of citation rates, or the increase in the number of distinct institutions working on a topic. These are intuitive but sensitive to baseline effects (a keyword going from 1 to 3 papers is 200% growth but not meaningful emergence).

## Worked example {#sec-emerging-example}

### Preparing temporal keyword data

```{r}
#| label: fetch-emerging-data
#| cache: true
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2015-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 600, seed = 42)
)

text_df <- works |>
  filter(!is.na(abstract), nchar(abstract) > 50) |>
  transmute(
    doc_id = id,
    text = paste(display_name, abstract, sep = ". "),
    year = year(publication_date)
  )

cat(glue("Documents: {nrow(text_df)}\n"))
cat(glue("Year range: {min(text_df$year)}--{max(text_df$year)}\n"))
```

```{r}
#| label: build-keyword-timeseries
corp <- corpus(text_df, docid_field = "doc_id", text_field = "text")
docvars(corp, "year") <- text_df$year

toks <- tokens(corp, remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en")) |>
  tokens_remove(c("study", "paper", "results", "research", "analysis",
                   "also", "however", "using", "based"))

dfmat <- dfm(toks) |>
  dfm_trim(min_termfreq = 10, min_docfreq = 5)

kw_by_year <- convert(dfm_group(dfmat, groups = year), to = "data.frame") |>
  pivot_longer(-doc_id, names_to = "term", values_to = "count") |>
  rename(year = doc_id) |>
  mutate(year = as.integer(year))

cat(glue("Terms tracked: {n_distinct(kw_by_year$term)}\n"))
```

### Growth rate detection

```{r}
#| label: growth-rates
growth <- kw_by_year |>
  group_by(term) |>
  arrange(year) |>
  mutate(
    growth = (count - lag(count)) / pmax(lag(count), 1),
    total = sum(count)
  ) |>
  ungroup() |>
  filter(!is.na(growth), total >= 20)

recent_growth <- growth |>
  filter(year >= 2021) |>
  group_by(term) |>
  summarise(mean_growth = mean(growth), total = first(total),
            .groups = "drop") |>
  filter(mean_growth > 0) |>
  arrange(desc(mean_growth))

recent_growth |>
  head(15) |>
  gt() |>
  fmt_percent(columns = mean_growth, decimals = 0)
```

### Simple burst detection

```{r}
#| label: burst-detection
detect_burst <- function(counts, threshold = 2) {
  mean_count <- mean(counts)
  sd_count <- sd(counts)
  if (sd_count == 0) return(rep(FALSE, length(counts)))
  z_scores <- (counts - mean_count) / sd_count
  z_scores > threshold
}

bursts <- kw_by_year |>
  group_by(term) |>
  filter(n() >= 5, sum(count) >= 20) |>
  arrange(year) |>
  mutate(is_burst = detect_burst(count, threshold = 1.5)) |>
  ungroup()

burst_summary <- bursts |>
  filter(is_burst) |>
  group_by(term) |>
  summarise(
    burst_years = paste(year, collapse = ", "),
    max_count = max(count),
    .groups = "drop"
  ) |>
  arrange(desc(max_count))

burst_summary |> head(15) |> gt()
```

### Visualising keyword trajectories

```{r}
#| label: fig-emerging-trajectories
#| fig-cap: "Temporal trajectories of emerging keywords with burst periods highlighted."
#| fig-alt: "Line chart showing the frequency of several keywords over time. Burst periods (unusually high frequency) are highlighted with background shading."
#| fig-width: 7
#| fig-height: 5
top_burst_terms <- burst_summary |> head(6) |> pull(term)

burst_plot_data <- bursts |>
  filter(term %in% top_burst_terms)

ggplot(burst_plot_data, aes(x = year, y = count)) +
  geom_rect(data = burst_plot_data |> filter(is_burst),
            aes(xmin = year - 0.5, xmax = year + 0.5,
                ymin = -Inf, ymax = Inf),
            fill = "gold", alpha = 0.3, inherit.aes = FALSE) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_wrap(~ term, scales = "free_y", ncol = 3) +
  labs(x = "Year", y = "Frequency") +
  theme_sci(base_size = 9)
```

### Rolling-window topic tracking

```{r}
#| label: rolling-window
window_size <- 3
years <- sort(unique(text_df$year))
windows <- years[1:(length(years) - window_size + 1)]

window_top_terms <- map_dfr(windows, function(start_year) {
  end_year <- start_year + window_size - 1
  docs <- docvars(dfmat, "year") >= start_year &
          docvars(dfmat, "year") <= end_year
  if (sum(docs) < 10) return(tibble())
  top <- topfeatures(dfmat[docs, ], 10)
  tibble(window = glue("{start_year}-{end_year}"),
         term = names(top), frequency = unname(top))
})
```

```{r}
#| label: fig-emerging-windows
#| fig-cap: "Top terms in each 3-year rolling window, showing vocabulary evolution."
#| fig-alt: "Heatmap or faceted bar chart showing the top terms for each rolling time window, revealing which terms gain or lose prominence."
#| fig-width: 7
#| fig-height: 5
window_top_terms |>
  group_by(window) |>
  slice_max(frequency, n = 8) |>
  mutate(term = reorder_within(term, frequency, window)) |>
  ggplot(aes(x = frequency, y = term)) +
  geom_col(fill = palette_sci(1)) +
  facet_wrap(~ window, scales = "free_y", ncol = 3) +
  scale_y_reordered() +
  labs(x = "Frequency", y = NULL) +
  theme_sci(base_size = 8)
```

## Diagnostics and interpretation {#sec-emerging-diagnostics}

- **Baseline effects**: A keyword going from 2 to 6 occurrences shows 200% growth but may not represent meaningful emergence. Set minimum frequency thresholds before computing growth rates.
- **Database artefacts**: A newly indexed journal or a batch import of conference proceedings can create spurious "bursts". Cross-check with independent evidence.
- **Burst duration**: Short bursts (1 year) may reflect a single influential paper or a special issue. Sustained bursts (3+ years) more likely indicate genuine emergence.
- **Window sensitivity**: Rolling-window results change with window width. Report the chosen width and test sensitivity with wider and narrower windows.

## Limitations and responsible use {#sec-emerging-limitations}

::: {.callout-warning}
## Limitations and responsible use

- **Emergence is retrospective.** By the time bibliometric methods detect emergence, the topic has already gained momentum. These methods identify "recently emerged" topics, not future breakthroughs.
- **Coverage bias.** OpenAlex coverage has expanded rapidly, meaning recent years have more records. Growth rates may reflect database growth rather than intellectual emergence [@priem2022].
- **Language of emergence.** English-language terms dominate; emerging topics in non-English literatures may be invisible until translated or adopted by English-language publications.
- **Not predictive.** Burst detection identifies past patterns. It does not predict which topics will continue growing. Many bursts are transient [@kleinberg2003].
- **Strategic gaming.** If emergence metrics are used for funding decisions, researchers may strategically choose "hot" keywords, creating feedback loops [@hicks2015].
:::

## Common pitfalls {#sec-emerging-pitfalls}

::: {.callout-important}
## Common pitfalls

- **Confusing absolute growth with relative emergence.** A field that publishes 10,000 papers/year growing by 5% adds 500 papers. A niche area growing from 10 to 30 papers shows 200% growth. The niche area is "emerging" but the large field is adding more absolute volume.
- **Not controlling for corpus growth.** If the overall number of publications in the corpus grows, all keywords will grow. Use relative frequency (keyword count / total documents) rather than absolute counts.
- **Overreacting to single-year spikes.** One conference proceedings volume or special issue can create a one-year spike. Require multi-year sustained growth before declaring emergence.
- **Ignoring synonyms.** "Deep learning" and "neural networks" may both refer to the same emerging trend. Merge synonyms before analysis.
:::

## Exercises {#sec-emerging-exercises}

1. **Relative burst detection.** Modify the burst detection to use relative frequency (keyword proportion) instead of absolute counts. Do the same terms still burst?

2. **Emergence by journal.** Compare the most frequent emerging keywords in two different journals. Do they share emerging topics or focus on different frontiers?

3. **Novelty scoring.** For each document, compute its cosine distance to the centroid of all prior-year documents (using TF-IDF vectors). Plot novelty scores over time. Are recent papers becoming more or less novel?

4. **Validation.** Select 5 terms identified as "emerging" by burst detection. Search for qualitative evidence (review articles, conference themes) confirming their emergence. What is the false positive rate?

::: {.callout-tip collapse="true"}
## Solutions

Solutions are provided in @sec-cheatsheets.
:::

## Further reading {#sec-emerging-reading}

- @kleinberg2003 --- Bursty patterns in temporal data: the foundational burst detection algorithm.
- @price1963 --- Early observations on the growth of scientific literature and research fronts.
- @priem2022 --- OpenAlex temporal coverage patterns relevant to emergence detection.
- @callon1983 --- Co-word dynamics; an early approach to tracking topical evolution.

## Session info {#sec-emerging-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
