---
title: "Interactive Dashboards"
---

## Learning objectives {#dashboard-objectives}

After completing this chapter, you will be able to:

- Create a Quarto dashboard layout with multiple panels and pages
- Convert ggplot2 figures to interactive plotly charts
- Build searchable, filterable data tables with the DT package
- Design effective dashboard layouts for bibliometric data exploration
- Deploy dashboards as self-contained HTML files

## Setup {#dashboard-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
library(plotly)
library(DT)
library(glue)

set.seed(20260509)

source(here::here("R", "api_helpers.R"))
source(here::here("R", "utils.R"))
source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#dashboard-background}

Bibliometric analysis often produces more data than can be communicated through static figures and tables. Interactive dashboards allow stakeholders — research managers, librarians, policy makers — to explore data on their own terms: filtering by year, hovering for details, sorting tables, and drilling down into specific subsets.

**Quarto dashboards** use the `format: dashboard` output type to arrange content in cards, rows, and columns. They support static and interactive content side by side, and can be deployed as self-contained HTML files (no server needed) or as server-backed applications.

**plotly** converts ggplot2 figures into interactive JavaScript visualisations with hover tooltips, zooming, panning, and selection. The `ggplotly()` function provides a zero-effort upgrade path from static to interactive figures, though fine-tuning often requires direct plotly syntax.

**DT** (DataTables) renders data frames as interactive HTML tables with search, sort, pagination, and filtering. For bibliometric data, DT is invaluable: users can search for specific authors, sort by citation count, and filter by year — all without writing code.

The key design principle for dashboards is **progressive disclosure**: show the most important summary first, then let users drill down into details. Avoid overwhelming users with every metric at once.

## Worked example {#dashboard-example}

### Preparing dashboard data

```{r}
#| label: fetch-dashboard-data
#| cache: true
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2019-01-01",
  to_publication_date = "2023-12-31",
  options = list(sample = 500, seed = 42)
)

dash_data <- works |>
  transmute(
    id, title = display_name,
    year = year(publication_date),
    cited_by_count, oa_status, type, source_display_name
  )

cat(glue("Dashboard data: {nrow(dash_data)} records\n"))
```

### Interactive trend chart

```{r}
#| label: fig-dashboard-trend
#| fig-cap: "Interactive publication trend with hover details."
#| fig-alt: "Interactive line chart showing annual publication counts with tooltips on hover."
#| fig-width: 6
#| fig-height: 3.5
#| eval: !expr knitr::is_html_output(excludes = c("epub","epub2","epub3"))
# Interactive plotly widget — only rendered for HTML output.
trend <- dash_data |>
  count(year)

p <- ggplot(trend, aes(x = year, y = n)) +
  geom_line(colour = palette_sci(1), linewidth = 1) +
  geom_point(colour = palette_sci(1), size = 3) +
  labs(x = "Year", y = "Publications") +
  theme_sci()

ggplotly(p)
```

### Interactive citation distribution

```{r}
#| label: fig-dashboard-citedist
#| fig-cap: "Interactive citation distribution with hover details."
#| fig-alt: "Interactive histogram of citation counts where hovering reveals bin details."
#| fig-width: 6
#| fig-height: 3.5
#| eval: !expr knitr::is_html_output(excludes = c("epub","epub2","epub3"))
# Interactive plotly widget — only rendered for HTML output.
p2 <- ggplot(dash_data, aes(x = cited_by_count)) +
  geom_histogram(binwidth = 5, fill = palette_sci(1), colour = "white") +
  labs(x = "Citation count", y = "Papers") +
  theme_sci()

ggplotly(p2)
```

### Searchable data table

```{r}
#| label: data-table
#| eval: !expr knitr::is_html_output(excludes = c("epub","epub2","epub3"))
# Interactive DT widget — only rendered for HTML output.
dash_data |>
  select(title, year, cited_by_count, oa_status) |>
  arrange(desc(cited_by_count)) |>
  datatable(
    options = list(pageLength = 10, searchHighlight = TRUE),
    filter = "top",
    rownames = FALSE
  )
```

### OA status breakdown (interactive)

```{r}
#| label: fig-dashboard-oa
#| fig-cap: "Interactive OA status breakdown by year."
#| fig-alt: "Interactive stacked bar chart showing OA status proportions by year."
#| fig-width: 6
#| fig-height: 4
#| eval: !expr knitr::is_html_output(excludes = c("epub","epub2","epub3"))
# Interactive plotly widget — only rendered for HTML output.
oa_trend <- dash_data |>
  count(year, oa_status)

p3 <- ggplot(oa_trend, aes(x = factor(year), y = n, fill = oa_status)) +
  geom_col() +
  scale_fill_manual(values = palette_sci(n_distinct(oa_trend$oa_status))) +
  labs(x = "Year", y = "Count", fill = "OA Status") +
  theme_sci()

ggplotly(p3)
```

## Diagnostics and interpretation {#dashboard-diagnostics}

- **Performance**: Large datasets (> 10,000 rows) can make plotly and DT slow. Aggregate or sample data before rendering interactive widgets.
- **Accessibility**: Interactive figures are not accessible to screen readers. Always provide alt text and consider static fallbacks.
- **File size**: Self-contained HTML dashboards embed all data and JavaScript libraries. Files can reach 5–50 MB for complex dashboards.
- **Browser compatibility**: Test in Chrome, Firefox, and Safari. Some plotly features render differently across browsers.

## Limitations and responsible use {#dashboard-limitations}

<div class="callout callout-warn">
## Limitations and responsible use

- **Interactivity is not analysis.** A dashboard lets users explore data, but exploration without statistical rigour can produce misleading impressions. Pair dashboards with proper analysis.
- **Dashboards can mislead.** Default axis ranges, binning, and colour choices affect interpretation. Design dashboards to prevent misreading, not enable it.
- **Data privacy.** Interactive tables may expose individual-level data (author names, paper titles). Consider whether the dashboard's audience should see this level of detail.
- **Maintenance burden.** Dashboards that query live APIs need updating when APIs change. Static dashboards based on cached data are more sustainable [@hicks2015].
</div>

## Common pitfalls {#dashboard-pitfalls}

<div class="callout callout-stop">
## Common pitfalls

- **Too many widgets.** A dashboard with 15 interactive panels is overwhelming. Start with 3–5 and add more only if users need them.
- **Not filtering large datasets.** Rendering 50,000 rows in DT or 100,000 points in plotly will crash the browser. Pre-aggregate.
- **Ignoring PDF fallbacks.** If your Quarto book renders to PDF, interactive widgets become static. Ensure the static version is still informative.
- **Using ggplotly without cleanup.** `ggplotly()` produces verbose tooltips. Customise hover text with `tooltip` parameter or direct plotly.
</div>

## Exercises {#dashboard-exercises}

1. **Multi-page dashboard.** Create a Quarto dashboard (format: dashboard) with two pages: "Overview" (trend + citation distribution) and "Details" (searchable table).

2. **Linked brushing.** Use plotly's selection features to highlight papers in a scatter plot and simultaneously filter a DT table below.

3. **Custom tooltips.** Modify the trend chart to show "Year: 2023, Papers: 87, Mean cites: 12.3" on hover instead of the default tooltip.

<div class="callout callout-ok">
## Solutions

Solutions are provided in \@ref(solutions).
</div>

## Further reading {#dashboard-reading}

- @aria2017 --- `bibliometrix` and `biblioshiny` for interactive bibliometric exploration.
- @priem2022 --- OpenAlex as a data source for dashboard construction.

## Session info {#dashboard-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
