---
title: "Building a Bibliometric Shiny App"
---

## Learning objectives {#shiny-objectives}

After completing this chapter, you will be able to:

- Design a Shiny app architecture for bibliometric analysis
- Implement reactive data fetching from OpenAlex
- Build interactive UI components: dropdowns, sliders, and tabbed panels
- Create server-side logic for citation analysis and network visualisation
- Deploy a Shiny app to shinyapps.io or a self-hosted server

## Setup {#shiny-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(shiny)
library(openalexR)
library(glue)

set.seed(20260509)

source(here::here("R", "sci_palette.R"))
```

## Conceptual background {#shiny-background}

Shiny transforms R scripts into web applications. Users interact through widgets (dropdowns, sliders, buttons) and see results update in real time. For bibliometrics, Shiny apps serve two audiences: researchers who want to explore data without writing code, and decision-makers who need accessible analytics dashboards.

A Shiny app has two components: the **UI** (user interface, defining layout and inputs) and the **server** (the R logic that reacts to user input and generates output). **Reactive programming** connects inputs to outputs: when a user changes a dropdown, only the computations that depend on that input are re-executed.

For bibliometric apps, the typical architecture involves:

1. **Input panel**: Select a journal, institution, or search query; choose date range and sample size.
2. **Data fetching**: Reactively query OpenAlex based on user selections.
3. **Analysis tabs**: Trend charts, citation distributions, author rankings, and optionally network visualisation.
4. **Export**: Download buttons for data (CSV) and figures (PNG).

**`biblioshiny()`** from the `bibliometrix` package [@aria2017] provides a ready-made Shiny interface for many standard analyses. Building a custom app gives full control over the analysis pipeline and user experience.

Deployment options include shinyapps.io (hosted by Posit, free tier available), Posit Connect (enterprise), and self-hosted via Shiny Server (open source).

## Worked example {#shiny-example}

### App structure

We demonstrate the key components of a bibliometric Shiny app. The code below shows the UI and server logic separately for clarity.

### UI definition

```{r}
#| label: shiny-ui
#| eval: false
ui <- fluidPage(
  titlePanel("Bibliometric Explorer"),

  sidebarLayout(
    sidebarPanel(
      textInput("journal_id", "OpenAlex Source ID:",
                value = "S148561398"),
      sliderInput("year_range", "Publication Years:",
                  min = 2015, max = 2023, value = c(2020, 2023)),
      numericInput("sample_size", "Sample Size:", value = 200,
                   min = 50, max = 1000),
      actionButton("fetch", "Fetch Data", class = "btn-primary"),
      hr(),
      downloadButton("download_csv", "Download CSV")
    ),

    mainPanel(
      tabsetPanel(
        tabPanel("Trends",
                 plotOutput("trend_plot"),
                 tableOutput("summary_table")),
        tabPanel("Citations",
                 plotOutput("cite_dist"),
                 plotOutput("cite_boxplot")),
        tabPanel("Data",
                 DT::dataTableOutput("data_table"))
      )
    )
  )
)
```

### Server logic

```{r}
#| label: shiny-server
#| eval: false
server <- function(input, output, session) {

  works_data <- eventReactive(input$fetch, {
    withProgress(message = "Fetching from OpenAlex...", {
      oa_fetch(
        entity = "works",
        primary_location.source.id = input$journal_id,
        from_publication_date = paste0(input$year_range[1], "-01-01"),
        to_publication_date = paste0(input$year_range[2], "-12-31"),
        type = "article",
        options = list(sample = input$sample_size, seed = 42)
      ) |>
        transmute(
          id, title = display_name,
          year = year(publication_date),
          cited_by_count, oa_status
        )
    })
  })

  output$trend_plot <- renderPlot({
    req(works_data())
    works_data() |>
      count(year) |>
      ggplot(aes(x = factor(year), y = n)) +
      geom_col(fill = palette_sci(1)) +
      labs(x = "Year", y = "Publications") +
      theme_sci()
  })

  output$cite_dist <- renderPlot({
    req(works_data())
    ggplot(works_data(), aes(x = cited_by_count)) +
      geom_histogram(binwidth = 5, fill = palette_sci(1), colour = "white") +
      labs(x = "Citations", y = "Count") +
      theme_sci()
  })

  output$cite_boxplot <- renderPlot({
    req(works_data())
    works_data() |>
      filter(!is.na(oa_status)) |>
      ggplot(aes(x = oa_status, y = cited_by_count + 1)) +
      geom_boxplot(fill = palette_sci(1), alpha = 0.7) +
      scale_y_log10() +
      labs(x = "OA Status", y = "Citations (log)") +
      theme_sci()
  })

  output$summary_table <- renderTable({
    req(works_data())
    works_data() |>
      summarise(
        Articles = n(),
        `Mean Citations` = round(mean(cited_by_count), 1),
        `Median Citations` = median(cited_by_count),
        `OA Rate` = scales::percent(mean(oa_status != "closed", na.rm = TRUE))
      )
  })

  output$data_table <- DT::renderDataTable({
    req(works_data())
    works_data() |>
      select(title, year, cited_by_count, oa_status) |>
      arrange(desc(cited_by_count))
  })

  output$download_csv <- downloadHandler(
    filename = function() paste0("biblio_data_", Sys.Date(), ".csv"),
    content = function(file) write_csv(works_data(), file)
  )
}
```

### Running the app

```{r}
#| label: run-app
#| eval: false
shinyApp(ui = ui, server = server)
```

### biblioshiny as an alternative

```{r}
#| label: biblioshiny
#| eval: false
# bibliometrix provides a ready-made Shiny interface:
# library(bibliometrix)
# biblioshiny()
```

## Diagnostics and interpretation {#shiny-diagnostics}

- **Responsiveness**: API calls can take several seconds. Use `withProgress()` to show loading indicators.
- **Error handling**: Wrap API calls in `tryCatch()` to display user-friendly error messages instead of stack traces.
- **Memory**: Each user session loads data into R memory. For apps with many concurrent users, pre-aggregate data or use database backends.
- **Testing**: Use `shinytest2` to automate testing of Shiny apps.

## Limitations and responsible use {#shiny-limitations}

<div class="callout callout-warn">
## Limitations and responsible use

- **Shiny apps are not analyses.** They enable exploration but do not substitute for rigorous, documented analytical pipelines. Use Shiny for discovery, not for producing final results.
- **API rate limits.** Live OpenAlex queries from a Shiny app can hit rate limits if many users fetch data simultaneously. Use caching or pre-fetched data for production apps.
- **Maintenance.** Shiny apps require ongoing maintenance: API changes, package updates, and server costs. Plan for long-term sustainability before deploying.
- **Interpretive guardrails.** Non-technical users may misinterpret metrics. Build in tooltips, descriptions, and warnings — especially around citation-based indicators [@hicks2015; @dora2012].
</div>

## Common pitfalls {#shiny-pitfalls}

<div class="callout callout-stop">
## Common pitfalls

- **Fetching data on every input change.** Use `eventReactive()` with an action button instead of `reactive()` to prevent re-fetching on every slider adjustment.
- **Not validating inputs.** Check that the journal ID exists and the date range is valid before making API calls.
- **Monolithic server functions.** Break the server logic into modules for maintainability as the app grows.
- **Deploying without testing.** Test with realistic data volumes before deploying. An app that works with 100 records may fail with 10,000.
</div>

## Exercises {#shiny-exercises}

1. **Add network tab.** Extend the app with a tab that builds and displays a co-authorship network using `visNetwork`.

2. **Caching.** Implement server-side caching so that repeated queries with the same parameters return instantly.

3. **Deployment.** Deploy the app to shinyapps.io. Test with different journal IDs and date ranges.

4. **biblioshiny comparison.** Run `biblioshiny()` and compare its features with your custom app. What does biblioshiny offer that your app does not?

<div class="callout callout-ok">
## Solutions

Solutions are provided in \@ref(solutions).
</div>

## Further reading {#shiny-reading}

- @aria2017 --- `bibliometrix` and `biblioshiny()` for turnkey bibliometric exploration.
- @priem2022 --- OpenAlex API for reactive data fetching.
- @hicks2015 --- The Leiden Manifesto: designing interfaces that prevent metric misuse.
- @dora2012 --- DORA: responsible assessment in interactive tools.

## Session info {#shiny-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
