---
title: "Chapter Title"
---

<!-- TEMPLATE: Copy this file to create a new chapter. Replace all placeholder
     content. Delete these HTML comments before committing. Every chapter must
     include all 10 sections below, in this order. -->

## Learning objectives {#CHID-objectives}

After completing this chapter, you will be able to:

- Objective 1 (use action verbs: compute, interpret, construct, evaluate, …)
- Objective 2
- Objective 3

## Setup {#CHID-setup}

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(openalexR)
# Add chapter-specific packages here

set.seed(20260509)
```

## Conceptual background {#CHID-background}

<!-- ≤500 words. Introduce the theoretical motivation and key concepts.
     Cite foundational literature from references.bib. -->

## Worked example {#CHID-example}

<!-- Full reproducible R workflow. All data from OpenAlex or bundled samples.
     Break into subsections as needed. -->

### Acquiring the data

```{r}
#| label: fetch-data
#| cache: true
# Example: fetch works from OpenAlex
# works <- fetch_openalex(
#   entity = "works",
#   search = "your query",
#   from_publication_date = "2020-01-01"
# )
```

### Analysis

```{r}
#| label: analyse
# Analysis code here
```

### Visualization

```{r}
#| label: fig-CHID-main
#| fig-cap: "Descriptive caption for the figure."
#| fig-alt: "Accessible text description of the figure content."
#| fig-width: 6
#| fig-height: 4.5
# ggplot code here, using theme_sci() and palette_sci()
```

## Diagnostics and interpretation {#CHID-diagnostics}

<!-- Discuss how to interpret the results. What do the numbers mean?
     What are the benchmarks? How to sanity-check the output? -->

## Limitations and responsible use {#CHID-limitations}

<div class="callout callout-warn">
## Limitations and responsible use

<!-- MANDATORY in every quantitative chapter. Discuss:
     - What this method cannot tell you
     - Known biases in the data or indicator
     - How the metric can be gamed or misused
     - Cite Leiden Manifesto / DORA where relevant -->
</div>

## Common pitfalls {#CHID-pitfalls}

<div class="callout callout-stop">
## Common pitfalls

<!-- MANDATORY. List 3–5 concrete mistakes practitioners make:
     - Pitfall 1
     - Pitfall 2
     - Pitfall 3 -->
</div>

## Exercises {#CHID-exercises}

1. **Exercise 1.** Description of task. *(Hint: …)*
2. **Exercise 2.** Description of task.
3. **Exercise 3.** Description of task.

<div class="callout callout-ok">
## Solutions

Solutions are provided in \@ref(solutions).
</div>

## Further reading {#CHID-reading}

<!-- Annotated list, ≤8 entries, all present in references.bib. -->

- @hicks2015 — The Leiden Manifesto; foundational reading on responsible metrics.

## Session info {#CHID-sessioninfo}

```{r}
#| label: session-info
sessionInfo()
```
