---
title: "R Setup, renv, and Docker"
---

## R installation

Install the latest R release from [CRAN](https://cran.r-project.org/). This book was developed with R 4.4.x. We recommend [RStudio](https://posit.co/download/rstudio-desktop/) or [VS Code with the R extension](https://marketplace.visualstudio.com/items?itemName=REditorSupport.r) as your IDE.

## Package management with renv

This project uses **renv** for reproducible package management. After cloning the repository:

```{r}
#| eval: false
install.packages("renv")
renv::restore()
```

This installs the exact package versions recorded in `renv.lock`. To update the lockfile after adding a new package:

```{r}
#| eval: false
renv::install("newpackage")
renv::snapshot()
```

## Core packages

The table below lists the primary packages used throughout the book.

| Package | Purpose | Chapter(s) |
|:--------|:--------|:-----------|
| openalexR | OpenAlex API access | 5–37 |
| tidyverse | Data wrangling and visualisation | All |
| igraph | Network analysis | 15–20 |
| tidygraph | Tidy network manipulation | 15–20 |
| ggraph | Network visualisation | 15–20 |
| quanteda | Text mining and corpus management | 21–25 |
| topicmodels | LDA topic modelling | 22, CS-5 |
| stm | Structural topic models | 22 |
| gt | Publication-quality tables | Throughout |
| targets | Pipeline orchestration | 34 |
| plotly | Interactive charts | 35 |
| DT | Interactive data tables | 35 |
| shiny | Web applications | 37 |
| word2vec | Word embeddings | 24 |
| uwot | UMAP dimensionality reduction | 24 |

## Docker environment

A `Dockerfile` is provided at the repository root. It builds on `rocker/verse:4.4.1` and installs all required system libraries and R packages.

```bash
docker build -t scientometrics-in-r .
docker run --rm -p 8787:8787 -e DISABLE_AUTH=true scientometrics-in-r
```

Then open `http://localhost:8787` in your browser to access RStudio Server with all dependencies pre-installed.

## Quarto

Install [Quarto CLI](https://quarto.org/docs/get-started/) version 1.4 or later. To render the book locally:

```bash
quarto render
```

Rendered output appears in `_book/`. For iterative development, use `quarto preview` to get live-reloading in the browser.
