# Cheatsheets and Quick Reference

<!-- verify_chunks: skip-file
     Cheatsheet snippets are reference code, not evaluated at build time. -->

## openalexR quick reference

```{r}
#| eval: false
library(openalexR)

# Fetch works by journal (source)
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  from_publication_date = "2020-01-01",
  to_publication_date = "2023-12-31",
  type = "article"
)

# Fetch works by institution
works <- oa_fetch(
  entity = "works",
  authorships.institutions.id = "I63966007",
  from_publication_date = "2022-01-01",
  type = "article"
)

# Fetch works by topic
works <- oa_fetch(
  entity = "works",
  topics.id = "T10102",
  from_publication_date = "2023-01-01"
)

# Fetch author profile
author <- oa_fetch(entity = "authors", id = "A5023888391")

# Fetch institution profile
inst <- oa_fetch(entity = "institutions", id = "I63966007")

# Sampling
works <- oa_fetch(
  entity = "works",
  primary_location.source.id = "S148561398",
  options = list(sample = 500, seed = 42)
)
```

## quanteda pipeline

```{r}
#| eval: false
library(quanteda)

# Full text-analysis pipeline
corp <- corpus(df, docid_field = "id", text_field = "text")
toks <- tokens(corp, remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en"))
dfmat <- dfm(toks) |>
  dfm_trim(min_termfreq = 5, min_docfreq = 3)

# TF-IDF weighting
dfmat_tfidf <- dfm_tfidf(dfmat)

# Top features
topfeatures(dfmat, 20)

# Keyword-in-context
kwic(toks, pattern = "open access", window = 5)
```

## igraph / tidygraph / ggraph

```{r}
#| eval: false
library(igraph)
library(tidygraph)
library(ggraph)

# Build graph from edge list
g <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)

# Tidygraph conversion
tg <- as_tbl_graph(g) |>
  activate(nodes) |>
  mutate(
    degree = centrality_degree(),
    between = centrality_betweenness(),
    community = group_leiden(resolution = 1.0)
  )

# ggraph visualisation
ggraph(tg, layout = "fr") +
  geom_edge_link(aes(alpha = weight), show.legend = FALSE) +
  geom_node_point(aes(size = degree, colour = factor(community))) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void()
```

## Common data wrangling patterns

```{r}
#| eval: false
library(tidyverse)

# Unnest OpenAlex authorships
authors <- works |>
  select(work_id = id, authorships) |>
  unnest(authorships, names_sep = "_")

# Unnest OpenAlex topics
topics <- works |>
  select(work_id = id, topics) |>
  unnest(topics, names_sep = "_")

# Compute h-index
compute_h_index <- function(citations) {
  citations <- sort(citations, decreasing = TRUE)
  sum(citations >= seq_along(citations))
}

# Field normalisation
works |>
  group_by(field, year) |>
  mutate(
    field_mean = mean(cited_by_count),
    mncs = cited_by_count / pmax(field_mean, 1)
  )

# Fractional counting
works |>
  unnest(authorships, names_sep = "_") |>
  group_by(id) |>
  mutate(frac = 1 / n()) |>
  ungroup()
```

## Companion package functions

The `scientometricsInR` package provides helper functions used throughout the book:

| Function | Purpose | Example |
|:---------|:--------|:--------|
| `fetch_openalex(...)` | Cached, rate-limited OpenAlex queries | `fetch_openalex(entity = "works", ...)` |
| `dedupe_by_doi(df)` | Remove duplicate records by DOI | `works |> dedupe_by_doi()` |
| `compute_h_index(citations)` | Compute h-index from citation vector | `compute_h_index(c(10, 8, 5, 3, 1))` → 3 |
| `field_normalize(cites, means)` | MNCS normalisation (vector) | `field_normalize(c(20, 5), c(10, 10))` → `c(2.0, 0.5)` |
| `compute_mncs(df)` | MNCS for a grouped data frame | `df |> compute_mncs()` |
| `build_coauth_graph(works)` | Co-authorship igraph from works | `build_coauth_graph(works)` |
| `kleinberg_bursts(kw, dates)` | Keyword burst detection | `kleinberg_bursts(keywords, dates)` |
| `palette_sci(n)` | Viridis colour palette | `palette_sci(4)` |
| `theme_sci()` | Minimal ggplot2 theme | `+ theme_sci()` |

## Useful keyboard shortcuts

| Action | RStudio | VS Code |
|:-------|:--------|:--------|
| Run current line/selection | Ctrl+Enter | Ctrl+Enter |
| Run current chunk | Ctrl+Shift+Enter | Ctrl+Shift+Enter |
| Insert pipe `|>` | Ctrl+Shift+M | — |
| Insert assignment `<-` | Alt+- | — |
| Render document | Ctrl+Shift+K | — |
| Go to file | Ctrl+. | Ctrl+P |
