# Exercise Solutions

<!-- verify_chunks: skip-file
     Every chunk in this appendix is intentionally not evaluated; solutions
     reference data built during the corresponding chapter. -->

# Exercise Solutions {#solutions}

This appendix provides suggested solutions for all chapter exercises.
The code sketches below are meant to guide you toward a working answer; they are
not the only valid approach.
Because each solution references data built during the corresponding chapter,
code chunks are **not evaluated** here --- run them inside the chapter project
after completing the worked example.

## Chapter 1 --- What Is Scientometrics? {-}

**1. Explore a different field.**
Swap the search term and fetch a fresh sample. Comparing the publication trend
and top journals reveals how different fields vary in volume, growth rate, and
journal concentration.

```{r}
#| label: sol-ch01-1
#| eval: false
library(openalexR)
library(tidyverse)

works_ml <- oa_fetch(
 entity = "works",
 search = "machine learning",
 from_publication_date = "2013-01-01",
 to_publication_date = "2023-12-31",
 count_only = FALSE,
 per_page = 200
) |> slice_sample(n = 500)

works_ml |>
 count(publication_year) |>
 ggplot(aes(publication_year, n)) +
 geom_col() +
 labs(title = "Machine-learning publications per year",
      x = "Year", y = "Works")
```

*Key insight:* High-growth fields like machine learning show steeper annual
increases and a broader spread across journals compared to a niche field like
scientometrics.

**2. Uncited works.**
Filter to zero-citation records and examine the year distribution.

```{r}
#| label: sol-ch01-2
#| eval: false
works |>
 filter(cited_by_count == 0) |>
 count(publication_year) |>
 mutate(pct = n / sum(n)) |>
 ggplot(aes(publication_year, pct)) +
 geom_col() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "Distribution of uncited works by year",
      x = "Year", y = "Share of uncited works")
```

*Key insight:* Recent publication years dominate because papers need time to
accumulate citations --- a window effect.

**3. Open Access status.**
Compute yearly OA proportions to reveal the trend.

```{r}
#| label: sol-ch01-3
#| eval: false
works |>
 mutate(year = publication_year) |>
 group_by(year) |>
 summarise(oa_share = mean(is_oa, na.rm = TRUE)) |>
 ggplot(aes(year, oa_share)) +
 geom_line() +
 geom_point() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "Open-access share over time",
      x = "Year", y = "OA proportion")
```

*Key insight:* OA proportions have risen steadily, driven by funder mandates and
the growth of gold OA journals.

**4. Growth curve.**
Plot annual counts on a log scale; linearity implies exponential growth.

```{r}
#| label: sol-ch01-4
#| eval: false
works |>
 filter(publication_year >= 2000, publication_year <= 2023) |>
 count(publication_year) |>
 ggplot(aes(publication_year, n)) +
 geom_point() +
 geom_smooth(method = "lm") +
 scale_y_log10() +
 labs(title = "Publication growth (log scale)",
      x = "Year", y = "Works (log)")
```

*Key insight:* If the points fall roughly on a straight line, growth is
approximately exponential, consistent with Price's law.

## Chapter 2 --- Theoretical Underpinnings {-}

**1. Lotka exponent.**
Fit a linear model in log-log space to estimate the exponent.

```{r}
#| label: sol-ch02-1
#| eval: false
library(tidyverse)

author_prod <- works |>
 unnest(author) |>
 count(au_id, name = "n_papers")

freq_tbl <- author_prod |>
 count(n_papers, name = "n_authors")

fit <- lm(log(n_authors) ~ log(n_papers), data = freq_tbl)
coef(fit)  # slope ≈ -2 if Lotka holds

ggplot(freq_tbl, aes(n_papers, n_authors)) +
 geom_point() +
 scale_x_log10() +
 scale_y_log10() +
 geom_smooth(method = "lm", se = FALSE) +
 labs(title = "Lotka's law fit",
      x = "Papers per author (log)", y = "Number of authors (log)")
```

*Key insight:* Lotka predicted an exponent near 2; empirical values typically
range from 1.5 to 3.5, depending on the field and time window.

**2. Bradford zones.**
Divide journals into three zones of roughly equal article counts.

```{r}
#| label: sol-ch02-2
#| eval: false
journal_counts <- works |>
 count(so, sort = TRUE) |>
 mutate(cum = cumsum(n),
        total = sum(n),
        zone = case_when(
         cum <= total / 3 ~ "Core",
         cum <= 2 * total / 3 ~ "Mid",
         TRUE ~ "Periphery"
        ))

journal_counts |> count(zone)
```

*Key insight:* The core zone contains very few journals; the periphery contains
many. The ratio between successive zone sizes should approximate Bradford's
multiplier.

**3. Your field's Zipf distribution.**
Extract title words and plot rank vs. frequency on log-log axes.

```{r}
#| label: sol-ch02-3
#| eval: false
library(tidytext)

word_freq <- works |>
 select(id, display_name) |>
 unnest_tokens(word, display_name) |>
 anti_join(stop_words, by = "word") |>
 count(word, sort = TRUE) |>
 mutate(rank = row_number())

ggplot(word_freq, aes(rank, n)) +
 geom_line() +
 scale_x_log10() +
 scale_y_log10() +
 labs(title = "Zipf's law for title words",
      x = "Rank (log)", y = "Frequency (log)")
```

*Key insight:* A roughly linear relationship on log-log axes confirms Zipf's
law. Domain-specific terms dominate the top ranks.

**4. Matthew Effect simulation.**
Simulate preferential attachment and compare the resulting distribution to
Lotka's law.

```{r}
#| label: sol-ch02-4
#| eval: false
set.seed(20260509)
n_authors <- 100
papers <- rep(1, n_authors)

for (step in seq_len(50)) {
 probs <- papers / sum(papers)
 winner <- sample(n_authors, 1, prob = probs)
 papers[winner] <- papers[winner] + 1
}

tibble(papers = papers) |>
 count(papers, name = "n_authors") |>
 ggplot(aes(papers, n_authors)) +
 geom_point() +
 scale_x_log10() +
 scale_y_log10() +
 labs(title = "Simulated preferential attachment",
      x = "Papers (log)", y = "Authors (log)")
```

*Key insight:* Preferential attachment generates a skewed distribution
resembling Lotka's law, illustrating the Matthew Effect.

## Chapter 3 --- Ethics and Responsible Metrics {-}

**1. Leiden Manifesto audit.**
This is a qualitative exercise. Evaluate a real hiring policy against the ten
principles.

```{r}
#| label: sol-ch03-1
#| eval: false
# No code required. Suggested structure:
# For each of the 10 Leiden Manifesto principles:
#   1. State the principle
#   2. Quote the relevant policy text (if any)
#   3. Assess: satisfied / partially satisfied / violated
#   4. Explain your reasoning
# Summarise overall compliance in a paragraph.
```

*Key insight:* Most policies rely heavily on the h-index or JIF, violating
principles 6 (field variation) and 7 (qualitative portfolio assessment).

**2. Self-citation across fields.**
Compare self-citation rates between two different fields.

```{r}
#| label: sol-ch03-2
#| eval: false
library(openalexR)
library(tidyverse)

fields <- c("bibliometrics", "genomics")
results <- map(fields, \(term) {
 w <- oa_fetch(entity = "works", search = term,
               from_publication_date = "2020-01-01",
               to_publication_date = "2023-12-31") |>
   slice_sample(n = 200)
 w |> mutate(field = term)
}) |> list_rbind()

# Self-citation proxy: papers citing other papers by the same first author
results |>
 group_by(field) |>
 summarise(mean_cites = mean(cited_by_count, na.rm = TRUE),
           median_cites = median(cited_by_count, na.rm = TRUE))
```

*Key insight:* Self-citation baselines differ by field. Higher self-citation in
some fields is structural (smaller communities) rather than manipulative.

**3. DORA compliance check.**
A research exercise using the DORA signatory database.

```{r}
#| label: sol-ch03-3
#| eval: false
# No code required. Visit https://sfdora.org/signers/
# Select three major funders and compare their grant evaluation
# criteria against DORA commitments:
#   - Do they avoid using JIF in evaluation?
#   - Do they assess research outputs on their own merit?
#   - Do they consider a broad range of impact measures?
```

*Key insight:* Many signatories still reference journal prestige in practice,
revealing a gap between stated commitments and implementation.

**4. Goodhart's Law thought experiment.**
Propose a new indicator and analyse potential gaming strategies.

```{r}
#| label: sol-ch03-4
#| eval: false
# Example indicator: "Replication Impact Score"
#   = fraction of a researcher's papers that have been independently replicated
#
# Potential gaming strategies:
#   - Publish trivially replicable results
#   - Arrange replication exchanges with collaborators
#   - Focus on small, easily replicated studies rather than ambitious work
#
# The numerator (replications) and denominator (total papers) are both gameable.
# This illustrates Goodhart's Law: once a measure becomes a target,
# it ceases to be a good measure.
```

*Key insight:* Any ratio-based metric can be gamed on both numerator and
denominator. Indicator design must anticipate behavioural responses.

## Chapter 4 --- Data Sources Compared {-}

**1. Coverage gap analysis.**
Query the same DOIs in OpenAlex and Crossref and compare.

```{r}
#| label: sol-ch04-1
#| eval: false
library(openalexR)
library(rcrossref)
library(tidyverse)

dois <- works |> slice_sample(n = 20) |> pull(doi) |> na.omit()

oa_results <- map(dois, \(d) {
 tryCatch(oa_fetch(entity = "works", doi = d), error = \(e) NULL)
}) |> compact()
oa_found <- length(oa_results)

cr_results <- map(dois, \(d) {
 tryCatch(cr_works(d)$data, error = \(e) NULL)
}) |> compact()
cr_found <- length(cr_results)

tibble(
 Source = c("OpenAlex", "Crossref"),
 Found = c(oa_found, cr_found),
 Total = length(dois),
 Coverage = Found / Total
)
```

*Key insight:* OpenAlex generally has high DOI coverage because it ingests
Crossref data, but abstract availability may differ.

**2. Disciplinary bias.**
Compare indexed works across contrasting fields.

```{r}
#| label: sol-ch04-2
#| eval: false
fields <- list(
 biology = "https://openalex.org/C86803240",
 philosophy = "https://openalex.org/C138885662"
)

coverage <- imap(fields, \(concept_id, name) {
 oa_fetch(entity = "works", concept.id = concept_id,
          count_only = TRUE) |>
   mutate(field = name)
}) |> list_rbind()

coverage
```

*Key insight:* STEM fields are substantially better covered than humanities,
reflecting the historical journal-centricity of bibliometric databases.

**3. Citation count concordance.**
Correlate citation counts from two sources.

```{r}
#| label: sol-ch04-3
#| eval: false
library(rcrossref)

comparison <- works |>
 slice_sample(n = 50) |>
 filter(!is.na(doi)) |>
 rowwise() |>
 mutate(cr_cites = tryCatch(
   cr_works(doi)$data$is_referenced_by_count,
   error = \(e) NA_integer_
 )) |>
 ungroup()

cor(comparison$cited_by_count, comparison$cr_cites,
    use = "complete.obs", method = "spearman")

ggplot(comparison, aes(cited_by_count, cr_cites)) +
 geom_point() +
 geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
 labs(x = "OpenAlex citations", y = "Crossref citations",
      title = "Citation count concordance")
```

*Key insight:* Counts correlate strongly but are not identical; differences
arise from coverage scope and update frequency.

**4. Preprint tracking.**
Check whether bioRxiv DOIs appear in OpenAlex with published links.

```{r}
#| label: sol-ch04-4
#| eval: false
biorxiv_dois <- c(
 "10.1101/2023.01.01.123456",
 "10.1101/2023.02.15.234567"
 # ... add 8 more real bioRxiv DOIs
)

preprint_status <- map(biorxiv_dois, \(d) {
 tryCatch({
   w <- oa_fetch(entity = "works", doi = d)
   tibble(doi = d, found = TRUE,
          has_published = !is.null(w$ids$doi))
 }, error = \(e) tibble(doi = d, found = FALSE, has_published = FALSE))
}) |> list_rbind()

mean(preprint_status$found)
mean(preprint_status$has_published)
```

*Key insight:* Most recent bioRxiv preprints appear in OpenAlex; the proportion
linked to a published version depends on recency and field.

**5. Source selection exercise.**
This is a qualitative exercise requiring no code.

```{r}
#| label: sol-ch04-5
#| eval: false
# Recommended answer structure:
# 1. OpenAlex: free, broad coverage, good for large-scale analysis
#    - Limitation: may under-represent regional journals
# 2. African Journals Online (AJOL): specialised coverage of
#    Sub-Saharan African journals
# 3. Combination approach: use OpenAlex for global context,
#    supplement with AJOL for regional coverage
#
# Justification should address:
#   - Coverage of nursing literature
#   - Representation of Sub-Saharan African journals
#   - Cost (OpenAlex is free; Scopus/WoS require subscriptions)
#   - Reproducibility (OpenAlex is open; others are proprietary)
```

*Key insight:* No single source is sufficient; combining an open global database
with a regional index maximises both coverage and reproducibility.

## Chapter 5 --- APIs and Packages {-}

**1. Fetch works by institution.**
Find your institution in OpenAlex and retrieve its works.

```{r}
#| label: sol-ch05-1
#| eval: false
library(openalexR)
library(tidyverse)

# Step 1: find the institution
my_inst <- oa_fetch(entity = "institutions",
                    search = "University of Example")
my_inst_id <- my_inst$id[1]

# Step 2: fetch works
inst_works <- oa_fetch(entity = "works",
                       authorships.institutions.lineage = my_inst_id,
                       from_publication_date = "2020-01-01")

nrow(inst_works)
mean(!is.na(inst_works$doi))  # DOI coverage rate
```

*Key insight:* DOI coverage varies by institution; institutions publishing more
in regional or non-English journals tend to have lower DOI rates.

**2. Compare Crossref metadata.**
Retrieve the same work from both APIs and compare.

```{r}
#| label: sol-ch05-2
#| eval: false
library(rcrossref)

sample_doi <- inst_works$doi[1]

oa_meta <- oa_fetch(entity = "works", doi = sample_doi)
cr_meta <- cr_works(sample_doi)$data

tibble(
 field = c("citations", "title_match"),
 OpenAlex = c(oa_meta$cited_by_count,
              oa_meta$display_name),
 Crossref = c(cr_meta$is_referenced_by_count,
              cr_meta$title)
)
```

*Key insight:* Citation counts differ because each database has a different
citation graph. Crossref counts only DOI-based citations; OpenAlex resolves more.

**3. Build a multi-year corpus.**
Fetch all works from a journal across multiple years.

```{r}
#| label: sol-ch05-3
#| eval: false
journal_works <- oa_fetch(
 entity = "works",
 primary_location.source.id = "S4210174107",  # example journal ID
 from_publication_date = "2015-01-01",
 to_publication_date = "2023-12-31"
)

journal_works |>
 group_by(publication_year) |>
 summarise(n = n(), median_cites = median(cited_by_count)) |>
 pivot_longer(c(n, median_cites)) |>
 ggplot(aes(publication_year, value)) +
 geom_line() +
 geom_point() +
 facet_wrap(~name, scales = "free_y") +
 labs(title = "Journal trend: publications and citations",
      x = "Year", y = NULL)
```

*Key insight:* Median citation counts typically decrease for recent years due to
the citation time window, while publication counts may rise if the journal is
growing.

## Chapter 6 --- Parsing Native Exports {-}

**1. Export and parse.**
Parse a BibTeX file from Google Scholar using `bib2df`.

```{r}
#| label: sol-ch06-1
#| eval: false
library(bib2df)
library(tidyverse)

bib <- bib2df("scholar_export.bib")
ncol(bib)
map_int(bib, \(x) sum(!is.na(x))) |>
 enframe("field", "populated") |>
 arrange(desc(populated))
```

*Key insight:* Google Scholar BibTeX exports often lack abstracts and keywords,
making them less useful for text-based analyses.

**2. Format comparison.**
Parse the same records from CSV and RIS formats and compare fields.

```{r}
#| label: sol-ch06-2
#| eval: false
library(bibliometrix)

csv_data <- read_csv("scopus_export.csv")
ris_data <- convert2df("scopus_export.ris", dbsource = "scopus",
                       format = "plaintext")

cat("CSV columns:", ncol(csv_data), "\n")
cat("RIS columns:", ncol(ris_data), "\n")

# Compare overlapping fields
intersect(names(csv_data), names(ris_data))
```

*Key insight:* RIS files typically include structured reference data and
keywords that CSV exports may omit; CSVs are easier to manipulate in R.

**3. Standardization function.**
Write a reusable function to normalise columns.

```{r}
#| label: sol-ch06-3
#| eval: false
standardise_biblio <- function(df) {
 tibble(
   doi       = df$DI %||% NA_character_,
   title     = df$TI %||% df$Title %||% NA_character_,
   authors   = df$AU %||% NA_character_,
   year      = as.integer(df$PY %||% df$Year),
   journal   = df$SO %||% df$`Source title` %||% NA_character_,
   cited_by  = as.integer(df$TC %||% df$`Cited by` %||% 0L)
 )
}

# Usage:
# standardise_biblio(ris_data)
```

*Key insight:* A standardisation layer decouples downstream analysis from the
input format, improving reproducibility.

## Chapter 7 --- Building Reproducible Corpora {-}

**1. DOI coverage by year.**
Compute the DOI coverage rate per publication year.

```{r}
#| label: sol-ch07-1
#| eval: false
library(tidyverse)

works |>
 group_by(publication_year) |>
 summarise(doi_rate = mean(!is.na(doi))) |>
 ggplot(aes(publication_year, doi_rate)) +
 geom_line() +
 geom_point() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "DOI coverage by year", x = "Year", y = "Coverage")
```

*Key insight:* DOI coverage tends to improve for recent years as DOI assignment
became standard practice after the mid-2000s.

**2. Name variants.**
Detect potential author name duplicates.

```{r}
#| label: sol-ch07-2
#| eval: false
library(stringdist)

author_names <- works |>
 unnest(author) |>
 distinct(au_id, au_display_name) |>
 pull(au_display_name)

# Find pairs with low string distance
combos <- combn(author_names[1:200], 2, simplify = FALSE)
similar <- map(combos, \(pair) {
 d <- stringdist(pair[1], pair[2], method = "jw")
 if (d < 0.15) tibble(name1 = pair[1], name2 = pair[2], dist = d)
 else NULL
}) |> compact() |> list_rbind()

similar |> arrange(dist) |> head(20)
```

*Key insight:* Common heuristics include Jaro-Winkler distance, initial
matching, and shared co-author overlap. OpenAlex author IDs help but are not
always correct.

**3. ROR lookup.**
Use the ROR API to resolve institutions.

```{r}
#| label: sol-ch07-3
#| eval: false
library(httr2)

top_insts <- works |>
 unnest(authorships) |>
 unnest(institutions) |>
 count(display_name, sort = TRUE) |>
 head(5)

ror_lookup <- function(name) {
 resp <- request("https://api.ror.org/organizations") |>
   req_url_query(query = name) |>
   req_perform() |>
   resp_body_json()
 if (length(resp$items) > 0)
   tibble(name = name, ror_id = resp$items[[1]]$id,
          ror_name = resp$items[[1]]$name)
 else
   tibble(name = name, ror_id = NA, ror_name = NA)
}

map(top_insts$display_name, ror_lookup) |> list_rbind()
```

*Key insight:* ROR provides stable, disambiguated identifiers; matching by name
alone can produce false positives for common institutional names.

## Chapter 8 --- Working with Full Text {-}

**1. Extract and compare.**
Download a PDF, extract text, and compare the abstract.

```{r}
#| label: sol-ch08-1
#| eval: false
library(pdftools)
library(tidyverse)

pdf_text_raw <- pdf_text("sample_paper.pdf")
extracted_abstract <- str_extract(
 paste(pdf_text_raw, collapse = "\n"),
 "(?i)abstract\\s*\\.?\\s*(.+?)(?=\\n\\s*(introduction|keywords|1\\.))",
)

# Compare with OpenAlex abstract
oa_abstract <- works |> filter(doi == "10.xxxx/example") |> pull(ab)

cat("Extracted:\n", str_trunc(extracted_abstract, 300), "\n")
cat("OpenAlex:\n", str_trunc(oa_abstract, 300), "\n")
```

*Key insight:* PDF extraction often introduces line-break artefacts and header/
footer noise. The OpenAlex abstract (from publisher metadata) is typically
cleaner.

**2. Section detection.**
Write a function to split extracted text into sections.

```{r}
#| label: sol-ch08-2
#| eval: false
split_sections <- function(text) {
 lines <- str_split(text, "\n")[[1]]
 # Match numbered headings like "1. Introduction" or "2 Methods"
 heading_idx <- str_which(lines, "^\\s*\\d+\\.?\\s+[A-Z]")

 if (length(heading_idx) < 2) return(list(full_text = text))

 sections <- map2(heading_idx, c(heading_idx[-1] - 1, length(lines)),
                  \(start, end) {
   list(
     heading = str_trim(lines[start]),
     text = paste(lines[(start + 1):end], collapse = "\n")
   )
 })
 set_names(
   map_chr(sections, "text"),
   map_chr(sections, "heading")
 )
}

sections <- split_sections(paste(pdf_text_raw, collapse = "\n"))
names(sections)
```

*Key insight:* Heading detection is fragile; two-column layouts, supplementary
materials, and non-standard numbering all break simple regex patterns.

**3. Table extraction.**
Extract a table from a PDF using `tabulizer`.

```{r}
#| label: sol-ch08-3
#| eval: false
# tabulapdf needs Java (rJava); shown for reference, not executed.
# Install: install.packages("tabulapdf"); requires a working Java runtime.

tables <- extract_tables("sample_paper.pdf", pages = 3)
tbl <- as_tibble(tables[[1]], .name_repair = "unique")

# Typical cleaning steps:
tbl <- tbl |>
 row_to_names(row_number = 1) |>   # janitor::row_to_names
 mutate(across(where(is.character), str_trim)) |>
 filter(if_any(everything(), \(x) x != ""))
```

*Key insight:* Extracted tables almost always need header row promotion,
whitespace trimming, and removal of empty or merged-cell artefact rows.

## Chapter 9 --- Descriptive Bibliometrics {-}

**1. Lotka's law.**
Fit Lotka's law using `bibliometrix`.

```{r}
#| label: sol-ch09-1
#| eval: false
library(bibliometrix)
library(tidyverse)

bib_df <- convert2df(works)  # or use the already-converted object
lotka_result <- lotka(biblioAnalysis(bib_df))

# Plot observed vs expected
plot(lotka_result)
```

*Key insight:* If the Kolmogorov-Smirnov test p-value is above 0.05, the
observed distribution is consistent with the theoretical inverse-square law.

**2. Bradford's law.**
Identify core, mid, and peripheral journal zones.

```{r}
#| label: sol-ch09-2
#| eval: false
library(openalexR)
library(bibliometrix)

ml_works <- oa_fetch(entity = "works", search = "machine learning",
                     from_publication_date = "2020-01-01",
                     to_publication_date = "2023-12-31") |>
 slice_sample(n = 500)

bib_df <- convert2df(ml_works)
brad <- bradford(biblioAnalysis(bib_df))
brad$table |> head(20)
```

*Key insight:* The core zone typically contains very few journals that account
for a disproportionate share of articles --- Bradford's law in action.

**3. Three-field exploration.**
Use country instead of author name on the left field.

```{r}
#| label: sol-ch09-3
#| eval: false
library(bibliometrix)

threeFieldsPlot(bib_df, fields = c("AU_CO", "SO", "DE"),
                n = c(10, 10, 10))
```

*Key insight:* Country-level Sankey diagrams reveal geographic specialisation
patterns: certain countries cluster around specific journals and keywords.

**4. Annual growth rate.**
Compute the compound annual growth rate (CAGR).

```{r}
#| label: sol-ch09-4
#| eval: false
annual <- works |>
 count(publication_year) |>
 filter(publication_year >= 2010, publication_year <= 2023)

first_year <- annual$n[1]
last_year  <- annual$n[nrow(annual)]
n_years    <- nrow(annual) - 1

cagr <- (last_year / first_year)^(1 / n_years) - 1
cat(sprintf("CAGR: %.1f%%\n", cagr * 100))
```

*Key insight:* A positive CAGR indicates the field is growing; compare against
the global average (~4-5% for all scientific literature) to assess relative
growth.

## Chapter 10 --- Productivity and Impact Indicators {-}

**1. Career h-index.**
Compute h-index, g-index, and m-quotient for a specific author.

```{r}
#| label: sol-ch10-1
#| eval: false
library(openalexR)
library(tidyverse)

author_works <- oa_fetch(
 entity = "works",
 author.id = "A5023888391",  # replace with target author
 per_page = 200
)

cites <- sort(author_works$cited_by_count, decreasing = TRUE)

# h-index
h <- sum(cites >= seq_along(cites))

# g-index
cum_cites <- cumsum(cites)
g <- max(which(cum_cites >= seq_along(cites)^2))

# m-quotient
career_years <- max(author_works$publication_year) -
               min(author_works$publication_year) + 1
m <- h / career_years

tibble(h_index = h, g_index = g, m_quotient = round(m, 2))
```

*Key insight:* The h-index rewards consistency, the g-index rewards highly cited
papers, and the m-quotient adjusts for career length.

**2. Self-citation sensitivity.**
Pseudocode for removing self-citations from the h-index.

```{r}
#| label: sol-ch10-2
#| eval: false
# Pseudocode:
# 1. For each paper by the author, fetch its citing works
# 2. Check if any citing work shares the same author ID
# 3. Count only non-self-citations
# 4. Recompute h-index with adjusted citation counts
#
# Implementation sketch (expensive API calls):
# author_id <- "A5023888391"
# adjusted_cites <- map_int(author_works$id, \(work_id) {
#   citing <- oa_fetch(entity = "works", cites = work_id)
#   citing |>
#     unnest(author) |>
#     filter(au_id != author_id) |>
#     distinct(id) |>
#     nrow()
# })
# h_adjusted <- sum(sort(adjusted_cites, decreasing = TRUE) >=
#                   seq_along(adjusted_cites))
```

*Key insight:* Self-citation removal typically reduces the h-index by 1-3 points
for most researchers but can have large effects for prolific self-citers.

**3. Bootstrap uncertainty for MNCS.**
Use bootstrap resampling to compute a confidence interval for MNCS.

```{r}
#| label: sol-ch10-3
#| eval: false
set.seed(20260509)

mncs_boot <- replicate(1000, {
 idx <- sample(nrow(works), replace = TRUE)
 boot_sample <- works[idx, ]
 compute_mncs(boot_sample$cited_by_count,
              boot_sample$publication_year)
})

quantile(mncs_boot, c(0.025, 0.975))
```

*Key insight:* Small samples produce wide confidence intervals; MNCS values
should always be reported with uncertainty estimates.

**4. PP(top 10%) vs. MNCS.**
Compare two indicators by year.

```{r}
#| label: sol-ch10-4
#| eval: false
yearly_indicators <- works |>
 group_by(publication_year) |>
 summarise(
   mncs = compute_mncs(cited_by_count, publication_year),
   pp_top10 = mean(cited_by_count >= quantile(cited_by_count, 0.9))
 )

ggplot(yearly_indicators, aes(mncs, pp_top10)) +
 geom_point() +
 geom_text(aes(label = publication_year), nudge_y = 0.01) +
 labs(x = "MNCS", y = "PP(top 10%)",
      title = "MNCS vs PP(top 10%) by year")
```

*Key insight:* The two indicators usually agree but can diverge when a few
extremely highly cited papers inflate MNCS without affecting the 10% threshold.

## Chapter 11 --- Journal Metrics {-}

**1. Four-year CiteScore proxy.**
Extend the citation window to four years.

```{r}
#| label: sol-ch11-1
#| eval: false
library(openalexR)
library(tidyverse)

journal_id <- "S4210174107"  # example journal
target_year <- 2023

# Fetch articles published in the 4-year window
window_works <- oa_fetch(
 entity = "works",
 primary_location.source.id = journal_id,
 from_publication_date = paste0(target_year - 4, "-01-01"),
 to_publication_date   = paste0(target_year - 1, "-12-31")
)

citescore_4yr <- sum(window_works$cited_by_count) / nrow(window_works)
cat(sprintf("4-year CiteScore proxy: %.2f\n", citescore_4yr))
```

*Key insight:* A wider window captures slower-to-cite fields more fairly but
dilutes the signal for fast-moving fields.

**2. Rank correlation.**
Compare JIF and SNIP proxies with Spearman correlation.

```{r}
#| label: sol-ch11-2
#| eval: false
journal_metrics <- works |>
 group_by(so) |>
 summarise(
   jif_proxy  = mean(cited_by_count[publication_year >= 2021]),
   snip_proxy = jif_proxy / mean(cited_by_count)  # simplified
 ) |>
 filter(!is.na(jif_proxy), !is.na(snip_proxy))

cor(journal_metrics$jif_proxy, journal_metrics$snip_proxy,
    method = "spearman")
```

*Key insight:* Rank correlation is typically positive but imperfect, showing that
field normalisation reorders journals.

**3. Field comparison.**
Compare journals across two disciplines.

```{r}
#| label: sol-ch11-3
#| eval: false
physics_journals <- c("S137773608", "S71048024")
sociology_journals <- c("S64468286", "S128732326")

compute_jif <- function(jid) {
 w <- oa_fetch(entity = "works",
               primary_location.source.id = jid,
               from_publication_date = "2021-01-01",
               to_publication_date = "2022-12-31")
 sum(w$cited_by_count) / nrow(w)
}

tibble(
 journal = c(physics_journals, sociology_journals),
 field = c(rep("Physics", 2), rep("Sociology", 2)),
 jif_proxy = map_dbl(journal, compute_jif)
) |>
 ggplot(aes(field, jif_proxy)) +
 geom_boxplot() +
 labs(title = "Raw JIF proxy by field")
```

*Key insight:* Raw JIF values are much higher in fields with faster citation
accumulation; SNIP-style normalisation reduces this gap.

**4. Self-citation share.**
Estimate the proportion of journal self-citations.

```{r}
#| label: sol-ch11-4
#| eval: false
# Approximation: among citing works of this journal's papers,
# how many are also published in the same journal?
journal_works <- oa_fetch(
 entity = "works",
 primary_location.source.id = journal_id,
 from_publication_date = "2021-01-01",
 to_publication_date = "2022-12-31"
)

# Sample a subset and check citing works
sample_ids <- journal_works |> slice_sample(n = 20) |> pull(id)

self_cite_rates <- map_dbl(sample_ids, \(wid) {
 citing <- tryCatch(
   oa_fetch(entity = "works", cites = wid),
   error = \(e) tibble()
 )
 if (nrow(citing) == 0) return(NA_real_)
 mean(citing$so == journal_works$so[1], na.rm = TRUE)
})

mean(self_cite_rates, na.rm = TRUE)
```

*Key insight:* Journal self-citation rates above 20-30% are considered unusually
high and may warrant investigation.

## Chapter 12 --- Author-Level Analysis {-}

**1. Profile comparison.**
Compare two researchers on multiple indicators.

```{r}
#| label: sol-ch12-1
#| eval: false
library(openalexR)
library(tidyverse)

ids <- c("A5023888391", "A5048491430")  # two author IDs

profiles <- map(ids, \(aid) {
 w <- oa_fetch(entity = "works", author.id = aid)
 cites <- sort(w$cited_by_count, decreasing = TRUE)
 tibble(
   author_id = aid,
   n_papers  = nrow(w),
   h_index   = sum(cites >= seq_along(cites)),
   mean_cites = mean(w$cited_by_count)
 )
}) |> list_rbind()

profiles
```

*Key insight:* An author with many papers but a modest h-index publishes
frequently but with less consistent impact; high mean citations suggest a few
highly cited papers.

**2. ORCID validation.**
Compare works fetched by OpenAlex ID vs. ORCID filtering.

```{r}
#| label: sol-ch12-2
#| eval: false
oa_by_id <- oa_fetch(entity = "works",
                     author.id = "A5023888391")

oa_by_orcid <- oa_fetch(entity = "works",
                        author.orcid = "0000-0001-2345-6789")

cat("By ID:", nrow(oa_by_id), "works\n")
cat("By ORCID:", nrow(oa_by_orcid), "works\n")

# Find discrepancies
only_id <- setdiff(oa_by_id$id, oa_by_orcid$id)
only_orcid <- setdiff(oa_by_orcid$id, oa_by_id$id)
cat("Only in ID set:", length(only_id), "\n")
cat("Only in ORCID set:", length(only_orcid), "\n")
```

*Key insight:* Discrepancies arise from incomplete ORCID claiming, name
disambiguation errors, and delayed metadata updates.

**3. Name variant detection.**
Search for a common name and assess disambiguation.

```{r}
#| label: sol-ch12-3
#| eval: false
wang_wei <- oa_fetch(entity = "authors",
                     search = "Wang Wei")

nrow(wang_wei)  # many entities

wang_wei |>
 select(id, display_name, works_count, cited_by_count) |>
 arrange(desc(works_count)) |>
 head(20)

# Heuristics for identifying the correct author:
# - Match institutional affiliation
# - Check ORCID linkage
# - Verify co-author overlap with known collaborators
# - Inspect topic/concept alignment
```

*Key insight:* Common names generate hundreds of author entities; institutional
affiliation and ORCID are the most reliable disambiguation features.

**4. Fractional productivity.**
Recompute productivity using fractional counting.

```{r}
#| label: sol-ch12-4
#| eval: false
author_works <- oa_fetch(entity = "works",
                         author.id = "A5023888391")

fractional <- author_works |>
 mutate(n_authors = map_int(author, nrow),
        frac_credit = 1 / n_authors) |>
 group_by(publication_year) |>
 summarise(full_count = n(),
           frac_count = sum(frac_credit))

ggplot(fractional, aes(publication_year)) +
 geom_line(aes(y = full_count, colour = "Full")) +
 geom_line(aes(y = frac_count, colour = "Fractional")) +
 labs(title = "Full vs. fractional productivity",
      x = "Year", y = "Papers", colour = "Counting")
```

*Key insight:* Fractional counting penalises hyper-authored papers. Authors who
collaborate broadly see a larger gap between full and fractional counts.

**5. Self-citation ratio.**
Estimate the fraction of self-citations for an author.

```{r}
#| label: sol-ch12-5
#| eval: false
target_author <- "A5023888391"

# Sample a subset of the author's works
sample_works <- author_works |> slice_sample(n = 10)

self_cite_frac <- map_dbl(sample_works$id, \(wid) {
 citing <- tryCatch(
   oa_fetch(entity = "works", cites = wid),
   error = \(e) tibble()
 )
 if (nrow(citing) == 0) return(NA_real_)
 citing |>
   unnest(author) |>
   summarise(has_self = any(au_id == target_author)) |>
   pull(has_self) |>
   mean()
})

cat(sprintf("Estimated self-citation rate: %.1f%%\n",
            mean(self_cite_frac, na.rm = TRUE) * 100))
```

*Key insight:* Self-citation rates of 5-15% are typical; rates above 25%
may signal strategic self-citation behaviour.

## Chapter 13 --- Institutional and Country Analysis {-}

**1. Country-level analysis.**
Aggregate by country using full and fractional counts.

```{r}
#| label: sol-ch13-1
#| eval: false
library(tidyverse)

country_counts <- works |>
 unnest(authorships) |>
 unnest(countries) |>
 rename(country = countries) |>
 mutate(n_countries = n_distinct(country), .by = id) |>
 summarise(
   full_count = n_distinct(id),
   frac_count = sum(1 / n_countries),
   .by = country
 ) |>
 arrange(desc(full_count)) |>
 head(10)

country_counts |>
 mutate(ratio = full_count / frac_count) |>
 ggplot(aes(reorder(country, full_count), ratio)) +
 geom_col() +
 coord_flip() +
 labs(title = "Full / fractional count ratio by country",
      x = NULL, y = "Ratio")
```

*Key insight:* Countries with high international collaboration rates (e.g.,
Switzerland, Singapore) show a larger ratio, meaning full counting inflates
their apparent output.

**2. PP(top 10%) by institution.**
Compute the proportion of papers in the top 10% most cited.

```{r}
#| label: sol-ch13-2
#| eval: false
# Compute year-specific 90th percentile thresholds
thresholds <- works |>
 group_by(publication_year) |>
 summarise(p90 = quantile(cited_by_count, 0.9))

inst_top10 <- works |>
 unnest(authorships) |>
 unnest(institutions) |>
 left_join(thresholds, by = "publication_year") |>
 mutate(is_top10 = cited_by_count >= p90) |>
 summarise(
   pp_top10 = mean(is_top10),
   mncs = mean(cited_by_count),
   .by = display_name
 ) |>
 filter(!is.na(display_name)) |>
 arrange(desc(pp_top10))

ggplot(inst_top10 |> head(20), aes(mncs, pp_top10)) +
 geom_point() +
 ggrepel::geom_text_repel(aes(label = display_name), size = 3) +
 labs(x = "MNCS", y = "PP(top 10%)")
```

*Key insight:* PP(top 10%) and MNCS generally correlate but can diverge if an
institution has a few extremely highly cited outliers.

**3. Collaboration intensity.**
Relate the number of institutional affiliations per paper to counting method
effects.

```{r}
#| label: sol-ch13-3
#| eval: false
collab_intensity <- works |>
 unnest(authorships) |>
 unnest(institutions) |>
 summarise(
   n_insts = n_distinct(display_name),
   .by = id
 )

inst_collab <- works |>
 unnest(authorships) |>
 unnest(institutions) |>
 left_join(collab_intensity, by = "id") |>
 summarise(
   mean_collab = mean(n_insts),
   full_count = n_distinct(id),
   frac_count = sum(1 / n_insts),
   ratio = full_count / frac_count,
   .by = display_name
 ) |>
 filter(!is.na(display_name))

ggplot(inst_collab, aes(mean_collab, ratio)) +
 geom_point(alpha = 0.5) +
 geom_smooth(method = "lm") +
 labs(x = "Mean affiliations per paper",
      y = "Full / fractional ratio",
      title = "Collaboration intensity vs. counting effect")
```

*Key insight:* There is a positive relationship --- institutions with higher
collaboration intensity are more inflated by full counting.

## Chapter 14 --- Temporal Analysis and Citation Aging {-}

**1. Half-life comparison.**
Compare citation half-lives between a mathematics and a biomedical journal.

```{r}
#| label: sol-ch14-1
#| eval: false
library(openalexR)
library(tidyverse)

fetch_half_life <- function(journal_id, label) {
 w <- oa_fetch(entity = "works",
               primary_location.source.id = journal_id,
               from_publication_date = "2015-01-01",
               to_publication_date = "2018-12-31")
 w |>
   mutate(age = 2024 - publication_year) |>
   group_by(age) |>
   summarise(mean_cites = mean(cited_by_count)) |>
   mutate(cum_share = cumsum(mean_cites) / sum(mean_cites),
          journal = label)
}

math_hl   <- fetch_half_life("S154589771", "Mathematics")
biomed_hl <- fetch_half_life("S49861241", "Biomedicine")

bind_rows(math_hl, biomed_hl) |>
 ggplot(aes(age, cum_share, colour = journal)) +
 geom_line() +
 geom_hline(yintercept = 0.5, linetype = "dashed") +
 labs(title = "Citation half-life comparison",
      x = "Age (years)", y = "Cumulative citation share")
```

*Key insight:* Biomedical journals typically have shorter half-lives (3-5 years)
than mathematics journals (8-12 years) because of faster publication cycles.

**2. Price index computation.**
Compute the fraction of references less than 5 years old.

```{r}
#| label: sol-ch14-2
#| eval: false
# For papers with reference data available
sample_papers <- works |>
 filter(publication_year == 2020) |>
 slice_sample(n = 20)

price_indices <- map_dbl(sample_papers$id, \(wid) {
 refs <- tryCatch(
   oa_fetch(entity = "works", cited_by = wid),
   error = \(e) tibble()
 )
 if (nrow(refs) == 0) return(NA_real_)
 pub_year <- sample_papers$publication_year[sample_papers$id == wid]
 mean(refs$publication_year >= (pub_year - 5), na.rm = TRUE)
})

cat(sprintf("Mean Price index: %.2f\n", mean(price_indices, na.rm = TRUE)))
```

*Key insight:* A high Price index indicates the field relies on recent work
(fast-moving); a low Price index suggests reliance on older, established
literature.

**3. Sleeping beauty detection.**
Identify delayed-recognition papers.

```{r}
#| label: sol-ch14-3
#| eval: false
sleeping <- works |>
 filter(publication_year <= 2018,
        cited_by_count >= 20) |>
 mutate(age = 2024 - publication_year)

# Approximate: papers with low early citation rate but high total
# (exact per-year citation data requires counts_by_year from OpenAlex)
sleeping_candidates <- sleeping |>
 filter(
   map_lgl(counts_by_year, \(cy) {
     if (is.null(cy)) return(FALSE)
     early <- cy |> filter(year <= publication_year + 3)
     sum(early$cited_by_count, na.rm = TRUE) < 2
   })
 )

sleeping_candidates |>
 select(display_name, publication_year, cited_by_count) |>
 arrange(desc(cited_by_count))
```

*Key insight:* Sleeping beauties are rare but occur across all fields; they
often represent ahead-of-their-time ideas that gained traction only after a
paradigm shift.

**4. Median vs. mean aging.**
Replot aging curves using median citations.

```{r}
#| label: sol-ch14-4
#| eval: false
aging_comparison <- works |>
 mutate(age = 2024 - publication_year) |>
 group_by(age) |>
 summarise(
   mean_cites   = mean(cited_by_count),
   median_cites = median(cited_by_count)
 ) |>
 pivot_longer(c(mean_cites, median_cites),
              names_to = "stat", values_to = "value")

ggplot(aging_comparison, aes(age, value, colour = stat)) +
 geom_line() +
 geom_point() +
 labs(title = "Mean vs. median citation aging",
      x = "Age (years)", y = "Citations")
```

*Key insight:* The median curve is lower and flatter because the mean is pulled
up by highly cited outliers. Median is more robust for skewed distributions.

**5. Citation window sensitivity.**
Compute JIF-like proxies with different window lengths.

```{r}
#| label: sol-ch14-5
#| eval: false
windows <- c(2, 3, 5)

jif_by_window <- map(windows, \(w) {
 journal_works |>
   filter(publication_year >= (2023 - w),
          publication_year <= 2022) |>
   summarise(jif = sum(cited_by_count) / n()) |>
   mutate(window = w)
}) |> list_rbind()

ggplot(jif_by_window, aes(factor(window), jif)) +
 geom_col() +
 labs(x = "Citation window (years)", y = "JIF proxy",
      title = "JIF sensitivity to citation window")
```

*Key insight:* Longer windows generally produce higher values and favour fields
with slower citation accumulation.

## Chapter 15 --- Co-authorship Networks {-}

**1. Fractional counting.**
Weight edges by 1/(k-1) where k is the number of authors.

```{r}
#| label: sol-ch15-1
#| eval: false
library(igraph)
library(tidyverse)

edges_frac <- works |>
 unnest(author) |>
 group_by(id) |>
 mutate(k = n()) |>
 filter(k >= 2) |>
 reframe(
   pair = combn(au_id, 2, simplify = FALSE),
   weight = 1 / (k - 1)
 ) |>
 unnest_wider(pair, names_sep = "_") |>
 rename(from = pair_1, to = pair_2)

g_frac <- graph_from_data_frame(edges_frac, directed = FALSE)
E(g_frac)$weight <- edges_frac$weight

degree_frac <- strength(g_frac) |>
 sort(decreasing = TRUE) |>
 head(10)
degree_frac
```

*Key insight:* Fractional counting reduces the influence of mega-authored
papers. Authors primarily involved in large teams may drop in rank.

**2. Temporal slicing.**
Build yearly networks and track structural evolution.

```{r}
#| label: sol-ch15-2
#| eval: false
yearly_stats <- works |>
 group_by(publication_year) |>
 group_map(\(df, key) {
   edges <- df |>
     unnest(author) |>
     group_by(id) |>
     filter(n() >= 2) |>
     reframe(pair = combn(au_id, 2, simplify = FALSE)) |>
     unnest_wider(pair, names_sep = "_")
   if (nrow(edges) == 0) return(NULL)
   g <- graph_from_data_frame(edges, directed = FALSE)
   tibble(
     year = key$publication_year,
     n_nodes = vcount(g),
     n_components = components(g)$no,
     mean_degree = mean(degree(g))
   )
 }) |> list_rbind()

ggplot(yearly_stats, aes(year, mean_degree)) +
 geom_line() +
 geom_point() +
 labs(title = "Mean degree over time", x = "Year", y = "Mean degree")
```

*Key insight:* Mean degree typically increases over time as collaboration
becomes more common; the number of components may decrease as the field becomes
more connected.

**3. Betweenness vs. degree.**
Compare centrality measures for the giant component.

```{r}
#| label: sol-ch15-3
#| eval: false
g <- build_coauth_graph(works)
gc <- induced_subgraph(g, components(g)$membership ==
                         which.max(components(g)$csize))

centrality <- tibble(
 author = V(gc)$name,
 deg = degree(gc),
 btw = betweenness(gc, normalized = TRUE)
)

ggplot(centrality, aes(deg, btw)) +
 geom_point(alpha = 0.4) +
 labs(x = "Degree", y = "Betweenness (normalised)",
      title = "Betweenness vs. degree in co-authorship network")
```

*Key insight:* High-betweenness, low-degree nodes are "bridge" authors who
connect otherwise separate groups --- they may be interdisciplinary researchers.

**4. Resolution parameter.**
Vary the Leiden resolution and compare community counts.

```{r}
#| label: sol-ch15-4
#| eval: false
library(igraph)

resolutions <- c(0.5, 1.0, 2.0)

res_results <- map(resolutions, \(r) {
 comm <- cluster_leiden(gc, resolution = r,
                        objective_function = "modularity")
 tibble(resolution = r,
        n_communities = length(comm),
        modularity = modularity(comm))
})

list_rbind(res_results)
```

*Key insight:* Higher resolution produces more communities. The optimal value
depends on the application --- lower resolution reveals broad thematic clusters,
higher resolution reveals tight collaborative teams.

## Chapter 16 --- Co-citation and Bibliographic Coupling {-}

**1. Cosine normalisation.**
Normalise co-citation strengths by the geometric mean of individual citation
counts.

```{r}
#| label: sol-ch16-1
#| eval: false
library(igraph)
library(tidyverse)

# Assume cocit_matrix is the raw co-citation matrix
diag_sqrt <- sqrt(diag(cocit_matrix))
cosine_matrix <- cocit_matrix / outer(diag_sqrt, diag_sqrt)
diag(cosine_matrix) <- 0

g_cosine <- graph_from_adjacency_matrix(cosine_matrix,
                                         mode = "undirected",
                                         weighted = TRUE)

# Compare top pairs
top_pairs_raw <- E(g_raw)[order(-E(g_raw)$weight)][1:10]
top_pairs_cos <- E(g_cosine)[order(-E(g_cosine)$weight)][1:10]
```

*Key insight:* Cosine normalisation down-weights pairs that are individually
highly cited, surfacing topically related pairs that raw counts miss.

**2. Temporal co-citation.**
Split the corpus into two periods and build separate networks.

```{r}
#| label: sol-ch16-2
#| eval: false
periods <- list(
 early = works |> filter(publication_year %in% 2020:2021),
 late  = works |> filter(publication_year %in% 2022:2023)
)

period_graphs <- map(periods, \(df) {
 # Build co-citation network for this period's references
 refs <- df |> unnest(referenced_works)
 ref_pairs <- refs |>
   group_by(id) |>
   filter(n() >= 2) |>
   reframe(pair = combn(referenced_works, 2, simplify = FALSE)) |>
   unnest_wider(pair, names_sep = "_")

 ref_pairs |>
   count(pair_1, pair_2, name = "weight") |>
   graph_from_data_frame(directed = FALSE)
})

# Find stable pairs: edges present in both periods
shared_edges <- E(period_graphs$early) %in% E(period_graphs$late)
```

*Key insight:* Stable co-citation pairs represent the intellectual core of the
field; new pairs emerging in the later period may signal research-front shifts.

**3. Bibliographic coupling by year.**
Track modularity and fragmentation of the intellectual structure.

```{r}
#| label: sol-ch16-3
#| eval: false
yearly_bc <- works |>
 group_by(publication_year) |>
 group_map(\(df, key) {
   refs <- df |> unnest(referenced_works)
   pairs <- refs |>
     group_by(referenced_works) |>
     filter(n() >= 2) |>
     reframe(pair = combn(id, 2, simplify = FALSE)) |>
     unnest_wider(pair, names_sep = "_") |>
     count(pair_1, pair_2, name = "weight")
   if (nrow(pairs) == 0) return(NULL)
   g <- graph_from_data_frame(pairs, directed = FALSE)
   comm <- cluster_leiden(g, resolution = 1.0)
   tibble(year = key$publication_year,
          n_components = components(g)$no,
          modularity = modularity(comm))
 }) |> list_rbind()

ggplot(yearly_bc, aes(year, modularity)) +
 geom_line() +
 labs(title = "Bibliographic coupling modularity over time")
```

*Key insight:* Decreasing modularity suggests convergence across subfields;
increasing modularity suggests specialisation and fragmentation.

**4. Identifying research fronts.**
Label communities using top title words.

```{r}
#| label: sol-ch16-4
#| eval: false
library(tidytext)

bc_comm <- cluster_leiden(g_bc, resolution = 1.0)
V(g_bc)$community <- membership(bc_comm)

front_labels <- works |>
 filter(id %in% V(g_bc)$name) |>
 left_join(
   tibble(id = V(g_bc)$name, community = V(g_bc)$community),
   by = "id"
 ) |>
 unnest_tokens(word, display_name) |>
 anti_join(stop_words, by = "word") |>
 count(community, word, sort = TRUE) |>
 group_by(community) |>
 slice_head(n = 3) |>
 summarise(label = paste(word, collapse = ", "))

front_labels
```

*Key insight:* Community labels based on frequent title words typically
correspond to recognisable subfields or research themes.

## Chapter 17 --- Co-word and Keyword Co-occurrence {-}

**1. Author keywords vs. concepts.**
Build and compare two co-occurrence networks.

```{r}
#| label: sol-ch17-1
#| eval: false
library(igraph)
library(tidyverse)

# Author keywords network
kw_pairs <- works |>
 unnest(keywords) |>
 group_by(id) |>
 filter(n() >= 2) |>
 reframe(pair = combn(keyword, 2, simplify = FALSE)) |>
 unnest_wider(pair, names_sep = "_") |>
 count(pair_1, pair_2, name = "weight", sort = TRUE)

g_kw <- graph_from_data_frame(kw_pairs, directed = FALSE)

# OpenAlex concepts network
concept_pairs <- works |>
 unnest(concepts) |>
 group_by(id) |>
 filter(n() >= 2) |>
 reframe(pair = combn(display_name, 2, simplify = FALSE)) |>
 unnest_wider(pair, names_sep = "_") |>
 count(pair_1, pair_2, name = "weight", sort = TRUE)

g_concept <- graph_from_data_frame(concept_pairs, directed = FALSE)

cat("Author KW network:", vcount(g_kw), "nodes,",
    ecount(g_kw), "edges\n")
cat("Concept network:", vcount(g_concept), "nodes,",
    ecount(g_concept), "edges\n")
```

*Key insight:* Author keywords are more specific but inconsistent; OpenAlex
concepts are more standardised but coarser.

**2. Temporal evolution.**
Track the top keywords by degree over time.

```{r}
#| label: sol-ch17-2
#| eval: false
yearly_top_kw <- works |>
 unnest(keywords) |>
 group_by(publication_year) |>
 group_map(\(df, key) {
   pairs <- df |>
     group_by(id) |>
     filter(n() >= 2) |>
     reframe(pair = combn(keyword, 2, simplify = FALSE)) |>
     unnest_wider(pair, names_sep = "_")
   if (nrow(pairs) == 0) return(NULL)
   g <- graph_from_data_frame(pairs, directed = FALSE)
   tibble(
     year = key$publication_year,
     keyword = names(sort(degree(g), decreasing = TRUE)[1:5]),
     degree = sort(degree(g), decreasing = TRUE)[1:5]
   )
 }) |> list_rbind()

ggplot(yearly_top_kw, aes(year, degree, colour = keyword)) +
 geom_line() +
 labs(title = "Top keywords by degree over time")
```

*Key insight:* Some keywords are perennially central; others emerge or fade,
signalling evolving research interests.

**3. Strategic diagram.**
Plot clusters on a density-centrality plane.

```{r}
#| label: sol-ch17-3
#| eval: false
comm <- cluster_leiden(g_kw, resolution = 1.0)
V(g_kw)$community <- membership(comm)

strategic <- map(unique(membership(comm)), \(c) {
 nodes <- V(g_kw)[V(g_kw)$community == c]
 subg <- induced_subgraph(g_kw, nodes)
 tibble(
   community = c,
   density = edge_density(subg),
   centrality = mean(strength(g_kw, v = nodes))
 )
}) |> list_rbind()

ggplot(strategic, aes(centrality, density,
                      label = community)) +
 geom_point(size = 3) +
 geom_text(nudge_y = 0.01) +
 geom_hline(yintercept = median(strategic$density),
            linetype = "dashed") +
 geom_vline(xintercept = median(strategic$centrality),
            linetype = "dashed") +
 labs(title = "Strategic diagram",
      x = "Centrality (external)", y = "Density (internal)")
```

*Key insight:* Upper-right quadrant clusters are core themes (high density and
centrality); lower-left clusters are peripheral or emerging topics.

**4. Normalisation comparison.**
Compare raw and association-strength-normalised networks.

```{r}
#| label: sol-ch17-4
#| eval: false
# Association strength: w_ij / (w_i * w_j / 2m)
total_weight <- sum(E(g_kw)$weight)
node_strength <- strength(g_kw)

assoc_edges <- as_data_frame(g_kw) |>
 mutate(
   assoc = weight / (node_strength[from] * node_strength[to] /
                     (2 * total_weight))
 )

g_assoc <- graph_from_data_frame(assoc_edges, directed = FALSE)
E(g_assoc)$weight <- assoc_edges$assoc

# Compare top-centrality keywords
top_raw <- names(sort(strength(g_kw), decreasing = TRUE))[1:10]
top_assoc <- names(sort(strength(g_assoc), decreasing = TRUE))[1:10]

tibble(raw = top_raw, normalised = top_assoc)
```

*Key insight:* Normalisation reduces the dominance of high-frequency generic
terms, elevating more specific and meaningful keyword pairs.

## Chapter 18 --- Community Detection and Backbone Extraction {-}

**1. Alpha sensitivity.**
Run the disparity filter across multiple alpha values.

```{r}
#| label: sol-ch18-1
#| eval: false
library(igraph)
library(tidyverse)

alphas <- c(0.01, 0.05, 0.10, 0.20)

alpha_results <- map(alphas, \(a) {
 # Disparity filter: keep edge if p_ij < alpha
 el <- as_data_frame(g) |>
   mutate(
     s_from = strength(g)[from],
     k_from = degree(g)[from],
     p_ij = (1 - weight / s_from)^(k_from - 1)
   ) |>
   filter(p_ij < a)

 g_bb <- graph_from_data_frame(el, directed = FALSE)
 comm <- cluster_leiden(g_bb, resolution = 1.0)
 tibble(alpha = a,
        edges = ecount(g_bb),
        communities = length(comm),
        connected = components(g_bb)$no == 1)
}) |> list_rbind()

alpha_results
```

*Key insight:* Very low alpha values remove too many edges and disconnect the
network; very high values retain noise. Typical values of 0.05-0.10 balance
signal and connectivity.

**2. Threshold vs. disparity.**
Compare two backbone extraction methods using NMI.

```{r}
#| label: sol-ch18-2
#| eval: false
library(igraph)

# Threshold filtering: keep edges above median weight
threshold <- median(E(g)$weight)
g_thresh <- subgraph.edges(g, which(E(g)$weight > threshold))
comm_thresh <- cluster_leiden(g_thresh, resolution = 1.0)

# Disparity filtering (from exercise 1)
comm_disp <- cluster_leiden(g_bb, resolution = 1.0)

# NMI comparison on shared nodes
shared <- intersect(V(g_thresh)$name, V(g_bb)$name)
m1 <- membership(comm_thresh)[shared]
m2 <- membership(comm_disp)[shared]

compare(m1, m2, method = "nmi")
```

*Key insight:* The disparity filter preserves locally significant edges
(important for low-degree nodes) while the threshold filter favours globally
strong connections. NMI values below 0.7 indicate meaningfully different
community structures.

**3. Hierarchical structure.**
Check whether communities at different resolutions are nested.

```{r}
#| label: sol-ch18-3
#| eval: false
resolutions <- c(0.5, 1.0, 2.0)

memberships <- map(resolutions, \(r) {
 comm <- cluster_leiden(g_bb, resolution = r)
 tibble(node = V(g_bb)$name, community = membership(comm),
        resolution = r)
}) |> list_rbind()

# Check nesting: at resolution 2.0, are communities subsets of
# resolution 0.5 communities?
nesting <- memberships |>
 pivot_wider(names_from = resolution, values_from = community,
             names_prefix = "res_") |>
 count(res_0.5, res_2) |>
 group_by(res_2) |>
 mutate(is_nested = n_distinct(res_0.5) == 1)

mean(nesting$is_nested)  # proportion of fine communities nested
```

*Key insight:* Perfectly nested hierarchies are rare in real networks.
Partial nesting suggests a hierarchical thematic structure with some cross-
cutting topics.

**4. Weighted vs. unweighted community detection.**
Remove weights and compare results.

```{r}
#| label: sol-ch18-4
#| eval: false
g_unweighted <- g_bb
E(g_unweighted)$weight <- 1

comm_weighted   <- cluster_leiden(g_bb, resolution = 1.0)
comm_unweighted <- cluster_leiden(g_unweighted, resolution = 1.0)

compare(membership(comm_weighted), membership(comm_unweighted),
        method = "nmi")

cat("Weighted communities:", length(comm_weighted), "\n")
cat("Unweighted communities:", length(comm_unweighted), "\n")
```

*Key insight:* Removing weights often increases the number of communities because
weak connections that helped merge groups are no longer distinguishable from
strong ones.

## Chapter 19 --- Science Mapping and Overlay Maps {-}

**1. Multi-institution overlay.**
Overlay two institutions on the same base map.

```{r}
#| label: sol-ch19-1
#| eval: false
library(tidyverse)
library(ggrepel)

inst_ids <- c("I123456789", "I987654321")

profiles <- map(inst_ids, \(iid) {
 w <- oa_fetch(entity = "works",
               authorships.institutions.lineage = iid,
               from_publication_date = "2020-01-01")
 w |>
   unnest(concepts) |>
   count(display_name, name = "n_works") |>
   mutate(inst = iid)
}) |> list_rbind()

ggplot(profiles, aes(display_name, n_works, fill = inst)) +
 geom_col(position = "dodge") +
 coord_flip() +
 labs(title = "Research profile overlay")
```

*Key insight:* Overlapping areas indicate shared research strengths; divergent
areas reveal institutional specialisation.

**2. Temporal overlay.**
Compare the same institution's profile across two time periods.

```{r}
#| label: sol-ch19-2
#| eval: false
periods <- list(
 early = c("2018-01-01", "2019-12-31"),
 late  = c("2022-01-01", "2023-12-31")
)

temporal_profiles <- imap(periods, \(dates, label) {
 oa_fetch(entity = "works",
          authorships.institutions.lineage = "I123456789",
          from_publication_date = dates[1],
          to_publication_date   = dates[2]) |>
   unnest(concepts) |>
   count(display_name, name = "n_works") |>
   mutate(period = label)
}) |> list_rbind()

temporal_profiles |>
 pivot_wider(names_from = period, values_from = n_works,
             values_fill = 0) |>
 mutate(shift = late - early) |>
 arrange(desc(abs(shift))) |>
 head(15)
```

*Key insight:* Positive shifts indicate growing research areas; negative shifts
suggest declining focus or strategic reorientation.

**3. Country-level overlay.**
Compare a small and a large country's publication profiles.

```{r}
#| label: sol-ch19-3
#| eval: false
countries <- c("SG", "US")  # Singapore vs. United States

country_profiles <- map(countries, \(cc) {
 oa_fetch(entity = "works",
          authorships.countries = cc,
          from_publication_date = "2022-01-01",
          to_publication_date = "2022-12-31") |>
   slice_sample(n = 500) |>
   unnest(concepts) |>
   count(display_name, name = "n_works") |>
   mutate(share = n_works / sum(n_works),
          country = cc)
}) |> list_rbind()

country_profiles |>
 pivot_wider(names_from = country, values_from = share,
             values_fill = 0) |>
 ggplot(aes(US, SG)) +
 geom_point() +
 geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
 ggrepel::geom_text_repel(aes(label = display_name), size = 2) +
 labs(title = "Country profile: Singapore vs. US")
```

*Key insight:* Small countries tend to show concentrated profiles (above the
diagonal in selected areas); large countries are more evenly distributed.

## Chapter 20 --- Network Visualization and Interoperability {-}

**1. Layout comparison.**
Apply five layouts and compare quality and speed.

```{r}
#| label: sol-ch20-1
#| eval: false
library(igraph)
library(ggraph)
library(tidyverse)

layouts <- c("fr", "kk", "drl", "graphopt", "lgl")

layout_results <- map(layouts, \(algo) {
 t <- system.time({
   p <- ggraph(gc, layout = algo) +
     geom_edge_link(alpha = 0.1) +
     geom_node_point(aes(colour = factor(community)), size = 1) +
     labs(title = algo) +
     theme_void()
 })
 list(plot = p, time = t["elapsed"], layout = algo)
})

# Print timing
map(layout_results, \(x) tibble(layout = x$layout, secs = x$time)) |>
 list_rbind()
```

*Key insight:* Fruchterman-Reingold is a good default; Kamada-Kawai is better
for small networks; DRL scales to large graphs but can produce chaotic layouts.

**2. Interactive filtering.**
Create a filterable `visNetwork` visualisation.

```{r}
#| label: sol-ch20-2
#| eval: false
library(visNetwork)

nodes_df <- tibble(
 id = V(gc)$name,
 label = V(gc)$name,
 group = as.character(V(gc)$community)
)

edges_df <- as_data_frame(gc) |>
 select(from, to)

visNetwork(nodes_df, edges_df) |>
 visOptions(selectedBy = "group",
            highlightNearest = TRUE) |>
 visPhysics(stabilization = FALSE)
```

*Key insight:* Interactive filtering lets users focus on specific communities
and explore inter-community bridges.

**3. Gephi round-trip.**
Export to GraphML for use in Gephi.

```{r}
#| label: sol-ch20-3
#| eval: false
library(igraph)

# Export
write_graph(gc, "network.graphml", format = "graphml")

# After Gephi processing, compare:
# - Force Atlas 2 layout often reveals community separation better
# - R/ggraph offers more reproducible, scriptable figures
# - Gephi is better for interactive exploration and publication
#   quality figures via SVG/PDF export
```

*Key insight:* Gephi's Force Atlas 2 often produces cleaner community
separation; R workflows are more reproducible and automatable.

**4. Colour-blind check.**
Test colour distinguishability with increasing community counts.

```{r}
#| label: sol-ch20-4
#| eval: false
library(ggraph)
library(viridis)

n_communities <- c(3, 5, 10)

plots <- map(n_communities, \(k) {
 comm <- cluster_leiden(gc, resolution = k / 5)
 V(gc)$comm_k <- membership(comm)

 ggraph(gc, layout = "fr") +
   geom_edge_link(alpha = 0.05) +
   geom_node_point(aes(colour = factor(comm_k)), size = 2) +
   scale_colour_viridis_d(option = "D") +
   labs(title = paste(k, "communities")) +
   theme_void() +
   theme(legend.position = "none")
})

# Use colorBlindness package or online simulator to test
# library(colorBlindness)
# cvdPlot(plots[[3]])
```

*Key insight:* Viridis palettes remain distinguishable up to about 8 categories;
beyond that, consider shape or faceting as supplementary channels.

## Chapter 21 --- Text Mining Bibliographic Corpora {-}

**1. Bigram analysis.**
Build a DFM from bigrams.

```{r}
#| label: sol-ch21-1
#| eval: false
library(quanteda)
library(tidyverse)

corp <- corpus(works, text_field = "ab")
toks <- tokens(corp, remove_punct = TRUE, remove_numbers = TRUE) |>
 tokens_tolower() |>
 tokens_remove(stopwords("en")) |>
 tokens_ngrams(n = 2)

dfm_bi <- dfm(toks) |> dfm_trim(min_termfreq = 5)

topfeatures(dfm_bi, 20)
```

*Key insight:* Bigrams capture meaningful phrases like "machine_learning" or
"citation_analysis" that single words miss.

**2. Lexical diversity.**
Compute the type-token ratio (TTR) per year.

```{r}
#| label: sol-ch21-2
#| eval: false
library(quanteda)
library(quanteda.textstats)

ttr_by_year <- works |>
 filter(!is.na(ab)) |>
 group_by(publication_year) |>
 summarise(
   all_text = paste(ab, collapse = " "),
   .groups = "drop"
 ) |>
 mutate(
   n_types  = map_int(all_text, \(t) {
     toks <- tokens(t, remove_punct = TRUE) |> tokens_tolower()
     ntype(dfm(toks))
   }),
   n_tokens = map_int(all_text, \(t) {
     toks <- tokens(t, remove_punct = TRUE)
     ntoken(toks)
   }),
   ttr = n_types / n_tokens
 )

ggplot(ttr_by_year, aes(publication_year, ttr)) +
 geom_line() +
 labs(title = "Lexical diversity over time",
      x = "Year", y = "Type-token ratio")
```

*Key insight:* Declining TTR may indicate vocabulary standardisation; increasing
TTR suggests diversifying terminology.

**3. Keyness analysis.**
Identify terms that distinguish a specific year.

```{r}
#| label: sol-ch21-3
#| eval: false
library(quanteda)
library(quanteda.textstats)

corp <- corpus(works |> filter(!is.na(ab)),
               text_field = "ab",
               docid_field = "id")
docvars(corp, "year") <- works$publication_year[!is.na(works$ab)]

dfm_all <- dfm(tokens(corp, remove_punct = TRUE) |>
               tokens_tolower() |>
               tokens_remove(stopwords("en")))

# Compare 2023 vs. all other years
dfm_grouped <- dfm_group(dfm_all, groups = ifelse(
 docvars(dfm_all, "year") == 2023, "2023", "other"
))

keyness <- textstat_keyness(dfm_grouped, target = "2023")
textplot_keyness(keyness, n = 20)
```

*Key insight:* Keyness analysis surfaces terms unique to a period. For 2023,
terms related to LLMs and AI may dominate.

**4. Preprocessing sensitivity.**
Compare analyses with and without stemming.

```{r}
#| label: sol-ch21-4
#| eval: false
toks_nostem <- tokens(corp, remove_punct = TRUE) |>
 tokens_tolower() |>
 tokens_remove(stopwords("en"))

toks_stem <- tokens_wordstem(toks_nostem)

top_nostem <- topfeatures(dfm(toks_nostem), 20)
top_stem   <- topfeatures(dfm(toks_stem), 20)

tibble(
 rank = 1:20,
 unstemmed = names(top_nostem),
 stemmed = names(top_stem)
)
```

*Key insight:* Stemming merges variants (e.g., "citations" and "citation") but
can produce unintelligible stems. For keyword-based analyses, unstemmed terms
are often more interpretable.

## Chapter 22 --- Topic Modeling {-}

**1. K selection.**
Fit LDA models with multiple values of K and compare coherence.

```{r}
#| label: sol-ch22-1
#| eval: false
library(topicmodels)
library(tidyverse)
set.seed(20260509)

ks <- c(5, 8, 12, 15)

models <- map(ks, \(k) {
 LDA(dtm, k = k, method = "Gibbs",
     control = list(seed = 20260509, iter = 1000))
})

# Compute coherence (using top-10 terms per topic)
coherence_scores <- map_dbl(models, \(mod) {
 terms <- terms(mod, 10)
 # Simple PMI-based coherence approximation
 mean(map_dbl(seq_len(ncol(terms)), \(t) {
   top_terms <- terms[, t]
   pairs <- combn(top_terms, 2, simplify = FALSE)
   mean(map_dbl(pairs, \(p) {
     log((slam::col_sums(dtm[, p[1]] > 0 & dtm[, p[2]] > 0) + 1) /
         slam::col_sums(dtm[, p[1]] > 0))
   }))
 }))
})

tibble(k = ks, coherence = coherence_scores) |>
 ggplot(aes(k, coherence)) +
 geom_line() +
 geom_point() +
 labs(title = "Topic coherence by K", x = "K", y = "Coherence")
```

*Key insight:* The best K balances coherence (higher is better) and
interpretability. There is rarely a single "correct" K.

**2. STM with journal covariate.**
Fit a Structural Topic Model with journal as a prevalence covariate.

```{r}
#| label: sol-ch22-2
#| eval: false
library(stm)
set.seed(20260509)

# Prepare data from two journals
two_journals <- works |>
 filter(so %in% c("Scientometrics", "PLOS ONE"), !is.na(ab))

processed <- textProcessor(two_journals$ab,
                           metadata = two_journals)
prepped <- prepDocuments(processed$documents,
                         processed$vocab,
                         processed$meta)

stm_fit <- stm(prepped$documents, prepped$vocab,
               K = 10,
               prevalence = ~ so,
               data = prepped$meta,
               seed = 20260509)

effect <- estimateEffect(1:10 ~ so, stm_fit, metadata = prepped$meta)
plot(effect, covariate = "so", topics = 1:10)
```

*Key insight:* Topics with large journal effects represent disciplinary
specialisation; shared topics indicate cross-cutting themes.

**3. Topic labelling.**
Read top documents per topic to assign labels.

```{r}
#| label: sol-ch22-3
#| eval: false
library(topicmodels)
library(tidyverse)

best_model <- models[[which.max(coherence_scores)]]
gamma <- posterior(best_model)$topics

# For each topic, find the top-5 documents
topic_labels <- map(seq_len(ncol(gamma)), \(t) {
 top_docs <- order(gamma[, t], decreasing = TRUE)[1:5]
 titles <- works$display_name[top_docs]
 tibble(topic = t,
        top_terms = paste(terms(best_model, 5)[, t], collapse = ", "),
        sample_titles = paste(titles, collapse = " | "))
}) |> list_rbind()

topic_labels
# Now assign human-readable labels based on inspection
```

*Key insight:* Automated topic labels from top terms are a starting point;
reading representative documents is essential for meaningful labels.

**4. LDA vs. STM comparison.**
Compare topics from both models.

```{r}
#| label: sol-ch22-4
#| eval: false
# Extract top terms for each model
lda_terms <- terms(best_model, 10)
stm_terms <- labelTopics(stm_fit, n = 10)$prob

# Compute Jaccard similarity between matched topics
jaccard <- function(a, b) {
 length(intersect(a, b)) / length(union(a, b))
}

# Best-match each LDA topic to an STM topic
similarities <- map_dbl(seq_len(ncol(lda_terms)), \(i) {
 max(map_dbl(seq_len(nrow(stm_terms)), \(j) {
   jaccard(lda_terms[, i], stm_terms[j, ])
 }))
})

mean(similarities)
```

*Key insight:* STM topics may differ because the year covariate allows topic
prevalence to shift over time, producing more temporally coherent topics.

## Chapter 23 --- Keyword Extraction {-}

**1. Precision and recall.**
Evaluate RAKE keywords against author keywords.

```{r}
#| label: sol-ch23-1
#| eval: false
library(udpipe)
library(tidyverse)

sample_papers <- works |>
 filter(!is.na(ab), map_lgl(keywords, \(k) length(k) > 0)) |>
 slice_sample(n = 20)

eval_results <- map(seq_len(nrow(sample_papers)), \(i) {
 paper <- sample_papers[i, ]
 rake_kw <- keywords_rake(
   data.frame(doc_id = paper$id,
              sentence = paper$ab,
              stringsAsFactors = FALSE),
   term = "sentence", group = "doc_id"
 ) |> head(10) |> pull(keyword) |> tolower()

 author_kw <- tolower(unlist(paper$keywords))

 tibble(
   precision = mean(rake_kw %in% author_kw),
   recall = mean(author_kw %in% rake_kw)
 )
}) |> list_rbind()

eval_results |>
 summarise(across(everything(), list(mean = mean, sd = sd)))
```

*Key insight:* RAKE precision is typically low (0.1-0.3) because algorithms
and authors conceptualise keywords differently. Recall is also limited because
author keywords often include broad terms not in the abstract.

**2. Bigram keywords.**
Score bigrams by TF-IDF.

```{r}
#| label: sol-ch23-2
#| eval: false
library(tidytext)
library(tidyverse)

bigram_tfidf <- works |>
 filter(!is.na(ab)) |>
 unnest_tokens(bigram, ab, token = "ngrams", n = 2) |>
 separate(bigram, c("word1", "word2"), sep = " ") |>
 filter(!word1 %in% stop_words$word,
        !word2 %in% stop_words$word) |>
 unite(bigram, word1, word2, sep = " ") |>
 count(id, bigram) |>
 bind_tf_idf(bigram, id, n) |>
 group_by(bigram) |>
 summarise(mean_tfidf = mean(tf_idf)) |>
 arrange(desc(mean_tfidf))

bigram_tfidf |> head(20)
```

*Key insight:* Multi-word terms like "citation analysis" and "research
evaluation" are invisible to unigram methods but highly informative.

**3. Keyword enrichment.**
Use extracted keywords to build a co-word network for keyword-poor papers.

```{r}
#| label: sol-ch23-3
#| eval: false
library(igraph)
library(udpipe)

# Papers without author keywords
no_kw_papers <- works |>
 filter(map_lgl(keywords, \(k) length(k) == 0), !is.na(ab))

# Extract keywords via RAKE
extracted <- map(seq_len(nrow(no_kw_papers)), \(i) {
 kw <- keywords_rake(
   data.frame(doc_id = no_kw_papers$id[i],
              sentence = no_kw_papers$ab[i]),
   term = "sentence", group = "doc_id"
 ) |> head(5) |> pull(keyword)
 tibble(id = no_kw_papers$id[i], keyword = kw)
}) |> list_rbind()

# Build co-word network
pairs <- extracted |>
 group_by(id) |>
 filter(n() >= 2) |>
 reframe(pair = combn(keyword, 2, simplify = FALSE)) |>
 unnest_wider(pair, names_sep = "_") |>
 count(pair_1, pair_2, sort = TRUE)

g_enriched <- graph_from_data_frame(pairs, directed = FALSE)
```

*Key insight:* Extracted keywords produce a topical structure comparable to
author keywords, enabling co-word analysis even for papers without manual
keyword assignments.

## Chapter 24 --- Embeddings for Scientometrics {-}

**1. Weighted averaging.**
Use TF-IDF weights for document embedding computation.

```{r}
#| label: sol-ch24-1
#| eval: false
library(word2vec)
library(tidytext)
library(tidyverse)
library(umap)

# Compute TF-IDF weights
tfidf <- works |>
 filter(!is.na(ab)) |>
 unnest_tokens(word, ab) |>
 anti_join(stop_words, by = "word") |>
 count(id, word) |>
 bind_tf_idf(word, id, n)

# Get word vectors
wv <- predict(w2v_model, newdata = unique(tfidf$word), type = "embedding")

# Weighted average per document
doc_embeddings <- tfidf |>
 inner_join(
   as_tibble(wv, rownames = "word"),
   by = "word"
 ) |>
 group_by(id) |>
 summarise(across(starts_with("V"), \(x) weighted.mean(x, tf_idf)))

# UMAP
umap_result <- umap(as.matrix(doc_embeddings |> select(-id)))
```

*Key insight:* TF-IDF weighting down-weights common words, often producing
tighter, more meaningful clusters in the UMAP projection.

**2. K-means clustering.**
Apply k-means to embeddings and compare with LDA topics.

```{r}
#| label: sol-ch24-2
#| eval: false
set.seed(20260509)

embedding_matrix <- as.matrix(doc_embeddings |> select(-id))

km_results <- map(c(5, 8, 12), \(k) {
 km <- kmeans(embedding_matrix, centers = k, nstart = 25)
 tibble(k = k,
        cluster = km$cluster,
        id = doc_embeddings$id)
}) |> list_rbind()

# Compare k=8 clusters with LDA topics (K=8)
# Assuming lda_topics is a vector of topic assignments
km8 <- km_results |> filter(k == 8) |> pull(cluster)
compare_nmi <- igraph::compare(km8, lda_topics, method = "nmi")
cat(sprintf("NMI between k-means and LDA: %.3f\n", compare_nmi))
```

*Key insight:* Moderate NMI (0.3-0.6) is typical; the two methods capture
overlapping but not identical structure because they model different aspects of
the data.

**3. Pre-trained embeddings.**
Use SPECTER embeddings via reticulate.

```{r}
#| label: sol-ch24-3
#| eval: false
library(reticulate)

# Requires Python with transformers installed
transformers <- import("transformers")
torch <- import("torch")

tokenizer <- transformers$AutoTokenizer$from_pretrained(
 "allenai/specter2"
)
model <- transformers$AutoModel$from_pretrained(
 "allenai/specter2"
)

abstracts <- works$ab[!is.na(works$ab)][1:100]
inputs <- tokenizer(abstracts, padding = TRUE, truncation = TRUE,
                    return_tensors = "pt", max_length = 512L)
with(torch$no_grad(), {
 outputs <- model(inputs)
})
specter_emb <- outputs$last_hidden_state$mean(dim = 1L)$numpy()
```

*Key insight:* Pre-trained scientific embeddings like SPECTER capture domain
semantics better than generic word2vec, especially for similarity search.

**4. Analogy test.**
Test analogical reasoning in the word2vec model.

```{r}
#| label: sol-ch24-4
#| eval: false
library(word2vec)

# Test: "journal" - "article" + "book" ≈ ?
analogy <- predict(w2v_model,
                   newdata = c("journal", "article", "book"),
                   type = "nearest", top_n = 5)

# Alternative approach using vector arithmetic
wv <- as.matrix(w2v_model)
if (all(c("journal", "article", "book") %in% rownames(wv))) {
 target <- wv["journal", ] - wv["article", ] + wv["book", ]
 sims <- word2vec_similarity(
   matrix(target, nrow = 1), wv
 )
 sort(sims[1, ], decreasing = TRUE)[1:5]
}
```

*Key insight:* Domain-specific models may capture field-specific analogies (e.g.,
"h-index" - "author" + "journal" near "impact factor") but fail on generic
analogies, reflecting the training corpus.

## Chapter 25 --- Detecting Emerging Topics {-}

**1. Relative burst detection.**
Use proportional frequencies instead of absolute counts.

```{r}
#| label: sol-ch25-1
#| eval: false
library(tidyverse)

# Compute keyword proportions by year
kw_yearly <- works |>
 unnest(keywords) |>
 count(publication_year, keyword) |>
 group_by(publication_year) |>
 mutate(proportion = n / sum(n)) |>
 ungroup()

# Apply Kleinberg burst detection on proportions
bursts_relative <- kw_yearly |>
 group_by(keyword) |>
 filter(n() >= 3) |>  # need multiple years
 summarise(
   burst = list(kleinberg_bursts(proportion, publication_year)),
   .groups = "drop"
 ) |>
 unnest(burst) |>
 arrange(desc(strength))

bursts_relative |> head(20)
```

*Key insight:* Relative frequencies reduce bias toward high-volume years;
some terms that burst in absolute counts may be merely riding overall growth.

**2. Emergence by journal.**
Compare emerging keywords across two journals.

```{r}
#| label: sol-ch25-2
#| eval: false
journals <- c("Scientometrics", "PLOS ONE")

journal_bursts <- map(journals, \(j) {
 kw_j <- works |>
   filter(so == j) |>
   unnest(keywords) |>
   count(publication_year, keyword) |>
   group_by(keyword) |>
   filter(n() >= 3) |>
   summarise(
     burst = list(kleinberg_bursts(n, publication_year))
   ) |>
   unnest(burst) |>
   mutate(journal = j)
}) |> list_rbind()

# Compare top bursting terms
journal_bursts |>
 group_by(journal) |>
 arrange(desc(strength)) |>
 slice_head(n = 10) |>
 select(journal, keyword, strength)
```

*Key insight:* Different journals serve different research communities; emerging
terms may overlap only for broad methodological advances.

**3. Novelty scoring.**
Compute document novelty using cosine distance to prior-year centroids.

```{r}
#| label: sol-ch25-3
#| eval: false
library(quanteda)
library(tidyverse)

dfm_yearly <- dfm(tokens(corpus(works, text_field = "ab"),
                         remove_punct = TRUE) |>
                  tokens_tolower() |>
                  tokens_remove(stopwords("en"))) |>
 dfm_tfidf()

years <- sort(unique(works$publication_year))

novelty <- map(years[-1], \(yr) {
 prior <- dfm_subset(dfm_yearly, publication_year < yr)
 current <- dfm_subset(dfm_yearly, publication_year == yr)
 if (nrow(prior) == 0 || nrow(current) == 0) return(NULL)

 centroid <- colMeans(as.matrix(prior))
 current_mat <- as.matrix(current)

 cos_dist <- 1 - (current_mat %*% centroid) /
   (sqrt(rowSums(current_mat^2)) * sqrt(sum(centroid^2)))

 tibble(year = yr, novelty = as.numeric(cos_dist))
}) |> list_rbind()

novelty |>
 group_by(year) |>
 summarise(mean_novelty = mean(novelty, na.rm = TRUE)) |>
 ggplot(aes(year, mean_novelty)) +
 geom_line() +
 labs(title = "Document novelty over time",
      x = "Year", y = "Mean cosine distance to prior centroid")
```

*Key insight:* Increasing novelty scores suggest the field is exploring new
territory; stable scores indicate incremental progress within established
paradigms.

**4. Validation.**
Manually validate burst-detected emerging terms.

```{r}
#| label: sol-ch25-4
#| eval: false
# Select 5 terms identified as "emerging"
emerging_terms <- bursts_relative |>
 arrange(desc(strength)) |>
 head(5) |>
 pull(keyword)

# For each term, search for corroborating evidence:
# - Is there a review article about this topic?
# - Was it a conference theme in recent years?
# - Does Google Trends show a rising interest?

# Create a validation table
validation <- tibble(
 term = emerging_terms,
 review_found = c(TRUE, TRUE, FALSE, TRUE, FALSE),
 conference_theme = c(TRUE, FALSE, FALSE, TRUE, TRUE),
 google_trend_rising = c(TRUE, TRUE, TRUE, FALSE, TRUE),
 validated = review_found | conference_theme
)

cat(sprintf("False positive rate: %.0f%%\n",
            (1 - mean(validation$validated)) * 100))
```

*Key insight:* Burst detection has moderate precision; triangulating with
qualitative sources reduces false positives and increases confidence.

## Chapter 26 --- Altmetrics {-}

**1. OA citation advantage.**
Compare citations for OA vs. closed-access papers across journals.

```{r}
#| label: sol-ch26-1
#| eval: false
library(openalexR)
library(tidyverse)

journal_ids <- c("S4210174107", "S137773608")

oa_advantage <- map(journal_ids, \(jid) {
 w <- oa_fetch(entity = "works",
               primary_location.source.id = jid,
               from_publication_date = "2019-01-01",
               to_publication_date = "2022-12-31")
 w |>
   group_by(is_oa) |>
   summarise(
     n = n(),
     mean_cites = mean(cited_by_count),
     median_cites = median(cited_by_count)
   ) |>
   mutate(journal = jid)
}) |> list_rbind()

oa_advantage
```

*Key insight:* The OA citation advantage exists in most fields but varies in
magnitude. Self-selection (better papers go OA) may partially explain the
effect.

**2. Temporal attention.**
Track citation accumulation over time for early-attention papers.

```{r}
#| label: sol-ch26-2
#| eval: false
recent_works <- works |>
 filter(publication_year == 2022, !is.null(counts_by_year))

attention_trajectory <- recent_works |>
 unnest(counts_by_year) |>
 group_by(id) |>
 arrange(year) |>
 mutate(cum_cites = cumsum(cited_by_count)) |>
 ungroup()

# Classify: high vs. low early attention (first year citations)
early_cites <- attention_trajectory |>
 filter(year == 2022) |>
 mutate(early_attention = ifelse(cited_by_count >= 5,
                                  "High", "Low"))

attention_trajectory |>
 left_join(early_cites |> select(id, early_attention), by = "id") |>
 group_by(early_attention, year) |>
 summarise(mean_cum = mean(cum_cites, na.rm = TRUE)) |>
 ggplot(aes(year, mean_cum, colour = early_attention)) +
 geom_line() +
 labs(title = "Citation accumulation by early attention")
```

*Key insight:* Papers with high early attention tend to sustain their advantage,
consistent with cumulative advantage (Matthew Effect) in citations.

**3. Cross-field comparison.**
Compare citation distributions and reference list lengths across fields.

```{r}
#| label: sol-ch26-3
#| eval: false
biomed <- oa_fetch(entity = "works",
                   primary_location.source.id = "S49861241",
                   from_publication_date = "2020-01-01") |>
 mutate(field = "Biomedicine") |>
 slice_sample(n = 200)

math <- oa_fetch(entity = "works",
                 primary_location.source.id = "S154589771",
                 from_publication_date = "2020-01-01") |>
 mutate(field = "Mathematics") |>
 slice_sample(n = 200)

comparison <- bind_rows(biomed, math)

comparison |>
 group_by(field) |>
 summarise(
   mean_cites = mean(cited_by_count),
   median_cites = median(cited_by_count),
   mean_refs = mean(referenced_works_count, na.rm = TRUE)
 )

ggplot(comparison, aes(cited_by_count, fill = field)) +
 geom_histogram(position = "dodge", bins = 30) +
 scale_x_log10() +
 labs(title = "Citation distributions by field")
```

*Key insight:* Biomedical papers have higher citation counts and longer reference
lists than mathematics papers. This is why raw citation counts cannot be compared
across fields.

## Chapter 27 --- Open Science Indicators {-}

**1. Cross-journal OA comparison.**
Compare OA rates across three journals in different disciplines.

```{r}
#| label: sol-ch27-1
#| eval: false
library(openalexR)
library(tidyverse)

journals <- list(
 biology = "S49861241",
 physics = "S137773608",
 sociology = "S64468286"
)

oa_rates <- imap(journals, \(jid, field) {
 w <- oa_fetch(entity = "works",
               primary_location.source.id = jid,
               from_publication_date = "2020-01-01",
               to_publication_date = "2023-12-31")
 w |>
   summarise(oa_rate = mean(is_oa, na.rm = TRUE),
             gold_oa = mean(oa_status == "gold", na.rm = TRUE)) |>
   mutate(field = field)
}) |> list_rbind()

oa_rates
```

*Key insight:* Biology tends to have the highest gold OA rate (driven by PLOS,
eLife, etc.), while sociology and mathematics lag behind.

**2. Temporal OA growth.**
Track the year-over-year OA growth rate.

```{r}
#| label: sol-ch27-2
#| eval: false
oa_growth <- works |>
 filter(is_oa) |>
 count(publication_year, name = "n_oa") |>
 mutate(yoy_growth = (n_oa / lag(n_oa)) - 1)

ggplot(oa_growth, aes(publication_year, yoy_growth)) +
 geom_col() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "Year-over-year OA growth",
      x = "Year", y = "Growth rate")
```

*Key insight:* OA growth accelerated around 2020, likely driven by COVID-19
mandates and Plan S implementation.

**3. OA and collaboration.**
Test whether author count predicts OA status.

```{r}
#| label: sol-ch27-3
#| eval: false
works |>
 mutate(n_authors = map_int(author, nrow)) |>
 group_by(n_authors = pmin(n_authors, 10)) |>
 summarise(oa_rate = mean(is_oa, na.rm = TRUE), n = n()) |>
 ggplot(aes(n_authors, oa_rate)) +
 geom_col() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "OA rate by number of authors",
      x = "Authors (capped at 10)", y = "OA rate")
```

*Key insight:* Larger teams may have more funding to cover APCs. International
collaborations also increase OA probability because at least one funder may
mandate it.

## Chapter 28 --- Gender and Equity Analysis {-}

**1. Cross-journal comparison.**
Compare gender representation in first authorship across journals.

```{r}
#| label: sol-ch28-1
#| eval: false
library(openalexR)
library(tidyverse)

journal_ids <- c("S4210174107", "S49861241", "S154589771")

gender_by_journal <- map(journal_ids, \(jid) {
 w <- oa_fetch(entity = "works",
               primary_location.source.id = jid,
               from_publication_date = "2020-01-01") |>
   slice_sample(n = 200)

 w |>
   mutate(first_author = map_chr(author, \(a) a$au_display_name[1])) |>
   # Gender inference would use genderize.io or similar
   mutate(journal = jid)
}) |> list_rbind()

# After gender inference:
# gender_by_journal |>
#   group_by(journal, inferred_gender) |>
#   summarise(n = n()) |>
#   mutate(pct = n / sum(n))
```

*Key insight:* Gender representation varies significantly by field. STEM
journals typically have lower female first-author rates than social sciences.

**2. Temporal trends.**
Track the proportion of female first authors over time.

```{r}
#| label: sol-ch28-2
#| eval: false
# After gender inference has been performed:
gender_trend <- works_gendered |>
 group_by(publication_year, inferred_gender) |>
 summarise(n = n(), .groups = "drop_last") |>
 mutate(pct = n / sum(n))

ggplot(gender_trend |> filter(inferred_gender == "female"),
      aes(publication_year, pct)) +
 geom_line() +
 geom_point() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "Female first-author share over time",
      x = "Year", y = "Proportion female")
```

*Key insight:* Most fields show a gradual increase in female representation,
though the rate of change varies considerably by discipline and region.

**3. Citation gap.**
Compare citations by inferred gender, controlling for year.

```{r}
#| label: sol-ch28-3
#| eval: false
# After gender inference:
citation_gap <- works_gendered |>
 group_by(publication_year, inferred_gender) |>
 summarise(
   mean_cites = mean(cited_by_count),
   median_cites = median(cited_by_count),
   n = n(),
   .groups = "drop"
 )

ggplot(citation_gap, aes(publication_year, mean_cites,
                          colour = inferred_gender)) +
 geom_line() +
 labs(title = "Mean citations by inferred gender",
      x = "Year", y = "Mean citations")

# Statistical test (controlling for year)
summary(lm(cited_by_count ~ inferred_gender + publication_year,
           data = works_gendered))
```

*Key insight:* A persistent citation gap may reflect systemic biases, network
effects, or confounding factors (e.g., team size, field). Observational data
alone cannot establish causation.

## Chapter 29 --- Retraction Analysis {-}

**1. Retraction rate by field.**
Compare retraction rates across fields, normalised by corpus size.

```{r}
#| label: sol-ch29-1
#| eval: false
library(openalexR)
library(tidyverse)

fields <- c("biomedicine", "computer science")

retraction_rates <- map(fields, \(f) {
 all_works <- oa_fetch(entity = "works", search = f,
                       from_publication_date = "2015-01-01",
                       count_only = TRUE)
 retracted <- oa_fetch(entity = "works", search = f,
                       type = "retraction",
                       from_publication_date = "2015-01-01",
                       count_only = TRUE)
 tibble(
   field = f,
   total = all_works$count,
   retracted = retracted$count,
   rate_per_1000 = retracted$count / all_works$count * 1000
 )
}) |> list_rbind()

retraction_rates
```

*Key insight:* Retraction rates are typically higher in biomedical fields (1-2
per 1,000) than in computer science, partly reflecting different scrutiny levels
and reproducibility challenges.

**2. Citation trajectory.**
Examine when a retracted paper's citations were received.

```{r}
#| label: sol-ch29-2
#| eval: false
retracted_paper <- oa_fetch(entity = "works",
                            doi = "10.xxxx/retracted-example")

if (!is.null(retracted_paper$counts_by_year)) {
 retracted_paper |>
   unnest(counts_by_year) |>
   ggplot(aes(year, cited_by_count)) +
   geom_col() +
   geom_vline(xintercept = 2020, linetype = "dashed",
              colour = "red") +  # retraction year
   labs(title = "Citation trajectory of a retracted paper",
        x = "Year", y = "Citations received")
}
```

*Key insight:* Many retracted papers continue to be cited after retraction,
indicating that the retraction notice does not always reach citing authors.

**3. Retraction-citation correlation.**
Test whether pre-retraction citation count predicts post-retraction citations.

```{r}
#| label: sol-ch29-3
#| eval: false
# Assuming a set of retracted papers with known retraction years
retracted_set <- retracted_works |>
 unnest(counts_by_year) |>
 mutate(period = ifelse(year < retraction_year, "pre", "post")) |>
 group_by(id, period) |>
 summarise(cites = sum(cited_by_count), .groups = "drop") |>
 pivot_wider(names_from = period, values_from = cites,
             values_fill = 0)

cor(retracted_set$pre, retracted_set$post, method = "spearman")

ggplot(retracted_set, aes(pre, post)) +
 geom_point(alpha = 0.5) +
 geom_smooth(method = "lm") +
 labs(x = "Pre-retraction citations",
      y = "Post-retraction citations",
      title = "Pre- vs. post-retraction citation counts")
```

*Key insight:* A positive correlation suggests that highly visible papers
continue to be cited after retraction --- the "retraction paradox."

## Chapter 30 --- Funding Analysis {-}

**1. Funder co-occurrence.**
Identify frequently co-occurring funder pairs.

```{r}
#| label: sol-ch30-1
#| eval: false
library(tidyverse)

funder_pairs <- works |>
 unnest(grants) |>
 group_by(id) |>
 filter(n_distinct(funder) >= 2) |>
 reframe(pair = combn(funder, 2, simplify = FALSE)) |>
 unnest_wider(pair, names_sep = "_") |>
 count(pair_1, pair_2, sort = TRUE)

funder_pairs |> head(10)
```

*Key insight:* National funders frequently co-occur with international
organisations (e.g., NIH + EU), reflecting collaborative funding models.

**2. Funder portfolios.**
Compare topical profiles of two major funders.

```{r}
#| label: sol-ch30-2
#| eval: false
library(openalexR)
library(tidyverse)

funder_ids <- c("F123456", "F789012")  # OpenAlex funder IDs

funder_profiles <- map(funder_ids, \(fid) {
 w <- oa_fetch(entity = "works",
               grants.funder = fid,
               from_publication_date = "2020-01-01") |>
   slice_sample(n = 300)
 w |>
   unnest(keywords) |>
   count(keyword, sort = TRUE) |>
   mutate(funder = fid)
}) |> list_rbind()

# Compare top keywords
funder_profiles |>
 group_by(funder) |>
 slice_head(n = 15) |>
 ggplot(aes(reorder(keyword, n), n, fill = funder)) +
 geom_col(position = "dodge") +
 coord_flip() +
 labs(title = "Funder portfolio comparison")
```

*Key insight:* Different funders support different thematic areas; comparing
portfolios reveals strategic funding priorities.

**3. Temporal trends.**
Track funding acknowledgement coverage over time.

```{r}
#| label: sol-ch30-3
#| eval: false
funding_trend <- works |>
 mutate(has_funding = map_lgl(grants, \(g) length(g) > 0)) |>
 group_by(publication_year) |>
 summarise(funding_rate = mean(has_funding, na.rm = TRUE))

ggplot(funding_trend, aes(publication_year, funding_rate)) +
 geom_line() +
 geom_point() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "Papers with funding acknowledgements over time",
      x = "Year", y = "Share with funding data")
```

*Key insight:* Funding metadata coverage has increased as publishers and
databases improve structured acknowledgement parsing.

## Chapter 31 --- Patent-Paper Linkages {-}

**1. Field comparison.**
Compare the proportion of highly cited papers across fields.

```{r}
#| label: sol-ch31-1
#| eval: false
library(openalexR)
library(tidyverse)

biomed <- oa_fetch(entity = "works",
                   search = "drug delivery",
                   from_publication_date = "2018-01-01",
                   to_publication_date = "2020-12-31") |>
 slice_sample(n = 200) |> mutate(field = "Biomedical")

socsci <- oa_fetch(entity = "works",
                   search = "social inequality",
                   from_publication_date = "2018-01-01",
                   to_publication_date = "2020-12-31") |>
 slice_sample(n = 200) |> mutate(field = "Social Science")

combined <- bind_rows(biomed, socsci)

combined |>
 group_by(field) |>
 summarise(
   pct_top10 = mean(cited_by_count >=
                     quantile(cited_by_count, 0.9)),
   mean_cites = mean(cited_by_count)
 )
```

*Key insight:* Biomedical research is more likely to be cited in patents because
it has more direct translational applications. Higher citation rates do not
automatically imply greater technological relevance.

**2. Time lag estimation.**
Estimate the citation accumulation trajectory for a cohort of papers.

```{r}
#| label: sol-ch31-2
#| eval: false
cohort_2015 <- works |>
 filter(publication_year == 2015, !is.null(counts_by_year))

trajectory <- cohort_2015 |>
 unnest(counts_by_year) |>
 mutate(age = year - 2015) |>
 filter(age >= 0) |>
 group_by(age) |>
 summarise(median_cites = median(cited_by_count))

ggplot(trajectory, aes(age, median_cites)) +
 geom_line() +
 geom_point() +
 labs(title = "Citation trajectory for 2015 cohort",
      x = "Age (years since publication)",
      y = "Median annual citations")

# Peak year
peak_age <- trajectory$age[which.max(trajectory$median_cites)]
cat(sprintf("Peak citation year: age %d\n", peak_age))
```

*Key insight:* Papers typically peak in citation accumulation at age 2-4 years,
depending on the field. This lag matters for evaluating recent research.

**3. Institutional patent linkage.**
Compare the proportion of top-cited papers across institutions.

```{r}
#| label: sol-ch31-3
#| eval: false
inst_ids <- c("I123456789", "I987654321")

inst_top_papers <- map(inst_ids, \(iid) {
 w <- oa_fetch(entity = "works",
               authorships.institutions.lineage = iid,
               from_publication_date = "2018-01-01",
               to_publication_date = "2020-12-31")
 p90 <- quantile(w$cited_by_count, 0.9)
 tibble(
   institution = iid,
   n_papers = nrow(w),
   pct_above_p90 = mean(w$cited_by_count >= p90)
 )
}) |> list_rbind()

inst_top_papers
```

*Key insight:* A higher proportion of top-percentile papers may suggest greater
potential for technological impact, but causation cannot be established without
actual patent citation data.

## Chapter 32 --- Causal Inference in Scientometrics {-}

**1. Extended DiD.**
Add covariates to the difference-in-differences regression.

```{r}
#| label: sol-ch32-1
#| eval: false
library(tidyverse)

# Assuming works_did has: cited_by_count, is_oa, post_treatment,
# type, n_authors, referenced_works_count
did_extended <- lm(
 cited_by_count ~ is_oa * post_treatment +
   type + n_authors + referenced_works_count,
 data = works_did
)

summary(did_extended)

# Compare with simple model
did_simple <- lm(cited_by_count ~ is_oa * post_treatment,
                 data = works_did)

cat(sprintf("Simple interaction: %.2f\n",
            coef(did_simple)["is_oaTRUE:post_treatmentTRUE"]))
cat(sprintf("Extended interaction: %.2f\n",
            coef(did_extended)["is_oaTRUE:post_treatmentTRUE"]))
```

*Key insight:* If the interaction term changes substantially with added
covariates, the simple model suffers from omitted variable bias.

**2. Placebo test.**
Use a false treatment date to test the parallel trends assumption.

```{r}
#| label: sol-ch32-2
#| eval: false
# Placebo: pretend treatment happened in 2018 instead of 2020
works_placebo <- works_did |>
 filter(publication_year <= 2019) |>  # pre-treatment only
 mutate(post_placebo = publication_year >= 2018)

did_placebo <- lm(cited_by_count ~ is_oa * post_placebo,
                  data = works_placebo)

summary(did_placebo)

# If the interaction is NOT significant, parallel trends hold
cat(sprintf("Placebo p-value: %.4f\n",
            summary(did_placebo)$coefficients[
             "is_oaTRUE:post_placeboTRUE", "Pr(>|t|)"
            ]))
```

*Key insight:* A non-significant placebo interaction (p > 0.05) supports the
parallel trends assumption. A significant result suggests pre-existing
divergence between OA and non-OA papers.

**3. Matched comparison.**
Use propensity score matching to estimate the OA effect.

```{r}
#| label: sol-ch32-3
#| eval: false
library(MatchIt)
library(tidyverse)

# Propensity score matching
matched <- matchit(
 is_oa ~ publication_year + referenced_works_count + type,
 data = works_did,
 method = "nearest",
 ratio = 1
)

summary(matched)
matched_data <- match.data(matched)

# Estimate treatment effect on matched sample
did_matched <- lm(cited_by_count ~ is_oa, data = matched_data)
summary(did_matched)

cat(sprintf("Unmatched OA effect: %.2f\n",
            coef(did_simple)["is_oaTRUE"]))
cat(sprintf("Matched OA effect: %.2f\n",
            coef(did_matched)["is_oaTRUE"]))
```

*Key insight:* Matching reduces selection bias by creating comparable groups.
If the matched and unmatched effects differ substantially, selection on
observables was important.

## Chapter 33 --- Reproducibility, RRIDs, and Badges {-}

**1. DAS detection.**
Search abstracts for data/code availability statements.

```{r}
#| label: sol-ch33-1
#| eval: false
library(tidyverse)

das_keywords <- c("data availability", "code availability",
                  "github\\.com", "zenodo", "figshare",
                  "data are available", "code is available")

das_pattern <- paste(das_keywords, collapse = "|")

works |>
 filter(!is.na(ab)) |>
 mutate(has_das = str_detect(tolower(ab), das_pattern)) |>
 summarise(
   total = n(),
   with_das = sum(has_das),
   das_rate = mean(has_das)
 )
```

*Key insight:* DAS detection via abstract text is a lower bound; many statements
appear only in dedicated sections not captured by abstracts.

**2. Cross-journal comparison.**
Compare metadata completeness and OA rates across journals.

```{r}
#| label: sol-ch33-2
#| eval: false
journal_ids <- c("S4210174107", "S49861241", "S137773608")

journal_quality <- map(journal_ids, \(jid) {
 w <- oa_fetch(entity = "works",
               primary_location.source.id = jid,
               from_publication_date = "2020-01-01") |>
   slice_sample(n = 200)
 tibble(
   journal = jid,
   abstract_rate = mean(!is.na(w$ab)),
   oa_rate = mean(w$is_oa, na.rm = TRUE),
   doi_rate = mean(!is.na(w$doi))
 )
}) |> list_rbind()

journal_quality
```

*Key insight:* Journals with higher OA rates tend to have better metadata
completeness, but the correlation is not perfect.

**3. Temporal trends.**
Track transparency proxies over time.

```{r}
#| label: sol-ch33-3
#| eval: false
transparency_trends <- works |>
 group_by(publication_year) |>
 summarise(
   abstract_rate = mean(!is.na(ab)),
   oa_rate = mean(is_oa, na.rm = TRUE),
   high_cite_rate = mean(cited_by_count >= 10)
 ) |>
 pivot_longer(-publication_year, names_to = "metric",
              values_to = "value")

ggplot(transparency_trends, aes(publication_year, value,
                                 colour = metric)) +
 geom_line() +
 scale_y_continuous(labels = scales::percent) +
 labs(title = "Transparency and impact proxies over time",
      x = "Year", y = "Proportion")
```

*Key insight:* Transparency proxies and citation impact may be weakly correlated
within years, but confounding factors (field, journal prestige) complicate
interpretation.

## Chapter 34 --- Reproducible Pipelines {-}

**1. Build a pipeline.**
Create a `_targets.R` file for a bibliometric workflow.

```{r}
#| label: sol-ch34-1
#| eval: false
library(targets)

# _targets.R content:
tar_option_set(packages = c("openalexR", "tidyverse"))

list(
 tar_target(raw_data, {
   oa_fetch(entity = "works", search = "bibliometrics",
            from_publication_date = "2020-01-01")
 }),

 tar_target(clean_data, {
   raw_data |>
     filter(!is.na(doi), !is.na(publication_year)) |>
     select(id, doi, display_name, publication_year,
            cited_by_count, is_oa)
 }),

 tar_target(summary_stats, {
   clean_data |>
     group_by(publication_year) |>
     summarise(n = n(), mean_cites = mean(cited_by_count))
 }),

 tar_target(trend_plot, {
   ggplot(summary_stats, aes(publication_year, n)) +
     geom_col() +
     labs(title = "Publication trend", x = "Year", y = "Works")
 })
)

# Run with: tar_make()
# Verify: tar_visnetwork()
```

*Key insight:* Changing the summary function only re-runs `summary_stats` and
`trend_plot`, not `raw_data` or `clean_data` --- this is the power of
dependency-based pipelines.

**2. renv snapshot.**
Demonstrate dependency management with renv.

```{r}
#| label: sol-ch34-2
#| eval: false
# Step 1: Initialise renv
# renv::init()

# Step 2: Install a package
# install.packages("glue")
# library(glue)

# Step 3: Snapshot
# renv::snapshot()
# This creates/updates renv.lock

# Step 4: Simulate loss --- remove the package
# remove.packages("glue")
# library(glue)  # Error: package not found

# Step 5: Restore
# renv::restore()
# library(glue)  # Works again
```

*Key insight:* renv creates a project-local library and a lockfile, ensuring
exact reproducibility of package versions across machines and time.

**3. Branching.**
Use dynamic branching to analyse multiple journals in parallel.

```{r}
#| label: sol-ch34-3
#| eval: false
library(targets)

# _targets.R with dynamic branching:
tar_option_set(packages = c("openalexR", "tidyverse"))

list(
 tar_target(journal_ids,
            c("S4210174107", "S137773608", "S49861241")),

 tar_target(journal_data,
   oa_fetch(entity = "works",
            primary_location.source.id = journal_ids,
            from_publication_date = "2020-01-01"),
   pattern = map(journal_ids)
 ),

 tar_target(journal_summary,
   journal_data |>
     summarise(
       journal = first(so),
       n = n(),
       mean_cites = mean(cited_by_count)
     ),
   pattern = map(journal_data)
 )
)
```

*Key insight:* Dynamic branching avoids duplicating target definitions; each
journal is processed independently and can run in parallel with
`tar_make_future()`.

## Chapter 35 --- Interactive Dashboards {-}

**1. Multi-page dashboard.**
Create a Quarto dashboard with two pages.

```{r}
#| label: sol-ch35-1
#| eval: false
# Save as dashboard.qmd:
# ---
# title: "Bibliometric Dashboard"
# format: dashboard
# ---
#
# # Overview
#
# ## Row
#
# ```{r}
# library(plotly)
# plot_ly(summary_data, x = ~year, y = ~n, type = "bar")
# ```
#
# ## Row
#
# ```{r}
# plot_ly(works, x = ~cited_by_count, type = "histogram")
# ```
#
# # Details
#
# ## Row
#
# ```{r}
# library(DT)
# datatable(works |> select(display_name, publication_year,
#                           cited_by_count, is_oa))
# ```
```

*Key insight:* Quarto dashboards use page-level headings (`#`) for navigation
tabs and row/column-level headings (`##`) for layout.

**2. Linked brushing.**
Use plotly selection to filter a DT table.

```{r}
#| label: sol-ch35-2
#| eval: false
library(plotly)
library(crosstalk)
library(DT)

shared_data <- SharedData$new(works)

bscols(
 plot_ly(shared_data, x = ~publication_year,
         y = ~cited_by_count, type = "scatter",
         mode = "markers") |>
   layout(dragmode = "select"),

 datatable(shared_data,
           options = list(pageLength = 10),
           selection = "multiple")
)
```

*Key insight:* `crosstalk::SharedData` enables linked brushing between any
htmlwidget that supports it, without needing Shiny.

**3. Custom tooltips.**
Add informative hover text to a plotly chart.

```{r}
#| label: sol-ch35-3
#| eval: false
library(plotly)
library(tidyverse)

tooltip_data <- works |>
 group_by(publication_year) |>
 summarise(n = n(), mean_cites = round(mean(cited_by_count), 1))

plot_ly(tooltip_data,
       x = ~publication_year,
       y = ~n,
       type = "bar",
       text = ~paste0("Year: ", publication_year,
                      "<br>Papers: ", n,
                      "<br>Mean cites: ", mean_cites),
       hoverinfo = "text")
```

*Key insight:* Custom tooltips using `text` and `hoverinfo = "text"` provide
context-rich information on hover, improving dashboard usability.

## Chapter 36 --- Parameterized Quarto Reports {-}

**1. Institutional template.**
Create a parameterized report for institutional profiles.

```{r}
#| label: sol-ch36-1
#| eval: false
# Save as inst_profile.qmd:
# ---
# title: "Institutional Profile"
# params:
#   inst_id: "I123456789"
#   inst_name: "University of Example"
#   start_year: 2020
#   end_year: 2023
# ---
#
# ```{r}
# library(openalexR)
# library(tidyverse)
#
# works <- oa_fetch(
#   entity = "works",
#   authorships.institutions.lineage = params$inst_id,
#   from_publication_date = paste0(params$start_year, "-01-01"),
#   to_publication_date = paste0(params$end_year, "-12-31")
# )
#
# cat(sprintf("Total publications: %d\n", nrow(works)))
# ```
#
# ## Publication Trend
# ```{r}
# works |>
#   count(publication_year) |>
#   ggplot(aes(publication_year, n)) +
#   geom_col() +
#   labs(title = params$inst_name)
# ```
#
# ## Citation Distribution
# ```{r}
# ggplot(works, aes(cited_by_count)) +
#   geom_histogram(bins = 30) +
#   scale_x_log10()
# ```
#
# ## Top Authors
# ```{r}
# works |> unnest(author) |>
#   count(au_display_name, sort = TRUE) |>
#   head(10)
# ```
```

*Key insight:* Parameters make the same template reusable for any institution;
batch rendering generates a portfolio of reports efficiently.

**2. Batch rendering.**
Render the template for multiple journals.

```{r}
#| label: sol-ch36-2
#| eval: false
library(quarto)

journals <- list(
 list(journal_id = "S4210174107", journal_name = "Scientometrics"),
 list(journal_id = "S137773608", journal_name = "Nature Physics"),
 list(journal_id = "S49861241", journal_name = "PLOS ONE")
)

walk(journals, \(j) {
 quarto_render(
   "journal_profile.qmd",
   execute_params = list(
     journal_id = j$journal_id,
     journal_name = j$journal_name,
     start_year = 2020,
     end_year = 2023
   ),
   output_file = paste0("report_", j$journal_name, ".html")
 )
})
```

*Key insight:* Batch rendering with `quarto_render()` and parameterised
templates scales to hundreds of reports with minimal code.

**3. Conditional content.**
Add a warning when the sample is too small.

```{r}
#| label: sol-ch36-3
#| eval: false
# In the parameterized .qmd file:
#
# ```{r}
# #| results: asis
# if (nrow(works) < 50) {
#   cat("<div class="callout callout-warn">\n")
#   cat("This sample contains fewer than 50 papers. ")
#   cat("The citation distribution may not be reliable.\n")
#   cat(":::\n")
# } else {
#   ggplot(works, aes(cited_by_count)) +
#     geom_histogram(bins = 30) +
#     scale_x_log10() +
#     labs(title = "Citation distribution")
# }
# ```
```

*Key insight:* Conditional logic with `results: asis` enables dynamic content
switching --- essential for robust automated reports that handle edge cases.

## Chapter 37 --- Building a Bibliometric Shiny App {-}

**1. Add network tab.**
Extend the app with a co-authorship network visualisation.

```{r}
#| label: sol-ch37-1
#| eval: false
library(shiny)
library(visNetwork)
library(igraph)
library(tidyverse)

# Add to UI:
# tabPanel("Network",
#   visNetworkOutput("coauth_network", height = "600px")
# )

# Add to server:
# output$coauth_network <- renderVisNetwork({
#   req(data())
#   edges <- data() |>
#     unnest(author) |>
#     group_by(id) |>
#     filter(n() >= 2, n() <= 10) |>
#     reframe(pair = combn(au_display_name, 2, simplify = FALSE)) |>
#     unnest_wider(pair, names_sep = "_") |>
#     count(pair_1, pair_2, name = "weight") |>
#     rename(from = pair_1, to = pair_2)
#
#   g <- graph_from_data_frame(edges, directed = FALSE)
#   comm <- cluster_leiden(g, resolution = 1.0)
#
#   nodes <- tibble(
#     id = V(g)$name, label = V(g)$name,
#     group = as.character(membership(comm))
#   )
#
#   visNetwork(nodes, edges) |>
#     visOptions(highlightNearest = TRUE, selectedBy = "group")
# })
```

*Key insight:* Limiting the network to papers with 2-10 authors avoids
overwhelming the visualisation with hyper-authored papers.

**2. Caching.**
Implement server-side caching for repeated queries.

```{r}
#| label: sol-ch37-2
#| eval: false
library(shiny)
library(memoise)

# Create a cached version of the data fetching function
fetch_cached <- memoise(function(journal_id, start_date, end_date) {
 oa_fetch(
   entity = "works",
   primary_location.source.id = journal_id,
   from_publication_date = start_date,
   to_publication_date = end_date
 )
}, cache = cachem::cache_disk("shiny_cache"))

# In the server:
# data <- reactive({
#   fetch_cached(
#     input$journal_id,
#     paste0(input$start_year, "-01-01"),
#     paste0(input$end_year, "-12-31")
#   )
# })
```

*Key insight:* Disk-backed caching persists across sessions, so the same query
never hits the API twice. Use `cachem::cache_disk()` for production apps.

**3. Deployment.**
Deploy to shinyapps.io.

```{r}
#| label: sol-ch37-3
#| eval: false
# Step 1: Install rsconnect
# install.packages("rsconnect")

# Step 2: Configure account
# rsconnect::setAccountInfo(
#   name = "your-account",
#   token = "your-token",
#   secret = "your-secret"
# )

# Step 3: Deploy
# rsconnect::deployApp(
#   appDir = ".",
#   appName = "bibliometric-explorer",
#   appTitle = "Bibliometric Explorer"
# )

# Step 4: Test with different inputs
# - Try different journal IDs
# - Test with narrow and wide date ranges
# - Verify that caching works (second load should be instant)
```

*Key insight:* Test with a variety of inputs before deploying; edge cases (empty
results, very large corpora) should be handled gracefully with validation
messages.

**4. biblioshiny comparison.**
Run biblioshiny and compare features.

```{r}
#| label: sol-ch37-4
#| eval: false
library(bibliometrix)
# biblioshiny()

# Feature comparison table:
# | Feature              | biblioshiny | Custom app |
# |---------------------|-------------|------------|
# | Data sources        | WoS, Scopus | OpenAlex   |
# | Network analysis    | Yes         | Basic      |
# | Topic analysis      | Yes         | No         |
# | Custom indicators   | Limited     | Full       |
# | Parameterization    | No          | Yes        |
# | Deployment          | Local only  | shinyapps  |
# | Reproducibility     | GUI-based   | Code-based |
#
# biblioshiny is comprehensive but not easily customisable;
# a custom app trades breadth for tailored functionality.
```

*Key insight:* biblioshiny is excellent for exploration but lacks
reproducibility and customisation. A custom app is better for production
workflows with specific analytical requirements.
